---
title: "Mitochondria integrate oncogenic and developmental signals to time Myc-driven mammary tumourigenesis"
subtitle: "ARCHIVED -- the July 2026 exploratory walkthrough (superseded)"
author: "Szabadkai lab"
date: today
---

::: {.callout-important}
## ARCHIVED 2026-08-14 -- superseded by `paper/analysis_record.qmd`

**What this is.** The July 2026 walkthrough of the **exploratory** figure layer,
`figures/fig01`-`fig05` / `figS1`-`figS8` and the four assembled `figure*.R` composites. It
predates the written Results, the 25-panel `figures/panels/` layer, and scripts 40-44.

**What replaced it.** `paper/analysis_record.qmd` is the record of the analysis as it stands:
the three written Results sections verbatim, all 25 manuscript panels with their legend
blocks, the retracted corpus and the open questions.

**Why it is kept rather than deleted.** Its prose on the global axis, the correlation ceiling
and the mitochondrial-group survey is not reproduced anywhere else, and the record document
cites it. Its **figure manifest and slot table below are OUT OF DATE** -- `figures/panels/
PANELS.md` is the manifest of record. Nothing here should be quoted without checking it
against the record document first.
:::

::: {.callout-note}
**Skeleton, not the finished document.** This is the container the reflective write-up
will grow into as the publication panels are finalised. It mirrors the exploratory arc of
scripts 32-39 and reuses the same `results/*.rds` objects and the `figures/` panel layer.
Each section below is a stub: prose to be written, panels to be embedded. The organising
principle, settled 2026-07: **genotype CONTRASTS survived; cross-sectional COUPLINGS sit
at a correlation ceiling and mostly did not.**
:::

```{r}
#| label: setup
#| include: false
library(here)
library(ggplot2)

# Pin here() to the repo root: this .qmd lives under paper/, and a knitr render
# starts the session there, so anchor explicitly (sourced panel scripts inherit it).
here::i_am("paper/archive/myc_mito.qmd")

# Display panels inline without re-exporting their canonical PDFs to outputs/.
options(myc.fig.nosave = TRUE)

# Read-only load of the analysis objects. The figure layer never re-runs analysis;
# it reads what scripts 00-39 already wrote to results/.
res <- function(x) readRDS(here::here("results", paste0(x, ".rds")))

# Loaded lazily per section as the prose is written; listed here as the manifest:
#   content   <- res("mito_content_proxies")        # script 32 -- Fig content
#   mtaxis    <- res("mtdna_axis_and_coupling_null") # script 33 -- adjusted +21%
#   loading   <- res("pathway_loading")              # script 37 -- global axis / dim
#   coupling  <- res("linear_pathway_coupling")      # script 36 -- ceiling origin
#   lesner    <- res("lesner_mechanism_test")        # script 39 -- external validation
```

::: {.callout-tip collapse="true"}
## Figure manifest -- build number vs manuscript figure

**The bioinformatics is TWO main figures of up to six panels plus two supplementary**
(author, 2026-07-27), and they follow the order of the Results paragraph: what MYC builds,
then what development takes away and what that hands to the cell experiments.

| paper figure | script | panels | source analysis | section |
|--------------|--------|--------|-----------------|---------|
| **Figure 1** | `figure1_myc_mitochondrion.R` | A-D | 32, 08, 42 | @sec-content |
| **Figure 2** | `figure2_developmental_window.R` | A-F | 43, 44, 42 | @sec-window |
| Supp 1 | `figureS1_compartment_detail.R` | A-C | 32, 08 | @sec-mito-survey |
| Supp 2 | `figureS2_controls.R` | A-D | 40, 41, 44 | @sec-attenuation |

Everything else under `figures/` -- `fig01`-`fig05`, `figS1`-`figS8` -- stays as the
**exploratory layer** this document walks through. Those scripts are not renamed: their
numbers are what the commits, the dated docs and `outputs/figures/` cite.

:::

# Overview {#sec-overview}

**What was measured.** Bulk RNA-seq of **purified mammary epithelial cells** from MMTV-Myc
transgenic mice and their wild-type littermates at **6 and 12 weeks** -- four groups
(`6W_neg`, `6W_pos`, `12W_neg`, `12W_pos`), **six replicates each, n=24**, analysed with DESeq2
on a `~ timepoint * myc_status` interaction design. Purification is by **enzymatic dissociation
only, with no sorting step**, and that single fact sets two standing rules. Any stromal,
endothelial or immune signal is **contamination rather than tissue composition** (3-12%
residual, which is what no sort predicts), so it is modelled as a nuisance and never read as
biology. And the warm digest is itself a stressor: the immediate-early signature it induces
(van den Brink 2017) is a genuine axis in these samples, which is why prep-stress appears as a
covariate wherever a per-sample quantity is interpreted.

::: {.callout-important}
**Batch = timepoint, and it governs every reading below.** The 6W and 12W cohorts were
extracted as two separate batches, so batch is perfectly confounded with age. Genotype is
**balanced within each batch**, so the interaction design absorbs the batch effect into the
timepoint main effect. **Every genotype gap and every interaction is therefore clean** -- and
so is any *difference between two temporal contrasts*, because the shared offset cancels in a
difference. What is not clean is the pure between-timepoint comparison, which is exactly the
developmental reading; with no dissociation-batch, viability or RIN metadata on disk it cannot
be separated post hoc. Developmental statements here are consequently **described and bounded,
never claimed on their own**, and where one carries weight (@sec-window) it is carried by a
comparison made *within* the timecourse against expression-matched nulls, not by the
timecourse itself.
:::

**Three rulers, which answer different questions and should not be mixed.** **Absolute
content** -- a gene set's share of the library in raw counts -- answers *how much of this
compartment is there*, and is the ruler of @sec-content. **mitoPPS** (Monzel 2025) is
deliberately **content-blind**: a within-compartment pairwise ratio answering *what is
prioritised inside the mitochondrion*, and because it is a ratio it cancels the per-sample
common mode (@sec-reallocation). **fGSEA on the Wald statistic** reads absolute
transcriptional enrichment for a given contrast. A fourth quantity, the **per-sample composite
score**, runs through the exploratory layer and is the one to distrust on sight: those scores
are nearly one-dimensional and correlate with one another at ~0.8 by default, for reasons
@sec-axis sets out in full.

Hence the rule this document holds to throughout: **contrasts are claims, couplings are
rankings.** At six replicates per group nothing in the coupling layer clears multiple testing,
and it is presented as what it is -- an ordered list of propositions for the bench
(@sec-leads), not as evidence. The contrast layer is where the results are.

**How to read what follows.** The document walks the exploratory process and then lands the
defensible part, in the order of the Results paragraph. @sec-content is what MYC builds
(Figure 1); @sec-window is what development takes away and what that hands to the cell
experiments (Figure 2); the two supplementary figures are the controls each of those rests on,
embedded at @sec-mito-survey and @sec-attenuation. Everything between is the exploratory
layer -- the compartment survey, OXPHOS complex by complex, the reallocation ranking, the
global axis and correlation ceiling, and the external comparison with Lesner et al.
(@sec-lesner). Results reported as the author's own bench data -- the MYC and BAX/BIM/PUMA
blots, the MYC-ER, MYAZ and Bcl-xL experiments -- are marked as such where they appear and are
taken as given, not reanalysed here. The analytical companion to this work is Menegollo,
Bentham et al., *Cancer Res* 2024.

# MYC builds a death-competent mitochondrion {#sec-content}

**MANUSCRIPT FIGURE 1.** At constant dose (@fig-myc-dose), MYC raises the **nuclear**
mitochondrial arms -- mass markers, nuclear MitoCarta and nuclear OXPHOS -- while the 13
mtDNA-encoded transcripts, the reference arm, stay flat: a content increase of ~**+21%**
(p=0.0006) after adjusting for prep-stress and contamination, and a *lower bound*, because MYC
amplifies total RNA per cell (@fig-myc-mito A). The biogenesis arm behaves the same way and is
the exploratory detail below.

**But the compartment is reprioritised, not simply enlarged.** Ranking every MitoPathway by
MYC's effect on mitoPPS -- a within-compartment ratio, blind to how many mitochondria a cell
has -- separates promoted from demoted arms in an order structured by function
(@fig-myc-mito B). Plotting the two rulers against each other shows what that means: the
content-down / priority-up quadrant is **empty** (@fig-myc-mito C). Demoted arms are not
switched off; they rise more slowly than the rest and lose share. Reprioritisation here is a
skew on a broad rise, not a redistribution of a fixed budget.

**And the organelle MYC builds can execute its own death.** The same genotype contrast raises
`Htra2` (**+1.49**, padj 2.6e-08) and `Bax` (**+0.48**, padj 1.4e-04) alongside the import,
cristae and OXPHOS arms, while lowering `Bcl2l1`/BCL-xL (**-0.42**, padj 0.040)
(@fig-myc-mito D). Asset and liability, built in one move -- the trade-off written as
molecules rather than as a model.

```{r}
#| label: fig-myc-mito
#| fig-cap: "MANUSCRIPT FIGURE 1 (figure1). MYC builds a death-competent mitochondrion. (A) Compartment share across the four groups for three nuclear arms, with the 13 mtDNA-encoded transcripts as the reference arm that does not move; brackets are the genotype comparison at each age, which is the batch-clean axis. (B) All MitoPathways ranked by MYC's effect on mitoPPS. (C) Content against priority, one point per pathway; the content-down / priority-up quadrant is empty. (D) The curated death-and-mitochondrion roster on the same genotype contrast. Genotype contrast throughout, n=6/group, raw log2FC."
#| fig-width: 9
#| fig-height: 7.7
#| echo: false
source(here::here("figures", "figure1_myc_mitochondrion.R"))
print(p)
```

The rest of this section is the exploratory detail behind panel A.

The biogenesis arm is **MitoCarta central dogma + protein import/sorting/homeostasis** (314
genes: mtDNA replication and transcription, mtRNA metabolism, mitochondrial translation and
ribosome, import, chaperones, proteases). It replaces the earlier mitoribosome facet for two
reasons: only **7 of its 314 genes** are in nuclear OXPHOS, so *OXPHOS vs general biogenesis*
is a genuinely disjoint comparison rather than a whole-vs-part one; and it is the same
`mito_biogenesis` axis the coupling analyses use (script 28), so figure and analysis name one
thing. Deliberately **not** the `Biogenesis_discrimination` panel (`CORE_MITO`, `MYC_MITO`,
`DEVELOPMENTAL_MITO`): those are regulons, and `MYC_MITO` is Myc targets by construction --
circular on a genotype axis.

```{r}
#| label: fig-content
#| fig-cap: "Myc raises the nuclear mitochondrial arms; mtDNA-encoded output stays flat. Genotype is the clean axis; the time axis is cohort-confounded. n=6/group."
#| fig-width: 9
#| fig-height: 4.4
#| echo: false
# Reuse the panel script: it builds `p` and (as a side effect) writes the PDF to
# outputs/figures/. Here we just display the same object inline.
source(here::here("figures", "fig01_mito_content.R"))
print(p)
```

The compositional critique of the share panel -- "the fraction can rise because *other*
genes fall" -- is answered by the same four comparisons in per-gene log2FC space
(`figures/fig01b_mito_content_lfc.R`, from `interaction_results.rds`): the arm genes rise
in **absolute** terms and **coordinately** (Myc@6W: 83-95% of each nuclear arm's genes up,
biogenesis highest at 95%; mtDNA-encoded genes are *not* induced -- 0-15% up), so the share
rises because the genes rise, not by subtraction.

::: {.callout-important}
## Two cross-panel tensions the narrative must own (both real, not artifacts)

1. **6W: LFC says strong, share says marginal -- except on the biogenesis arm.** fig01b shows
   369 individually significant up-genes in Nuclear MitoCarta at 6W, yet fig01's *share*
   bracket at 6W is only p=0.11 there (mass markers 0.085, nuclear OXPHOS 0.12). Per-gene
   induction is large; the n=6 aggregate *fraction* is noisy at 6W, so the genes clearly move
   where the share is underpowered. **Mito biogenesis is the exception and the arm to lead
   with**: p=**0.035** at 6W and p=**0.0015** at 12W -- the only arm significant at both, and
   the same arm fig01b independently ranks most coordinate (94.7% of genes up at 6W). Share
   space and LFC space agree once the arm is a real programme rather than a sub-arm; part of
   the old tension was arm choice, not only sample size.

2. **12W: significant counts undersell the coordinate shift.** OXPHOS@12W has a clearly
   positive box (87% of genes up) but only **1** gene clears FDR, because attenuation shrinks
   per-gene magnitude and n=6 limits power at 12W -- exactly where fig01's share is *firmest*
   (p 0.003). The `% up` line on fig01b (power-independent) is what keeps 12W from looking
   artificially weak; read it alongside the significant-gene counts. The biogenesis arm makes
   the point explicit: at 12W it sits at essentially the same 88% up as OXPHOS yet retains
   **14** FDR-significant genes to OXPHOS's 1 -- the counts are tracking set size and power,
   not a difference in biology. This previews @sec-attenuation.

Integrated reading: **6W = strong per-gene induction, noisy aggregate; 12W = attenuated
per-gene signal, consolidated compartment share.**

Not a tension but the mirror of the claim: the only significant *time* bracket in fig01 is
mtDNA-encoded within Myc+ (p=0.0079, 6W->12W), matching fig01b's 85% up. It is on the
batch-confounded axis (batch = timepoint) and stays a described observation, not a result --
see @sec-lesner for why the mtDNA arm needs qPCR to settle.
:::

*(stub: the content narrative -- mass-panel coherence, where the imbalance is NOT
[the mtDNA machinery is not the transcriptional bottleneck], the three ceilings, and why
mtDNA qPCR / blot is the settling experiment. Possible future panel: the four contrasts as
the sides of a square [Myc@6W / Myc@12W / WT-time / Myc+-time] -- a 2x2 contrast diagram.)*

# The mitochondrial-group survey: what rises when OXPHOS falls {#sec-mito-survey}

**SUPPLEMENTARY FIGURE 1** is the three controls behind @sec-content, in the order the
claims are made: the compartment share rises because the *genes* rise and not by subtraction
(the compositional objection to a fraction); something has to rise while OXPHOS falls, and
the sixteen-group survey names it as the substrate-metabolism arms, so a flat compartment
total is reallocation rather than stasis; and the respiratory chain moves as one block, so
"MYC raises OXPHOS" is the right sentence rather than a claim about one complex.

```{r}
#| label: fig-compartment-detail
#| fig-cap: "SUPPLEMENTARY FIGURE 1 (figureS1). The compartment in detail. (A) Per-gene direction across the four contrasts for the Figure 1 arms; numbers inside the bars are FDR-significant genes, which track set size and power, while the % up line does not. (B) Median log2FC across the 16 MitoCarta groups; read the WT column. (C) The five respiratory complexes. Set membership resolved with the vintage reconciler, which recovers 17 of Complex V's 24 subunits that a plain symbol match drops. Genotype columns are the batch-clean axis; the temporal columns are confounded and described, not claimed."
#| fig-width: 9
#| fig-height: 8.2
#| echo: false
source(here::here("figures", "figureS1_compartment_detail.R"))
print(p)
```

The rest of this section is the exploratory version of panels B and C.

fig01 shows OXPHOS is the mitochondrial arm that *falls* across the WT 6->12W timeline,
yet the total nuclear MitoCarta share is flat -- so other groups must rise to hold it
there. This supplementary survey resolves which. It takes the seven canonical MitoCarta
MitoPathway top-level categories -- **with Metabolism split into its nine level-2 children**
(hierarchy read from MitoCarta Sheet 4) -- and reads them the same two ways as fig01/fig01b:
compartment share across the four groups (@fig-survey-share) and per-gene log2FC across the
four contrasts (@fig-survey-lfc). **OXPHOS is shown as its two halves** -- nuclear OXPHOS
(`MITOCARTA_OXPHOS_NU`, 155 genes, no mt-* genes) and mtDNA OXPHOS (the 13 mtDNA-encoded
subunits, which *are* `MITOCARTA_MTDNA_ENCODED`) -- so the nuclear category carries no mtDNA
bias and there is no redundant "mtDNA reference" facet. `share_nomt` (denominator drops only
the 13 mt-* genes) is the consistent scale.

```{r}
#| label: fig-survey-share
#| fig-cap: "Compartment share of all 16 main MitoCarta groups across the four groups (% of the non-mtDNA transcriptome, free y). Genotype is the clean axis; the time axis is batch-confounded. n=6/group."
#| fig-width: 9
#| fig-height: 8
#| echo: false
source(here::here("figures", "figS1_mitocarta_survey_share.R"))
print(p)
```

```{r}
#| label: fig-survey-lfc
#| fig-cap: "Per-gene raw log2FC across the four fig01 contrasts for the 16 groups. A: median LFC per group x contrast (n = genes at padj<0.05; -n = mostly down). B: the LFC distributions. Read the WT 6->12W column."
#| fig-width: 9
#| fig-height: 8.5
#| echo: false
source(here::here("figures", "figS1b_mitocarta_survey_lfc.R"))
print(p)
```

The answer: in the WT timeline **nuclear OXPHOS is the outlier** -- only ~38% of its genes up
(6 individually significant *down*) -- while the biosynthetic and substrate-metabolism groups
*rise* (sulfur 80%, carbohydrate 77%, amino-acid 76%, lipid 68% of genes up; metals &
cofactors, detoxification, small-molecule transport and electron carriers all >55%). The flat
total is therefore a **reallocation**, not stasis: as the developmental nuclear-OXPHOS
programme winds down over the timeline, substrate-metabolism groups take up the share. Note
that **mtDNA OXPHOS moves opposite to nuclear OXPHOS** -- it rises over the timeline (@fig-survey-lfc
panel A) -- the same nuclear-down / mtDNA-up mitonuclear discordance discussed in @sec-lesner,
though here on the batch-confounded time axis and so described, not claimed.

Two guards on reading this. (1) The WT-timeline columns sit on the batch-confounded axis
(batch = timepoint), so the rise is **described, not claimed** -- the clean read is the
genotype columns, where Myc induces essentially *every* group at 6W and that induction fades
by 12W (attenuation, @sec-attenuation). (2) The median-LFC fills and the significant-gene
counts track set size and power; the direction (% up) is the power-independent read.

The companion table lists, per group, the genes whose LFC varies most across the four
contrasts (`spread = max - min`) -- the context-dependent movers behind each group's summary
(full CSV at `outputs/figures/figS1b_diverse_genes.csv`).

```{r}
#| label: tbl-diverse-genes
#| tbl-cap: "Most divergent genes per MitoCarta group (top by LFC spread across the four contrasts)."
#| echo: false
knitr::kable(diverse_genes, digits = 2,
             col.names = c("group", "gene", "Myc@6W", "Myc@12W", "WT 6->12W",
                           "Myc+ 6->12W", "spread", "sd"))
```

# OXPHOS, complex by complex {#sec-oxphos-complexes}

The survey isolates **nuclear OXPHOS** as the arm that rises with Myc and falls across the
WT timeline. Does one respiratory complex carry that, or all five together? This drills into
the five nuclear-encoded complexes (Complex I-V; the 13 mtDNA subunits are the separate mtDNA
OXPHOS facet) the same two ways -- share (@fig-cplx-share) and per-gene LFC (@fig-cplx-lfc).

```{r}
#| label: fig-cplx-share
#| fig-cap: "Transcriptome share of each respiratory complex across the four groups (% of the non-mtDNA transcriptome, free y). Brackets = genotype/ time p<0.05 only. n=6/group."
#| fig-width: 9
#| fig-height: 4.2
#| echo: false
source(here::here("figures", "figS2_oxphos_complex_share.R"))
print(p)
```

```{r}
#| label: fig-cplx-lfc
#| fig-cap: "Per-gene raw log2FC of the five complexes across the four contrasts. A: median LFC per complex x contrast (n = genes at padj<0.05). B: LFC distributions."
#| fig-width: 9
#| fig-height: 5.8
#| echo: false
source(here::here("figures", "figS2b_oxphos_complex_lfc.R"))
print(p)
```

The reading is **uniformity, not selectivity**: Myc raises all five complexes coordinately at
6W (the share is genotype-significant for CI, CIII, CIV and CV; CII -- 8 nuclear genes, the one
complex with no mtDNA subunit -- is the underpowered exception), and every complex attenuates in
step by 12W. No complex is singled out; the oncogene drives the respiratory chain as one block,
consistent with the coordinate biogenesis reading of @sec-content.

::: {.callout-note}
## Mapping note (why Complex V needed care)
MitoCarta's complex sets carry the **old ATP-synthase gene names** (`Atp5a1/b/c1/...`), while
the count matrix uses the **current** ones (`Atp5f1a..`, `Atp5mc*`, `Atp5pd/pf/po`). A
current-symbol match silently drops 17 of Complex V's 24 genes. These figures therefore resolve
membership through MitoCarta's **own Ensembl IDs** (Sheet 2) rather than by symbol, recovering
CV 24/24. They are computed self-contained (not via script 32), so fig01/figS1 are untouched.
:::

```{r}
#| label: tbl-cplx-diverse
#| tbl-cap: "Most divergent genes per complex (top by LFC spread across the four contrasts)."
#| echo: false
knitr::kable(diverse_genes, digits = 2,
             col.names = c("complex", "gene", "Myc@6W", "Myc@12W", "WT 6->12W",
                           "Myc+ 6->12W", "spread", "sd"))
```

# Myc reprioritises the compartment {#sec-reallocation}

The content figures use one ruler -- absolute abundance. mitoPPS (Monzel 2025) adds the
second: it is **content-blind**, reading *relative* resource reallocation *within* the
compartment (each MitoPathway is ratio-normalized to 1; the pairwise ratio cancels the
per-sample common mode, mitoPPS ambient ~0.42 vs GSVA's ~0.80). The main point is simply that
**reprioritisation happens** -- Myc does not merely raise the compartment, it reshuffles
priority within it, coherently by pathway tier and at both ages (@fig-reallocation).

```{r}
#| label: fig-reallocation
#| fig-cap: "MitoPathways ranked by their Myc mitoPPS effect (Myc+ - WT, 6W), promoted (left) to demoted (right); one dot per pathway coloured by tier, * = padj<0.05, the 7 top-level categories labelled, and 12W shown as a small point joined to 6W (the fade). Reprioritisation is structured by function. n=6/group."
#| fig-width: 9
#| fig-height: 4.7
#| echo: false
source(here::here("figures", "fig02_reallocation_ranked.R"))
print(p)
```

The ranking is **structured by pathway function**, not random: the **demoted** end is dominated
by dynamics & surveillance, apoptosis and signalling (and the mtDNA-encoded OXPHOS arm at the
very bottom), while the **promoted** end gathers biosynthetic metabolism and the import
machinery (SAM, chaperones). The seven top-level categories (labelled) span the whole range --
dynamics most demoted, import/central-dogma most promoted. **OXPHOS is visibly *split***: its
structural subunits sit high (promoted) and its assembly factors low (demoted), which is why
OXPHOS lands ~neutral as a whole tier. This is the **clean genotype axis**, but **descriptive /
exploratory** (n=6/group): the claim is the ordered, tier-coherent pattern, not any single pathway.

The fully-labelled ranking -- every pathway named, with 6W and 12W dots and their connectors --
is @fig-reallocation-full.

```{r}
#| label: fig-reallocation-full
#| fig-cap: "Supplementary (figS4): every MitoPathway (~144) ranked by Myc mitoPPS effect, fully labelled; filled = 6W, open = 12W, line = the 6W->12W change. The full version of @fig-reallocation. n=6/group."
#| fig-width: 8.5
#| fig-height: 12
#| echo: false
source(here::here("figures", "figS4_reallocation_full.R"))
print(p)
```

The pathway-level detail behind this -- and the one place the two rulers *disagree* -- is in
@fig-reallocation-detail: across all ~148 pathways, absolute content (DESeq set log2FC) is
positive almost everywhere, yet within **OXPHOS** the structural *subunits* are promoted in
priority (CIII-subunits mitoPPS +0.20, CV-subunits +0.17) while the *assembly factors* are
demoted (CII −0.26, CIII −0.14 padj 0.022, respirasome −0.11) and the mtDNA arm falls (−0.34) --
all on positive content. That subunit-over-assembly split is why OXPHOS reads ~neutral at tier
level, and it only surfaces at all-pathways granularity.

```{r}
#| label: fig-reallocation-detail
#| fig-cap: "Supplementary: content (absolute set log2FC) vs priority (relative mitoPPS) Myc effect across all ~148 MitoPathways, tier-blocked. The colour flip -- red on the left, blue on the right -- is the reallocation, sharpest within OXPHOS (subunits vs assembly). n=6/group."
#| fig-width: 8.5
#| fig-height: 13
#| echo: false
source(here::here("figures", "figS3_reallocation_two_rulers.R"))
print(p)
```

Collapsing the two rulers into one scatter with marginal distributions
(@fig-reallocation-scatter) makes the logic plain. **Content is one-sided**: ~94% of pathways
rise (the top marginal is a single hump right of zero) -- Myc raises nearly the whole
compartment. **Priority is wide**: it straddles zero, only ~56% up (the right marginal). That
wide priority spread, sitting on a near-uniform content increase, *is* the reprioritisation.
Two features stand out. (1) The reallocation is **directional**: the content-down / priority-up
quadrant is **empty** -- nothing that falls in content gains priority, so the demoted pathways
are not switched off, they simply rise *slower than the biogenesis machinery* and lose relative
share. (2) The priority extremes name candidates: **most promoted** are the import chaperones and
SAM, the immune response and anaplerotic metabolism (creatine, pyruvate); **most demoted (raised
yet)** are the OXPHOS assembly factors, cAMP-PKA signalling, the dynamics/apoptosis pathways and
the mtDNA-encoded arm. No trend line is drawn: mitoPPS is a ratio (~content relative to the
compartment mean), so a content/priority slope is partly built in. Descriptive/exploratory
(n=6/group).

```{r}
#| label: fig-reallocation-scatter
#| fig-cap: "Supplementary (figS5): priority (relative mitoPPS effect, x) vs content (set-average log2FC, y) per pathway at 6W, with marginal densities; the 7 top-level categories labelled. Content is one-sided-positive (~94% up), priority wide both ways (~56% up) -- that spread is the reallocation. Ringed = padj<0.05. n=6/group."
#| fig-width: 9
#| fig-height: 6.6
#| echo: false
source(here::here("figures", "figS5_reallocation_scatter.R"))
print(p)
```

## Reallocation over the timeline {#sec-reallocation-time}

The reallocation above is the Myc *genotype* effect within a timepoint. A **separate**
phenomenon plays out over the timeline: the **WT background itself reprioritises** as the tissue
develops. Ranking the pathways by their WT 6->12W mitoPPS change (@fig-reallocation-time) shows a
full developmental reallocation in the WT alone -- and the dropline to the Myc+ 6->12W change
shows where Myc **bends** that trajectory. The two are largely shared (WT-time and Myc+-time
mitoPPS correlate r=0.75, 70% same direction, @fig-reallocation-time-div): most of the temporal
reallocation is developmental, common to both genotypes. Where Myc departs is the **fade** -- the
biogenesis / biosynthesis and OXPHOS arms it promoted at 6W lose relative priority hardest by 12W
in Myc+ (they sit *below* the y=x line), the mitoPPS reading of the attenuation (@sec-attenuation).
The apoptosis / mitophagy pathways move with the developmental background rather than diverging --
i.e. no Myc-specific temporal priming signal in this lens.

::: {.callout-warning}
The **time axis is batch-confounded** (batch = timepoint), so the individual WT-time and Myc+-time
changes are **descriptive, not claims**. The clean quantity is their **divergence** -- the dropline
in @fig-reallocation-time, the deviation from y=x in @fig-reallocation-time-div -- because the
batch effect cancels in the difference. Both are exploratory (n=6/group).
:::

```{r}
#| label: fig-reallocation-time
#| fig-cap: "Supplementary (figS6): MitoPathways ranked by WT 6->12W mitoPPS change (filled), with a dropline to Myc+ 6->12W (open) -- the Myc-specific bend. The 7 top-level categories labelled. n=6/group."
#| fig-width: 9
#| fig-height: 5
#| echo: false
source(here::here("figures", "figS6_reallocation_temporal_ranked.R"))
print(p)
```

```{r}
#| label: fig-reallocation-time-div
#| fig-cap: "Supplementary (figS7): WT 6->12W vs Myc+ 6->12W mitoPPS change per pathway. On the y=x line = shared/developmental (r=0.75); off it = where Myc bends the WT trajectory; below-diagonal = the biogenesis arms Myc fades. n=6/group."
#| fig-width: 6.5
#| fig-height: 6.8
#| echo: false
source(here::here("figures", "figS7_reallocation_temporal_divergence.R"))
print(p)
```

# The global axis and the correlation ceiling {#sec-axis}

Every claim this document defends is a **contrast** -- a genotype gap, an interaction, a
difference of differences. Not one is a per-sample correlation between two pathway scores.
This section is the reason, and it belongs before the reader meets a coupling anywhere in the
exploratory layer: in this dataset a correlation of 0.8 between two composites is the
*default*, not a finding.

**Score the whole library and it is almost one-dimensional.** Across the 885 sets scored per
sample, PC1 takes **77%** of the variance (PC2 8%, PC3 4%), effective dimensionality is
**1.7** of a possible 23 at n=24, and **79% of set loadings share one sign** -- a common mode,
not a contrast. PC1 is essentially the per-sample mean of all 885 scores (r=0.98), and nuclear
OXPHOS tracks that mean at **r=0.972**. The library is not resolving 885 programmes. It is
resolving roughly one, plus departures from it.

**What that does to a coupling.** Arbitrary set-to-set pairs already sit at median |rho|
**0.55**, and the ambient coupling of `mito_oxphos` to an arbitrary library set is **0.80**.
Three explanations one would reach for first all fail. It is **not low n**: permuting the
axis's own sample labels puts the chance floor at **0.12**. It is **not curation**: arbitrary
pairs are at 0.55 before any curation is involved. It is **not the genotype contrast**: within
`6W_pos` alone -- one genotype, one age, the contrast gone -- the ambient is still **0.83**.
And it is **removable**: regress the per-sample global mean out of every score and the ambient
falls to **0.13**, which is chance.

**The axis is not an artifact to delete; it is a phenotype.** The raw PC1 correlates with the
four-group label at r=0.47, so a large part of the raw ceiling *is* the design (much of that
MYC). Remove the design and a within-group axis of the same kind remains -- residual PC1
**71%** of variance -- and its strongest correlates are epithelial marker share **+0.79**,
contamination **-0.70** and proliferation **+0.65**. Biological state and prep tells load
together, so the axis is neither cleanly biological nor cleanly technical; with no
dissociation-batch, viability or RIN metadata on disk, the two are not adjudicable. Report it
as a phenotype, and remove it before asking a specific question.

**Where it comes from at gene level -- and the flip that matters.** The largest gene-level
axis (top-500 VST PC1, 44% of gene variance) is **genotype-independent** (rho +0.02): it is
the epithelial-purity versus immune-infiltration composition axis (epithelial +0.84, immune
-0.90), i.e. residual dissociation contamination. Its metabolic pole is cell **identity** --
epithelial MECs are mitochondria-dense, immune infiltrate is not -- and not MYC dose, which is
why the design correctly drops it. Gene-PC2 (13%) is the MYC/E2F/proliferation programme
(proliferation +0.76, MYC +0.49, genotype +0.26) with OXPHOS and mitochondrial translation
baked in. The pathway axis tracks **PC2 more than PC1** (+0.60 against +0.44) and *sharpens*
the genotype signal from +0.26 to **+0.63**. So compositing foregrounds the Myc-mito programme
and demotes the nuisance: **mitochondria ride inside the MYC programme rather than beside
it**, which is this document's entry point, not a caveat to it. The gene-level root of the
ceiling sits in the same sentence -- OXPHOS enriches at *both* poles.

::: {.callout-important}
**The interpretability trap -- check this before reading any global-adjusted coupling.** The
mitochondrial axes **are** the global factor. `mito_oxphos` carries r2 = **0.94** with the
per-sample global mean, so after design -> PC1 -> residual only ~4% of its variance remains,
and that remainder is noise. Its global-adjusted "coupling" to proliferation prints **-0.65**
and is sign-consistent across all three quantifiers -- and it is **not antagonism**, it is
noise on noise at n=24. A global-adjusted coupling is readable only where the axis retains
roughly **10% or more** residual variance. Three of the mitochondrial pairs fail that outright
(`mito_oxphos` to proliferation, to TEB-dedifferentiation, to priming: residual fraction
0.038-0.039) and two more are borderline.
:::

**What survives is not "OXPHOS is central".** That proposition fails in the strongest available
form: not merely unsupported but **untestable**, because OXPHOS has essentially no variance
independent of the broad state, so no specific coupling exists to test in either direction.
The positive reframe is the robust half. As a composite, OXPHOS **is** the dominant axis
(design-residual loading **0.969**) -- but it is indistinguishable from a tightly co-loading
cluster: TCA 0.968, nucleotide 0.946, MYC signatures 0.934, biogenesis 0.911, proliferation
0.863. All of them essentially *are* the axis, which is exactly why nothing in the cluster is
separable from the rest. The dominant programme is that whole anabolic-proliferative-
mitochondrial core, not OXPHOS alone. Read as a **level** this is near-tautological and
therefore safe -- it is a description of the axis: OXPHOS member sets sit at the **78th
percentile** of all 885 absolute loadings, MitoCarta as a whole loads above the rest (Wilcoxon
p=0.015), and the **redox** control -- the one axis MYC does not drive -- sits at the 43rd
(loading 0.26).

The framework's one non-tautological move is the loading **change** by genotype
(`S ~ G*genotype*timepoint`), a contrast the axis itself cannot touch. Within groups MYC hints
at *tightening* the metabolic core's alignment (OXPHOS +58%, p=4.3e-05; TCA +42%, p=0.008)
while *loosening* proliferation's (-56%, p=0.012). That is a lead and not a result: the slopes
are near-collinear at n=6/group and `mito_biogenesis` shows no change at all (p=0.22).

**Mito-led, not mito-specific** -- and specifically not an argument for mitochondria-*primary*
regulation. Mito-defined sets are 44% of the library but **100% of the top 100 loaders**. The
enrichment is real and it is also inflated by construction: the top 100 splits MitoCarta-proper
**6** / build-tautological `_MITO` lanes **94** / non-mito **0**, and `TFT_MYC_GRAY_BA_LE_MITO`
loads 0.993 (rank 6) while the *same* transcription factor's non-mito lane
`TFT_MYC_GRAY_AP_HE` loads 0.014 (rank 883). Loading therefore tracks mitochondrial gene
**content**, not the regulator. Non-mitochondrial growth programmes load nearly as high
(METABRIC MB2 bicluster and mammary-development lanes at 0.97-0.98) against the redox control
at 0.26. Loading is covariation; primacy needs perturbation.

**The one lead that survives the correction is not a mitochondrial one.** `redox -> MB2
tumorigenic fork`: global-adjusted **-0.63**, same sign and -0.62 to -0.73 across all three
quantifiers, on an axis that barely loads on the global factor (r2 = 0.03, so ~92% of its
variance survives the correction) and that MYC does not drive. `cholesterol/mevalonate -> fork`
is **demoted** to +0.14 (range +0.14 to +0.27), "separable but weak" -- its earlier
96th-percentile headline was a raw-ambient artifact. And the negative control behaves as a
negative control should: `mito OXPHOS -> death priming` is untestable, its sign flipping
between -0.33 and +0.33 across methods. @sec-leads carries the bench versions.

::: {.callout-note}
**Method of record, and scope.** The primary quantifier is the **linear mean gene-wise
z-score** -- its correlation *is* average cross-gene covariance -- with singscore for
cohort-independent robustness and GSVA for continuity with the earlier analyses. The
correction is **design -> PC1 -> residual with the factor estimated after the design**, plus
leave-pair-out; it is **not** mean-subtraction, which produces an equal-loading, sum-to-zero
artifact. Gene-level and score-level design adjustment agree to ~0.02. A second, technical
lens (contamination and IEG anchors removed) correlates r=0.57 with the primary design-only
axis, so treating those as technical does not overturn the biological reading -- but the
anchors are RNA-derived and absorb r2=0.67 of design-residual *proliferation* variance, so
that lens removes biology along with prep noise and the design-only lens stays primary.
**Scope:** this section re-ranks the **exploratory couplings only**. The content result (raw
count shares, @sec-content) and the attenuation (DESeq2 log-fold-changes, @sec-attenuation)
never pass through GSVA or any of these couplings and are untouched by everything here.
Nothing in this section beats BH < 0.05 at n=24; it is a ranking.
:::

Sources on disk: `results/pathway_loading.rds` (script 37), `results/linear_pathway_coupling.rds`
(36) and `results/ambient_corrected_couplings.rds` (35), with panels in
`outputs/pathway_loading/` (`A_scree_dimensionality`, `A2_sample_pca_gene_vs_pathway`,
`A3_gene_axis_covariates`, `B_axis_loadings`, `D_loading_enrichment_mito`) and
`outputs/linear_pathway_coupling/` (`A_ceiling_origin`, `B_global_factor_phenotype`,
`C_coupling_hierarchy`). Those panels are the analysis layer's own, not promoted to the
paper figure set. The method notes this section implements are
`docs/GSVA_global_background_and_pathway_coupling.md`,
`docs/Gene_set_quantification_for_pathway_correlation.md` and
`docs/Low_dimensional_pathway_structure_and_global_axis_analysis.md`; the five-question
synthesis is `docs/2026-07-18_narrative_synthesis_five_questions.md` (sections 0.5, 0.6).

# The oncogenic programme attenuates across progression {#sec-attenuation}

**SUPPLEMENTARY FIGURE 2** carries the four controls @sec-window depends on. PUMA's collapse
is read against a global x0.55 rescaling, so before that reading is allowed: the x0.55 is
**not** a falling dose (every measure that would register one *widens* while MYC's output
narrows); it **is** real and uniform (the 12W profile is the 6W profile at 55% amplitude, its
coherence at the 100th percentile of matched shuffles); what MYC adds over the window is a
**downward shift that is near-constant across pathways** and is the batch-clean quantity; and
PUMA's collapse **survives removal of the BCL-xL denominator** on a genome-wide statistic.

That third control is the one worth stating in plain terms, because it is easy to overstate
in either direction. Across the window the wild-type gland drifts *up* (median +0.04; 103 of
143 pathways rise) while the Myc gland drifts *down* (median -0.13; 22 of 143 rise). Per
pathway, **Myc+ minus wild-type is negative in 133 of 143 cases, median -0.18** -- and that
difference *is* the Myc-specific temporal change, batch-clean because the shared offset
cancels in a difference. Decomposed as a regression (the panel's inset), the **shared** slope
of 0.75 is **not** distinguishable from its null (82nd percentile), while the MYC-specific
intercept of -0.17 lies beyond all 500 shuffles. So a large part of the Myc gland's movement
*is* shared with the wild-type's -- that part is simply unremarkable -- and what the wild-type
change cannot account for is the extra downward component.

```{r}
#| label: fig-controls
#| fig-cap: "SUPPLEMENTARY FIGURE 2 (figureS2). The controls behind Figure 2. (A) Driver against output: everything registering a falling MYC dose widens over the window while the genotype gap in Hallmark MYC targets V2 narrows, so the attenuation is downstream of dose. (B) The 12W MYC effect is the 6W effect at ~55% amplitude. (C) Myc+ minus wild-type per pathway, ranked: 133 of 143 move further down with MYC, median -0.18, and the flat band is the near-uniformity shown rather than inferred; the inset decomposes the same thing into a shared slope (not beyond its null) and a MYC-specific intercept (beyond all 500 shuffles). (D) The collapse statistic over 8,774 MYC-responsive genes, marking the two pre-specified genes and Myc itself. The nulls test against the data's own structure, not against sampling new animals."
#| fig-width: 9
#| fig-height: 8.1
#| echo: false
source(here::here("figures", "figureS2_controls.R"))
print(p)
```

Foreshadowed already in fig01b (@sec-content, tension 2): the per-gene induction shrinks
6W->12W (OXPHOS still 87% up at 12W but only 1 gene clears FDR), and the Time-Myc+ contrast
is net **down** (Nuclear MitoCarta 68 down vs 20 up) -- the fade, in the same panel. The
question this section settles is *why*: does the Myc effect fade because the **background**
the oncogene acts on has moved, or because the **programme itself** decays? @fig-background
answers it in four panels, all built on one exact identity -- the change in the genotype gap
equals (Myc+ temporal change) - (WT temporal change), which holds to machine precision on
both the content and the priority ruler.

**First, the trivial explanation: the driver does not fade** (@fig-myc-dose). Myc's own
genotype gap *widens* over the window, **+1.64 log2 at 6W (p=2.5e-4) to +1.80 at 12W
(p=9.6e-6)**, and no within-genotype decline could carry the fade (Myc+ 8142 -> 7578
normalised counts, p=0.65). The WT baseline itself drifts down borderline (2543 -> 2143,
p=0.052), but that sits on the batch-confounded axis -- which is precisely why the gap, the
batch-clean quantity, is the one read. The proximal MYC/MAX/MXD network is flat: across MAX,
the MXD1/3/4-MXI1-MNT-MGA repressors and the MLX arm, 40 simple-effect tests give 5 at nominal
p<0.05 and **none survive BH** (min 0.052), with no genome-wide interaction padj<0.05 for any
of them -- no repressor rises to titrate MAX away.

The contrast that settles it is driver against output. **Every** measure that would register a
falling dose widens over the window -- Myc **+0.17**, Myc:MAX **+0.32** (+1.42 -> +1.74),
Myc:repressor-sum **+0.25** (+1.65 -> +1.90) -- while Myc's *output*, the genotype gap in
Hallmark MYC targets V2, **narrows by -0.36** (+1.32 -> +0.95). Driver up, output down: the
attenuation is **downstream of dose**, and the rest of this section is the decomposition of
that downstream part. Two bounds on the claim. Reads map to the mouse *Myc* locus, so total
message in Myc+ is transgene plus endogenous and the two are not separable here (script 27
decomposes them at programme level) -- "constant dose" is a statement about total message,
which is what the substrate sees. And transcript is not protein: the 31-gene post-translational
stability panel (`outputs/myc_stability_panel/`, script 41) finds nothing Myc-specific (every
interaction padj>0.05), no excess movement over an expression-matched background, and
directions that predict a *more* stable MYC (Fbxw7 -0.35 and Huwe1 -0.24, the two established
degradative E3 ligases, both down) -- but that is absence of transcriptional evidence, not evidence of
absence. The settling experiments are protein-level: a MYC blot at 6W vs 12W, phospho-T58/S62,
and a cycloheximide chase.

```{r}
#| label: fig-myc-dose
#| fig-cap: "figS8: is the Myc dose constant? (A) Myc normalised counts across the four groups. (B) The proximal MYC/MAX/MXD network -- MAX, the MXD/MNT/MGA repressors and the MLX arm; brackets unlabelled, red = p<0.05. (C) Everything that would register a falling dose (Myc, Myc:MAX, Myc:repressor sum) against Myc's output (Hallmark MYC targets V2), as the genotype gap at each age; open = 6W, arrow to 12W. The driver gap widens while the output gap narrows -- the attenuation is downstream of dose. n=6/group."
#| fig-width: 9
#| fig-height: 7.4
#| echo: false
source(here::here("figures", "figS8_myc_network_levels.R"))
print(p)
```

**Myc's effect is rescaled, not reshaped.** Regressing the 12W Myc effect on the 6W effect
across the 144 MitoPathways gives **slope 0.55** (bootstrap 0.51-0.60), R2 0.80, rho 0.84 on
the content ruler and slope 0.64, R2 0.79, rho 0.93 on priority, with per-tier slopes
0.53-0.83 -- uniform (@fig-background B). The coherence sits at the **100th percentile** of
gene-label shuffles matched on set size, expression and pathway overlap. The 12W profile *is*
the 6W profile at 55% of the amplitude: the same pathways, in the same order, less of it.
This is the pathway-level form of the Issue #4 result that magnitude, not rank, shrinks.

**What Myc adds over the timeline is a downward shift, near-constant across pathways, rather
than a different set of pathways moving.** Regressing the Myc+ 6->12W change on the WT 6->12W
change (@fig-background C) gives slope 0.75 and intercept -0.17. The **slope** -- the part
both genotypes share -- is not distinguishable from the null (null median 0.63, 82nd
percentile). The **intercept** is, beyond all 500 shuffles. It is also the only batch-clean
quantity in the comparison, since an offset common to both temporal contrasts cancels in
their difference.

Two phrasings to avoid here, both tempting and both wrong. **"A uniform offset"** would be
exact only at slope 1; at 0.75 there is also a component scaling with how far the wild-type
moved, so the defensible claim is that the *intercept* is the extreme, batch-clean term, not
that the offset is literally constant. And **"the Myc change is not due to the WT change"** is
too strong the other way: slope 0.75 says much of the movement *is* shared, and the point is
that the shared part is unremarkable while an additional component is not.

**The background is not moving toward the Myc state.** The WT 6->12W vector projects onto the
Myc(6W) vector at just **0.064** -- against a null median of 0.24, the 9.8th percentile, i.e.
*less* than an arbitrary matched panel. Restricted to the biogenesis and OXPHOS arms the cos
angle is **-0.34** (content) / -0.39 (priority): the background moves *away* from where Myc
pushes. That aggregate is not homogeneous, and the per-tier breakdown is informative -- protein
import (**-0.63** content, **-0.77** priority) and OXPHOS (**-0.65**) move against Myc, while
mitochondrial central dogma (**+0.74**) and metabolism (+0.48) move with it. The Myc+
trajectory, by contrast, has cos angle **-0.84** with the Myc effect (0th percentile), and
-0.80 to -0.98 in every substantial tier: over the timeline the Myc+ compartment retraces the
Myc programme backwards. @fig-background A shows the same thing per tier -- a red Myc arrow
that keeps its direction and shortens, and a grey background arrow that points elsewhere.

**And the WT-convergence term does not survive its control.** The published decomposition
aligns each gene to Myc's induction direction and reads the WT temporal change along it -- but
the genotype contrast and the WT temporal contrast **share the 6W WT baseline with opposite
signs**, so noise in that baseline manufactures apparent convergence. The six 6W_neg mice are
therefore split over all 20 three-versus-three partitions -- and, because halving the baseline
also adds noise to the alignment direction (which pulls the statistic toward chance from either
side), each partition is run twice off the *same* half baseline, once shared between the two
contrasts and once split. Identical direction, identical universe, identical noise; the only
difference is the sharing.

Globally, sharing the baseline adds **+0.24** (sd 0.138) of apparent convergence, in 19 of the
20 partitions: `frac_wt_toward` falls from 0.720 to **0.478** -- chance -- and the percentage
from +61% to -12% (@fig-background D). And the artifact is **absent exactly where the published
analysis reported divergence**: +0.01 on the MYC-target core, -0.01 on OXPHOS subunits, so
their divergence stands as published. What survives on the other side is programme-specific:
the biosynthetic arms genuinely converge (amino acid 0.74 -> **0.65**, lipid 0.77 -> **0.63**,
TCA and nucleotide ~0.63-0.64). The programme-specificity was the whole result; the global
number was the artifact.

So the picture is **additive, not causal**: a reallocation of the mitochondrial compartment
that both genotypes undergo, plus a Myc programme that decays proportionally without changing
shape. The background's own move is 57-62% as large as the Myc effect and points in a
direction the Myc effect does not.

One number the section deliberately does not give is a single size for the attenuation. On
the same gene sets, the share ruler retains 77-83% of the genotype gap at 12W while the
set-average LFC rulers retain 54-63% -- the sign and direction agree, the magnitude is
normalisation-dependent (denominator: per-sample transcriptome sum vs median-of-ratios size
factors). The manuscript quotes the range.

::: {.callout-warning}
**Batch = timepoint.** Both temporal contrasts carry the same batch offset. It cancels in the
interaction -- @fig-background C's intercept, and the whole rescaling result -- but **not** in
the convergence/fade split, so calling the shared temporal vector "developmental" rather than
"batch" remains an interpretation. The sample-split removes the shared-*baseline* artifact; it
cannot remove the batch confound. Separately, a family of tempting statistics here is
**algebraically forced** and carries no information: cor(WT-time, change in Myc effect) is
-0.19 (content) / -0.33 (priority) and the value the marginals force is *exactly* the same,
because the change is the difference of the two temporal contrasts. The full list is in
`results/background_vs_myc.rds$artifact_ledger`. None of them may be cited.
:::

```{r}
#| label: fig-background
#| fig-cap: "fig03: is Myc's mitochondrial effect shaped by a changing background? (A) Each Level-1 MitoPathway tier as four states in priority x content; red arrows = the Myc effect at 6W (solid) and 12W (dashed), grey = each genotype's temporal move. (B) The 12W Myc effect is the 6W effect x0.55 -- rescaled, not reshaped. (C) Myc+ vs WT temporal change: the shared slope is not beyond the null, the Myc-specific intercept is. (D) WT-convergence on the matched control -- the same three-mouse 6W WT baseline shared between the two contrasts (open) vs split (filled), 20 partitions; tick = the published six-mouse value. The global term collapses to chance; the biosynthetic arms survive and OXPHOS/MYC-core divergence is untouched. n=6/group, exploratory."
#| fig-width: 9
#| fig-height: 8.6
#| echo: false
source(here::here("figures", "fig03_background_vs_myc.R"))
print(p)
```

Full numbers, nulls and the artifact ledger: `docs/2026-07-24_background_vs_myc_interpretation.md`
(script 40). The earlier per-gene decomposition (script 31) is unchanged on disk; script 40
tests it and reports the correction.

# Development opens the window {#sec-window}

**MANUSCRIPT FIGURE 2.** MYC builds that organelle on a substrate that is itself moving, and
what the substrate does over the same window is the other half of the story.

## The gland withdraws respiration, and only respiration

Between 6 and 12 weeks the wild-type gland withdraws from respiration **and from nothing else
that would explain it**. OXPHOS subunits fall **-0.255** in set-average log2FC, at the **0th
percentile** of 2000 expression-matched random sets, while the proliferative programme barely
moves (`PROLIF_*` pooled **-0.047**) and the **assembly factors of the very same complexes**
sit at **+0.001, the 50th percentile** (@fig-window A). Nucleotide metabolism and the
mitoribosome -- the two mitochondrial arms most tightly coupled to growth, chosen in advance
as the alternative hypothesis's best representatives -- do not move either. The steepest fall
in the whole comparator is the **TEB-versus-ductal signature, -0.422**: the terminal end bud
regresses, and the TEB is the apoptosis-loaded compartment, because duct cavitation is
achieved by apoptosis.

The claim is comparative, so it is tested as a comparison. `PROLIF_*` pooled is itself beyond
its own null (1.3rd percentile) -- it is *small*, not immobile -- so a one-set null cannot
carry the sentence. Redrawing **both** sets of a contrast together puts
OXPHOS-minus-proliferation at **-0.208, the 0th percentile** (@fig-window B). Nor is it an
artefact of the content rise: the withdrawal persists on mitoPPS, a within-compartment ratio
that is blind to how many mitochondria a cell has (@fig-window C). **The gland de-respires
without de-proliferating.**

## And it leaves the death machinery intact

Of the 25 pro- and 7 anti-apoptotic MitoCarta transcripts plus five non-MitoCarta brakes,
exactly **one** moves across the window at padj<0.05 -- and it moves *up* (`Bnip3` +0.63). No
brake is raised either: `Bcl2` padj 0.45, `Bcl2l1` 0.79, `Mcl1` 0.44, `Xiap` 0.79,
`Birc2/3/5` all non-significant (@fig-window D). **Development removes the input, not the
machine.** That is why what follows is a *competence* argument rather than an abundance one.

## What MYC therefore loses

Every arm of the MYC programme is rescaled by about **x0.55** across the window
(@sec-attenuation). Read against that rate, the `Bax:Bcl-xL` ratio retains **0.55** exactly --
the **50th percentile** of matched gene pairs conditioned on the same starting effect, i.e.
entirely unremarkable. The `PUMA:Bcl-xL` ratio does not fade: it **reverses**, from +0.674 at
6W (p 0.009) to -0.060 at 12W, a retention of **-0.09** at the **92nd percentile** of that
conditional null (interaction p 0.037; p 0.017 with epithelial and immune composition in the
model) (@fig-window E).

That result reproduces on a statistic that never touches the `Bcl2l1` denominator. Ranking all
8,774 MYC-responsive genes by their residual from the global rate puts `Bbc3` at the **0.51st
percentile** -- and `Bbc3` was **pre-specified** from the PGC-1a westerns, not found by
scanning (script 44; drawn in `figures/fig05_death_arm.R` panel D, which becomes Supplementary
Figure 2). Both halves of that prediction are reported and they split: `Bcl2l11`/BIM sits at
the **91.6th percentile**, so the in-vivo collapse is PUMA-specific. `Myc`'s own transcript
sits at the top of the same distribution (retention 1.09): the transgene message does not
attenuate, so the downstream fade is not transcript-level dose.

Two things this collapse is *not*. It has **no wild-type definition** -- the wild-type
`Bbc3:Bcl2l1` ratio drifts *up* (+0.168) and the whole effect sits in the genotype gap, the
batch-clean quantity. It is a loss of **inducibility**, not of level. And **PUMA is a solo**:
none of six pre-registered questions returns a module. What loses MYC-inducibility genome-wide
is luminal lineage identity (`MG_HS_GRAY` NES **-2.79**, padj 1.7e-17); what keeps it is the
MYC programme itself.

```{r}
#| label: fig-window
#| fig-cap: "MANUSCRIPT FIGURE 2 (figure2). Development opens the window. (A) Comparator arms on the wild-type window, each against its own expression-matched null; the assembly factors of the same complexes are the internal control. (B) The paired null on the difference, which is the form the comparative claim takes. (C) Content against priority: the withdrawal survives a content-blind ruler. (D) The apoptotic and brake transcripts across the same window: one moves, and it rises. (E) Retention against the x0.55 global rate, with the conditional matched-pair null. (F) The three propositions the mouse hands to the cell experiments, each with its control and its adjudicating experiment. Wild-type only in A-D, n=6/timepoint; F is a ranking, not a test."
#| fig-width: 9
#| fig-height: 10.7
#| echo: false
source(here::here("figures", "figure2_developmental_window.R"))
print(p)
```

## The three propositions the mouse hands to the cell experiments

These numbers rank hypotheses; they do not test them. Read that way the same data licenses
three propositions, two of which the cell experiments already answer (@fig-window F).

**H1 -- respiratory state gates PUMA priming.** Whether MYC raises the PUMA:BCL-xL ratio
depends on where an animal sits on the respiratory-priority axis: the interaction lands at the
**91.9th percentile** of a within-timepoint permutation null, while the same test on `redox` --
a matched mitochondrial axis MYC does *not* drive -- sits at the **50.9th, dead centre**. It is
not shared with `BAX:BCL-xL` in any readable way (67.2nd against its own control at 63.3rd).
*Adjudicated by: raising PGC-1a in MYC-driven cells.*

**H2 -- the TEB-to-ductal transition moves both together.** TEB is the steepest faller of any
arm and moderates the same ratio at the **90.8th percentile**, again PUMA-specific over BAX.
The terminal end bud both respires hardest and clears its lumen by apoptosis, so this is the
honest rival to H1 and it is **not separable from it at n=24**. *Adjudicated by: single-cell or
deconvolution, or TEB- versus ductal-derived cells -- open.*

**H3 -- respiration costs proliferation, unless death is blocked.** On the same axis, in the
same animals, MYC is *worse* at proliferating where respiratory priority is high (**6.3rd
percentile**; E2F **9.2nd**). In an animal death is intact, so the liability limb is the one
expressed. *Adjudicated by: PGC-1a in MYAZ with and without Bcl-xL -- and the sign inverts.*
**The mouse supplies the negative limb, the cells the positive one, and the switch is whether
the cell can die.**

::: {.callout-important collapse="true"}
## The bounds -- and they are severe

**Nothing in panel F is significant.** The best permutation p is 0.081. Three specific limits:
the respiratory axis is **collinear with the timepoint it would explain** (the term goes
p 0.005 -> 0.088 once `tp*myc` is in the model), which is the structure of a two-timepoint
cross-sectional design and is not repairable by statistics; H1 and H2 are not separable; and a
mediation test is uninformative here rather than supportive (adding OXPHOS makes the
interaction *stronger*, and the mediator is itself MYC-driven). One column of the source table
must be distrusted outright: adding timepoint makes the *proliferation* and composite terms
stronger while PUMA weakens, and terms that grow on adding a collinear variable are
collinearity artefacts.

**The composite version of the death result is dead three times over**, which is why the
finding is gene-level. (i) The GSVA-space mitochondria-to-death bridge is **circular** -- the
priming composite is built from `MITOCARTA_APOPTOSIS_PRO`/`_ANTI`, so it correlated
mitochondrial genes with mitochondrial genes. (ii) "MYC raises apoptotic priming" as a
composite **is the content effect**: against an expression-matched *MitoCarta* background the
pro-apoptotic set rises +0.196 where the background rises +0.364 (z -2.42), i.e. below average
for a mitochondrial gene, and on the 404 non-mitochondrial pro-death genes it does not
reproduce. (iii) The attenuation is **global**, so a composite priming signal fades because
the MYC programme fades. Nor does the gene-level result survive multiplicity: no pair beats BH
across the priming panel (minimum adjusted interaction p 0.33).

**One trap the panel encodes rather than captions:** `Bmf:Bcl-xL` sits *further* out on the
conditional null than `Bbc3:Bcl-xL` (97.5th percentile, p 0.025) but has **no 6W effect to
lose** (d6 0.24, p 0.40). It fails the precondition the whole statistic rests on.

**Batch = timepoint** bounds every wild-type temporal sentence above. Three things keep them
usable rather than uninterpretable: the comparison in panels A-B is *internal to one contrast*
(a batch offset shared by all gene sets cannot make OXPHOS subunits fall while the assembly
factors of the same complexes do not); it reproduces on two rulers with different
denominators; and the protein blots move the same way off the RNA batch entirely. Panel D's
flatness reads as "no detectable movement at n=6", not "no movement".

One caveat that is in no panel: within the 12 wild-type animals the OXPHOS and proliferation
scores correlate at **0.82**. The timepoint term survives adjustment for proliferation
(-0.499 -> -0.420, p 0.002) but at that collinearity a per-sample model cannot arbitrate. The
set-level dissociation in A-B, a contrast between gene sets rather than between samples, is
the load-bearing version.
:::

::: {.callout-note collapse="true"}
## A hypothesis about what carries the respiratory arm

Gray et al. (2023) describe a **"low-expressing" (LE)** state within each mammary lineage -- a
systematic suppression of the lineage programme, *not* an oestrogen axis; our library metadata
glosses it wrongly as "high- versus low-estrogen", recorded here so the error does not
propagate. LE states are pubertally enriched, and the ESRRA and GABPA mitochondrial regulons
are detected in the **AP (alveolar progenitor / LASP) LE** context. If the maturing gland loses
LE abundance, LE-marker sets should fall and HE-marker sets should not. In AP they do: LE
markers **-0.294** (0th percentile) against HE markers -0.049 (23rd); in BA the same. **In HS
the prediction fails** -- both arms fall -- so this is an AP/BA statement, not a global one.

Two reasons it stays a hypothesis. **25 of the 87 OXPHOS subunits are themselves AP LE-marker
genes**, so any lineage claim built on the raw sets is the OXPHOS claim restated; every number
here is from the mitochondria-removed version, and the raw self-adjusted figure (-0.044) must
never be quoted. And even mitochondria-removed, the two scores correlate at 0.93 across 12
samples. Separating a shift in state *abundance* from developmental *regulation* needs
deconvolution or single cells.

Relatedly, the withdrawal is **not** a whole-regulon event. `CORE_MITO` -- the ESRRA/NRF1/GABPA
targets intersecting MitoCarta, the closest DNA-level proxy for the PGC-1a axis -- is flat
overall (**+0.013**, 75th percentile) while its 59 OXPHOS-subunit members fall **-0.216**
(0.05th) and the other 241 rise **+0.069** (99.6th). `Ppargc1a` itself sits at baseMean 30 and
is not readable, which is why **PGC-1a is named in this paper only where the cell experiments
are described**.
:::

::: {.callout-note title="Planned experiment: MYAZ + PGC1a RNA-seq"}
Not yet done; the author intends to do it. It is **the only design that can define the
OXPHOS/apoptosis module by perturbation** rather than by correlation -- whatever PGC1a
co-induces that is both mitochondrial and apoptotic *is* the module, with no null to fight
and no correlation ceiling. In this dataset the module cannot be defined by co-expression:
the wild-type gland does not move its apoptotic transcripts at all, the PGC1a-axis regulon is
*depleted* of pro-apoptotic genes (fold 0.76) while MYC's is enriched (2.68x), and the
per-sample couplings sit at the ambient. See section 0.8 of
`docs/2026-07-26_introduction_alignment_and_the_question.md`. When the data exists it should
settle: whether the payload/trigger division of labour holds under direct PGC1a manipulation,
and why PUMA and BIM behave as a pair in cells but split in vivo.
:::

Sources on disk: `results/priming_arm_teb.rds` (script 42),
`results/substrate_specificity_tradeoff.rds` (43),
`results/collapse_module_ownership.rds` (44), and
`results/death_priming_reassessment.rds` (34, for the three composite negatives above). The
narrative and the propositions: `docs/2026-07-26_introduction_alignment_and_the_question.md`
sections 0, 0.5a, 0.7a, 0.8, 0.9, **0.10**. Blot values (MYC -50%; BAX/BIM/PUMA) and the
MYC-ER, MYAZ and Bcl-xL rescue results are the author's, reported here as given.

# External validation: Lesner et al. {#sec-lesner}

The discordance @sec-content is built on -- MYC raises the **nuclear** mitochondrial arms
while the 13 mtDNA-encoded transcripts stay flat -- has an obvious alternative explanation in
our hands. We purify MECs by warm enzymatic dissociation, and mitochondrial transcript
fraction is exactly the sort of quantity a stressful prep perturbs. Lesner et al. (bioRxiv
2026.07.13.738248) report the same discordance independently, in flash-frozen human tumours
(TCGA), a mouse HCC model, proteomics and mtDNA qPCR. That disposes of the artifact worry:
whatever our dissociation does, it cannot have produced a pattern visible in material that was
never dissociated. **They validate the question.** They do not validate our contrast -- our own
mtDNA arm sits on an axis that is genotype-independent (p ~ 0.49) and time-associated
(p = 0.0003), so it is underpowered and batch-exposed, not demonstrated.

What script 39 adds is a test of their **mechanisms** in our contrasts. The instrument is
deliberately DESeq2 contrasts -- clean genotype gaps and the Myc-specific interaction -- and
not per-sample couplings, for the reason @sec-axis gives.

**Their hypoxia -> MXI1 -> MYC-off route does not transfer.** Hypoxia is *lower* in Myc+
(6W NES **-1.34**, padj 0.010) and does not rise over the window in Myc+ (Myc+ timecourse
-1.14, padj 0.26). The positive interaction (+1.49, padj 0.003) is a **sign trap**: hypoxia
falls *less* in Myc+ than in wild-type (WT timecourse -1.58, padj 0.0018), which is not the
same as rising. At gene level `Hif1a` is low in Myc+ (-0.34), `Mxi1` is up at 6W (+0.33) but
flat over time (+0.06), and `Tfam` is flat (+0.18). One thread survives -- `Epas1`/HIF-2 rises
over time in Myc+ (interaction +0.45) -- and its predicted downstream fails: `Jun` *rises*
(+0.39) instead of being suppressed, and JUN targets are not suppressed either (interaction
NES +1.52). The HIF/MXI1/JUN axis is liver-specific.

**The OXPHOS decline does transfer, but in a different frame.** Hallmark OXPHOS is strongly
MYC-raised at both ages (+2.99 at 6W, +3.40 at 12W) and then declines over the window
Myc-specifically: Myc+ timecourse **-2.07** (padj 6.7e-09) against a non-significant wild-type
timecourse (-1.04, padj 0.33), interaction **-2.21** (padj 7.4e-11). But it falls in lockstep
with MYC targets (interaction -1.80) while hypoxia does not rise, so this is the general
programme fade of @sec-attenuation and not a hypoxia switch. One nuance has to be kept
straight, because it is where this section meets @sec-window: on the OXPHOS **subunit** set the
wild-type declines too (WT timecourse **-2.07**, padj 2.1e-05) -- the convergence component,
and the reason respiration is the one mitochondrial arm the normal gland withdraws from. The
Hallmark set gives the cleaner Myc-specific signal; the subunit set gives the developmental
one. Quoting either alone misreports the other.

**Their downstream observation transfers without the specificity read into it.** MYC raises
the mitochondrial turnover machinery at the genotype level -- DYNAMICS **+2.23** (padj
1.2e-08), FUSION +1.85, AUTOPHAGY +1.62, MITOPHAGY +1.56, FISSION +1.55 -- and it attenuates
like everything else (DYNAMICS interaction -1.81). Two things stop this being a mitophagy
programme. At gene level the specific effectors are weak (`Dnm1l`/DRP1 +0.22, `Fundc1` +0.14),
while the large movers -- `Bnip3`, `Bnip3l`, `Prkn` -- rise **equally in wild-type**, i.e. with
the batch-confounded timecourse rather than with MYC. And against an expression-matched
**MitoCarta** background the genotype rise is *below* average for a mitochondrial gene:
MITOPHAGY +0.24 against a matched +0.39 (z = -1.37), DYNAMICS +0.27 against +0.39 (z = -2.69),
AUTOPHAGY z = -2.06. So MYC raises mito-turnover transcripts largely **because it raises
mitochondrial content ~20% wholesale** (@sec-content), not because it runs a mitophagy
programme. The direction of the bias is what makes this safe to state: the matched null is
anti-conservative, so a *non*-specific verdict from it is conservative.

**Where that leaves the death arm.** Lesner read their discordance as protective **mitophagy**
(MYC -> NRF1 -> DRP1/FUNDC1) rather than as priming. Our data neither support that route
specifically nor exclude it: the machinery rises, but no more than mitochondrial content does,
and the transcript layer cannot distinguish a mitophagy flux from a larger compartment in any
case. It is better treated as a second thing worth measuring than as a rival to the substrate
reading of @sec-window. What we take from them is the endorsement of the same three
experiments: **mtDNA qPCR** for copy number, **BH3 profiling** for the priming state, and
NRF1/DRP1/FUNDC1 at protein level.

::: {.callout-warning}
Batch = timepoint, so the Myc+ and wild-type timecourses quoted above are batch-confounded and
descriptive; the **interaction** is the clean quantity, and it is the one every claim in this
section rests on. Set-level fGSEA and the expression-matched null are the claims; single-gene
log-fold-changes are illustration. The comparison is to Lesner et al.'s **reported** results in
a preprint, not a reanalysis of their data.
:::

Sources on disk: `results/lesner_mechanism_test.rds` (script 39), with panels
`outputs/lesner/A_mechanism_genes.pdf`, `B_fgsea_contrasts.pdf` and
`C_mitophagy_specificity_null.pdf` -- the analysis layer's own, not promoted to the paper
figure set.

# Leads for the bench {#sec-leads}

Hypothesis-generating, not results (n=24): `redox -> MB2 tumorigenic fork` (the one
robust Myc-independent coupling; test: redox/GSH assay) and `nuclear OXPHOS -> death-priming`
(test: **BH3 profiling** -- priming is an occupancy state at the outer membrane, and no
amount of transcriptomics substitutes for measuring it).

**`Foxo3`, new 2026-07-27 and the one live mechanistic lead** -- it collapses like PUMA
(0.87th percentile of its matched stratum; +0.243 at 6W -> -0.242 at 12W, interaction
p 0.0074, against `Bbc3`'s +0.258 -> -0.283, p 0.0081) and it is PUMA's canonical
p53-independent activator. Test: **FOXO3 western across the two ages, and in the PGC1a
arm.** Caveats that must travel with it: both 6W effects are non-significant, both
interactions are genome-wide null, two genes are not a module, and `Foxo3` *rises* in the
wild-type gland. Note also that our direction is **opposite** to the canonical
PGC-1a-sequesters-FOXO3a axis in `docs/library_reference/PUMA-and-its-relationships.md`. Not
for the Results on the RNA alone.

**Closed, do not propose it again:** the retrograde/ISR route to PUMA is tested and negative
(script 44 PART D -- the ATF4 target set and Hallmark UPR are null in the collapse ranking,
and `Chac1`, which looked collapsed, has no 6W effect to lose at p 0.47).

The OXPHOS->priming lead is **gene-level, not composite-level** (corrected 2026-07-27, it was
described here in its pre-script-43 composite form). The composite version is dead twice
over: script 34 showed it is circular, and script 43 PART B showed that of five outcomes only
the **gene-level `Bbc3:Bcl2l1` ratio** has a `redox` control that behaves (`myc x OXPHOS-priority`
+6.09, p 0.005, 91.7th percentile against a control at the 51.2nd). The claim is a **sign
contrast on one outcome**, not a magnitude comparison across five. Two further tests the
mechanism menu (section 0.9 of the alignment doc) names, neither answerable from RNA:
cardiolipin by lipidomics and cristae architecture by EM.

# Provenance {#sec-provenance}

```{r}
#| label: session
#| echo: false
sessionInfo()
```
