---
title: "Mitochondrial content sets an apoptotic threshold for breast cancer transformation"
subtitle: "The bioinformatics behind Results 1-3, in full, before the reduction"
author: "Szabadkai lab"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: 2
    number-sections: true
    code-fold: true
    embed-resources: true
    theme: cosmo
    fig-format: png
    fig-dpi: 300
---

```{r}
#| label: setup
#| include: false

# Pin here() to the repo root. This .qmd lives under paper/ and a knitr render
# starts the session there; paper/_quarto.yml sets execute-dir: project, but the
# anchor is declared explicitly so a chunk run interactively behaves the same.
library(here)
here::i_am("paper/analysis_record.qmd")

# The panel spine. Sourced here so this document can call legend_md() -- the
# SAME renderer rebuild_panels.R uses to build legends.md -- rather than keeping
# a second copy of the legend format.
source(here::here("figures", "panels", "_panel_common.R"))

PANEL_DIR <- here::here("figures", "panels")

# --- the panel-embedding mechanism -------------------------------------------
# THIS DOCUMENT RE-RUNS NO ANALYSIS. It sources figures/panels/*.R exactly as
# rebuild_panels.R does -- one fresh environment per script, parented on the
# global environment, so a panel object from one script cannot quietly satisfy
# the next.
#
# myc.fig.capture = TRUE makes save_panel_p() record the panel's DESIGNED size
# into the .myc_panel_sizes registry and write nothing (_panel_common.R:495-503).
# `p` and `LEGEND` are still built. So one source() pass gives the figure, its
# legend AND its size, and THE SIZE COMES FROM THE PANEL'S OWN save_panel_p()
# CALL rather than from a number copied into this file that could drift.
#
# The cache matters for more than speed: knitr evaluates a chunk's !expr options
# BEFORE the chunk body, so panel_w() triggers the source and the body then reads
# the same environment instead of running the script twice.
.panel_cache <- new.env(parent = emptyenv())

panel_env <- function(slug) {
  if (!is.null(.panel_cache[[slug]])) return(.panel_cache[[slug]])
  f <- file.path(PANEL_DIR, paste0(slug, ".R"))
  if (!file.exists(f)) stop("no panel script: ", slug, ".R", call. = FALSE)

  old <- options(myc.fig.capture = TRUE)
  on.exit(options(old), add = TRUE)          # inside a FUNCTION this is correct;
                                             # at the top level of a sourced file
                                             # it fires at once (panels_to_pdf.R)
  env <- new.env(parent = globalenv())
  source(f, local = env)

  if (!exists("p", envir = env, inherits = FALSE))
    stop(slug, " built no `p`", call. = FALSE)
  if (!exists("LEGEND", envir = env, inherits = FALSE))
    stop(slug, " carries no panel_legend() block", call. = FALSE)

  reg <- get0(".myc_panel_sizes", envir = globalenv())
  if (is.null(reg) || !exists(slug, envir = reg, inherits = FALSE))
    stop(slug, " did not register a size -- does it call save_panel_p(p, \"",
         slug, "\")?", call. = FALSE)
  sz <- get(slug, envir = reg)
  if (!identical(sz$units, "mm")) stop(slug, " reports size in ", sz$units)

  assign("SIZE_MM", c(width = sz$width, height = sz$height), envir = env)
  .panel_cache[[slug]] <- env
  env
}

# Designed size in inches, for the fig-width / fig-height chunk options.
panel_w <- function(slug) unname(panel_env(slug)$SIZE_MM[["width"]]  / 25.4)
panel_h <- function(slug) unname(panel_env(slug)$SIZE_MM[["height"]] / 25.4)

# The panel object itself.
panel_p <- function(slug) panel_env(slug)$p

# The legend block, rendered by the project's own legend_md() and wrapped in a
# collapsible callout: the figure reads clean, the detail is one click away.
# legend_md()'s first line is a `### <slot>  (`slug`)` heading; inside a callout
# the heading IS the title, so it is demoted rather than duplicated.
panel_legend_md <- function(slug) {
  md <- legend_md(panel_env(slug)$LEGEND, slug = slug)
  c("::::: {.callout-note collapse=\"true\"}",
    paste0("## Legend block — ", sub("^### ", "", md[1])),
    md[-1],
    ":::::")
}

# The slot a panel declares, for the pairing assertions and the manifests.
panel_slot <- function(slug) panel_env(slug)$LEGEND$slot

# Every panel script on disk, in name order -- the same glob rebuild_panels.R
# uses, so this document cannot silently omit a panel that exists.
ALL_SLUGS <- sub("\\.R$", "",
                 sort(list.files(PANEL_DIR, pattern = "^fig.*\\.R$")))

# Which slugs each chapter draws. Declared here, checked at the foot of the
# document: all 25 exactly once, and every "(alt)" beside its own original.
DRAWN <- list()
draw_here <- function(...) {
  s <- c(...)
  DRAWN[[length(DRAWN) + 1L]] <<- s
  invisible(s)
}
```

::: {.callout-important}
## What this document is

The bioinformatics of Results sections 1--3, **in full**, recorded on 2026-08-14 because the
manuscript text and figures are about to be cut to roughly half.

Most of what will be removed is not wrong. It is the evidence, the bounds and the negative
controls that *license* the sentences that stay. Removing it from the paper is right;
removing it from the project is not. This document is where it goes.

**It re-runs no analysis.** Every figure below is produced by sourcing the panel script that
the manuscript figure uses, and every number in every legend is computed by that script from
`results/*.rds`. Nothing here is transcribed by hand, so this document cannot drift away from
the panels it describes.
:::

# How to read this {#sec-how}

## The layers

```
scripts/00-44          the numbered pipeline. Hand-run in Positron, never by a figure.
   |                   DESeq2 (~ timepoint * myc_status), fGSEA, mitoPPS, GSVA,
   |                   and the exploratory arc 32-44.
   v
results/*.rds          what those scripts saved. Gitignored, generated at runtime.
   |
   v
figures/panels/*.R     one script per manuscript panel. Reads results/, computes its
   |                   own statistics, ASSERTS them against the analysis of record,
   |                   draws, and carries a panel_legend() block.
   v
this document          the panels in the order the Results cites them, each with its
                       legend block, plus what the analysis rejected and what is open.
```

`figures/panels/PANELS.md` is the working manifest: the slot map, a signed-off entry per
panel recording *why* it has the form it has, and the running list of sentences the data does
not support. This document is the readable face of it.

## Conventions

- **The written text is quoted verbatim**, in blockquotes, from the Google Doc
  `11.8.26_MK_Myc_Paper`, tab **CURRENT VERSION**, as read on **2026-08-14**. Paperpile
  citation links are reduced to author--year; nothing else is altered, including the figure
  callouts, because that mapping is part of what is being recorded.
- **Panels are drawn at their designed physical size.** A panel is 89 mm wide because that is
  one manuscript column, and the type is set for that size. Magnify with the browser's zoom.
- **Legend blocks are collapsed by default.** They are long on purpose: `what` is the
  sentence, `detail` carries n, units, the test and the thresholds, `bounds` carries the
  caveat that must travel with the claim, and `source` names the object every number came
  from.

# The measurement, and four standing constraints {#sec-constraints}

Bulk RNA-seq of **purified mammary epithelial cells** from MMTV-Myc transgenic mice and their
wild-type littermates at **6 and 12 weeks** --- four groups (`6W_wt`, `12W_wt`, `6W_myc`,
`12W_myc`), **six animals each, n = 24** --- analysed with DESeq2 on a
`~ timepoint * myc_status` interaction design.

Four facts about that design govern every reading below. Almost every bound in every legend
block is one of them.

### 1. Purification is by enzymatic dissociation, with no sorting step

So any stromal, endothelial or immune signal is **contamination, not tissue composition**
(3--12 % residual, which is what no sort predicts) and is modelled as a nuisance, never read
as biology. And the warm digest is itself a stressor: the immediate-early signature it
induces (van den Brink 2017) is a real axis in these samples, which is why prep-stress
appears as a covariate wherever a per-sample quantity is interpreted.

### 2. Batch = timepoint

The 6W and 12W cohorts were extracted as **two separate batches**, so batch is perfectly
confounded with age. Genotype is **balanced within each batch**, so the interaction design
absorbs the batch effect into the timepoint main effect.

**Consequences, and they are not symmetric.** Every genotype gap and every interaction is
clean --- and so is any *difference between two temporal contrasts*, because the shared
offset cancels. What is **not** separable is the pure between-timepoint reading on its own:
the developmental claim is confounded with batch, and no QC metadata exists on disk
(no dissociation batch, no viability, no RIN) with which to separate them. Where a temporal
contrast is drawn it is **described**, not claimed; what is claimed is the difference between
two of them.

### 3. n = 6 per cell

An interaction is the least powered thing this design can be asked for. The median `lfcSE` on
an interaction term is **0.333** against **0.233** for a genotype contrast. Nothing in the
death arm clears genome-wide BH correction, and the licence for testing `Bbc3` and `Foxo3` at
all is that they were **pre-specified from the cell experiments**, not that they cleared a
threshold.

### 4. There is one dominant per-sample axis, and it is a correlation ceiling

Score the 885 gene sets per animal and **PC1 is 76 % of the variance**, with effective
dimensionality 1.7 of a possible 23. `cor(mito_oxphos, mean of all 884 other scores) = 0.968`.
So an arbitrary pair of composites correlates at ~0.8 **by construction**, and a published
coupling of 0.8 is what the window hands you rather than a finding (script 36).

The axis is **genotype-independent (p = 0.49)** and **time-associated (p = 0.0003)**. That is
why the genotype and interaction results survive it and several per-sample coupling results
did not --- @sec-died.

::: {.callout-warning}
## The one inference rule that follows

At n = 24, with a 0.8 ambient correlation, **these analyses rank; they do not confirm.** A
residual correlation of ~0.29 arises under the null. Where this document quotes a coupling it
is a lead with an experiment attached (@sec-open), not a result. The claims that *are*
confirmatory are the contrasts: raw count shares, DESeq2 log2 fold changes, and Wald-ranked
fGSEA. Those never touch the couplings.
:::

# Section 1 --- MYC drives early mammary tumourigenesis {#sec-s1}

> ## MYC drives early mammary tumourigenesis by inducing OXPHOS and biosynthetic pathways
>
> To understand the MYC-driven changes in the transcriptome of mammary epithelial cells
> (MECs) during breast cancer initiation, we isolated MECs from mammary fat pads of the
> MMTV-Myc genetically engineered mouse model at two early timepoints of tumourigenesis, at 6
> and 12 weeks of age. Mammary fat pads of the MMTV-Myc+/+ model, known to recapitulate the
> heterogeneity and progression of the human disease (Grimm and Nordeen 1998; Stewart et al.
> 1984; Jamerson et al. 2004; Andrechek et al. 2009; Broeker et al. 2023), showed hyperplastic
> expansion of the developing ductal epithelial structures between these time points
> (**Fig. 1A**). The wild-type (MMTV Myc-/-) mammary ductal system undergoes significant
> developmental changes from 6 to 12 weeks. Estrogen-driven growth factor signaling promotes
> terminal end bud (TEB) cell proliferation during puberty at 6 weeks. By 12 weeks, the mature,
> adult ductal structure exhibits decreased estrogen receptor levels and reduced cell
> proliferation. To investigate how MYC influences the developmental trajectory, we compared
> the transcriptome of the purified MEC population using both longitudinal (6W versus 12W for
> WT or Myc+ genotypes) and cross-sectional (WT versus Myc+ at 6W or 12W) comparisons
> (**Fig. S1A**). This analysis used fGSEA ranking and GSVA scoring based on custom-curated
> genesets (**Fig. S1B**), derived from previous single-cell RNA-seq datasets (see detailed
> list in Methods). Consistent with earlier results, the overall distribution of primary
> mammary cell states remained predominantly unaltered, including the basal-myoepithelial
> (BMYO), luminal hormone-sensitive (LHS), and luminal alveolar secretory precursor (LASP)
> states (Gray et al. 2025). Nevertheless, a distinct transition from TEB to ductal states was
> observed at 12W, representing a clear divergence from the pubertal proliferative phenotype
> (**Fig. 1B**) (Gray et al. 2023). Importantly, an endogenous MYC program contributed to the
> pubertal TEB state in WT animals, promoting proliferation, evidenced by the reduction of the
> canonical MYC, TEB and proliferation signatures in the adult stage (WT 6W → 12W). Transgenic
> MYC activation amplified the TEB and proliferation state, suppressed the BMYO lineage, and
> strongly promoted a lineage suppressed (LE) de-differentiation pattern of the luminal states
> (Gray et al. 2023). This effect diminished in both the WT and Myc+ 6W → 12W trajectory,
> suggesting a suppression of the endogenous MYC program.

## The design, and the library {#sec-s1-design}

**Fig. 1A is bench material** --- whole-mount and histology of the hyperplastic ductal
expansion, assembled outside R. It now carries a second quantity, the Ki-67 : CC3 ratio,
which is the phenotype the whole of @sec-s3 exists to explain.

```{r}
#| label: fig-s1a
#| echo: false
#| fig-width: !expr panel_w("figS1_design_contrasts")
#| fig-height: !expr panel_h("figS1_design_contrasts")
invisible(draw_here("figS1_design_contrasts"))
print(panel_p("figS1_design_contrasts"))
```

```{r}
#| label: leg-s1a
#| echo: false
#| output: asis
cat(panel_legend_md("figS1_design_contrasts"), sep = "\n")
```

Four groups, six animals each. The schematic exists because **which contrast a number comes
from decides what it can mean** here more than in most designs --- see the batch constraint in
@sec-constraints.

```{r}
#| label: fig-s1b
#| echo: false
#| fig-width: !expr panel_w("figS1_geneset_library")
#| fig-height: !expr panel_h("figS1_geneset_library")
invisible(draw_here("figS1_geneset_library"))
print(panel_p("figS1_geneset_library"))
```

```{r}
#| label: leg-s1b
#| echo: false
#| output: asis
cat(panel_legend_md("figS1_geneset_library"), sep = "\n")
```

::: {.callout-warning}
## Three set counts are in play, and each panel uses a different one {#txt-counts}

**986** sets in the library (this panel, and the number the text gives) --- **885** scored per
sample by GSVA (Figs. 1C and S1C) --- **866** in the enrichment ranking (Fig. 1D: 816
fGSEA-eligible library sets clearing the minSize 10 / maxSize 500 filter, plus 50 fresh
Hallmark comparators). 986 is correct where it stands; the other two need naming where they
are cited.
:::

The fill is the **mitochondrial definition** of each set in three classes, and it carries a
number that matters for how @sec-s1-axis is read: the library is **40 % mitochondrial by
label but 13 % once the construction lanes are set aside** --- 260 of the 393 mito-labelled
sets are Gray `_MITO` / `_LE_MITO` transcription-factor lanes, which are MitoCarta subsets *by
build*.

## The developmental trajectory, and what the transgene does to it {#sec-s1-teb}

```{r}
#| label: fig-1b
#| echo: false
#| fig-width: !expr panel_w("fig1_myc_teb_proliferation")
#| fig-height: !expr panel_h("fig1_myc_teb_proliferation")
invisible(draw_here("fig1_myc_teb_proliferation"))
print(panel_p("fig1_myc_teb_proliferation"))
```

```{r}
#| label: leg-1b
#| echo: false
#| output: asis
cat(panel_legend_md("fig1_myc_teb_proliferation"), sep = "\n")
```

::: {.callout-warning}
## Four sentences in this paragraph need a word {#txt-teb}

- **"no major changes in the overall composition"** --- in within-group SD the wild-type
  6→12W shift is BMYO **+0.10**, LASP **−0.70**, LHS **−0.60**, TEB-ductal **−0.91**. BMYO is
  flat; the two luminal states fall by two thirds of what the TEB axis does. All four are
  non-significant at n = 6 (p 0.24--0.86), so the sentence survives --- but script 26's own
  header says the change *is* luminal decline, not a BMYO expansion.
- **"a clear TEB-ductal shift"** --- not supported by the sample-level test in wild type
  (**p = 0.24**). What licenses "clear" is the matched-null ruler: TEB-vs-ductal (HS)
  **−0.422, 0th percentile**. In Myc+ the shift *is* significant (**p = 0.001**).
- **"suppressed the BMYO lineage in favor of differentiation to LHS"** --- the BMYO half
  **holds** and deepens with age (−0.80 SD at 6W, −1.16 SD at 12W, p = 0.066). The LHS half is
  a **reversal, not a main effect**: −0.63 SD at 6W, **+0.79 SD at 12W** (pooled p = 0.86,
  interaction p = 0.097). True at 12 weeks, false at 6.
- **"the transgene amplified the TEB effects"** --- resolved in the author's favour, but only
  once split by age: pooled it is +0.63 SD, p = 0.14, and reads as a null; split it is
  **+1.02 SD at 6 weeks** and **+0.23 SD at 12**. The effect exists where the phenotype is.

**The de-differentiation reading is the stronger version of the sentence.** MYC raises every
lineage-suppressed (LE) composite and lowers every differentiated (HE) composite, in all three
lineages, at both ages (AP +1.51 / −0.82, BA +1.55 / −1.00, HS +1.31 / −0.85 SD at 6 weeks).
**The HE numbers must appear in the text even though only the LE arm is drawn**, because the
LE composites sit almost on the global common-mode axis (r 0.85--0.90) and the HE ones do not
--- what a common-mode axis cannot produce is *opposite signs within a lineage*, so the
opposition is what licenses the reading.
:::

::: {.callout-note collapse="true"}
## Built, currently cited nowhere --- the MB-fork specificity control

`figS1_mb_fork_specificity.R` held slot S1C, then S1D, and is now uncited: the written S1D is
the MYC western blot. It is the specificity control that Fig. 1B's `Human BRCA-MYC` row needs
--- MB1 and MB2 share a lower fork and diverge into two upper forks that are *both*
biogenesis-high and proliferative, so the only feature distinguishing them is MYC. It keeps
building and gets a letter back the day a sentence asks for one.

```{r}
#| label: fig-mbfork
#| echo: false
#| fig-width: !expr panel_w("figS1_mb_fork_specificity")
#| fig-height: !expr panel_h("figS1_mb_fork_specificity")
invisible(draw_here("figS1_mb_fork_specificity"))
print(panel_p("figS1_mb_fork_specificity"))
```

```{r}
#| label: leg-mbfork
#| echo: false
#| output: asis
cat(panel_legend_md("figS1_mb_fork_specificity"), sep = "\n")
```
:::

## The dominant axis {#sec-s1-axis}

```{r}
#| label: fig-1c
#| echo: false
#| fig-width: !expr panel_w("fig1_pathway_pca")
#| fig-height: !expr panel_h("fig1_pathway_pca")
invisible(draw_here("fig1_pathway_pca"))
print(panel_p("fig1_pathway_pca"))
```

```{r}
#| label: leg-1c
#| echo: false
#| output: asis
cat(panel_legend_md("fig1_pathway_pca"), sep = "\n")
```

**The one thing that makes this panel readable: PC1 is a genotype axis, not a time axis.**
`PC1 ~ timepoint * genotype` gives genotype **p = 0.0031**, timepoint p = 0.70, interaction
p = 0.43. Since batch = timepoint, an age-separating axis would have been uninterpretable.

```{r}
#| label: fig-s1c
#| echo: false
#| fig-width: !expr panel_w("figS1_axis_loadings")
#| fig-height: !expr panel_h("figS1_axis_loadings")
invisible(draw_here("figS1_axis_loadings"))
print(panel_p("figS1_axis_loadings"))
```

```{r}
#| label: leg-s1c
#| echo: false
#| output: asis
cat(panel_legend_md("figS1_axis_loadings"), sep = "\n")
```

::: {.callout-warning}
## "aligned almost entirely with mitochondria related terms" is bounded three ways {#txt-axis}

Supported at the top of the ranking: mito sets are 44 % of the library and **100 % of the top
50**, median \|loading\| 0.90 against 0.61. But:

1. **94 of the top 100 are build-tautological** `_MITO` lanes --- MitoCarta subsets by
   construction, so a high loading is a property of how the set was made.
2. **Non-mitochondrial growth programmes reach 0.98** --- the METABRIC biclusters, nucleotide
   metabolism, pentose phosphate, the MYC and E2F regulons.
3. **Redox, a mitochondrial set, loads 0.67**, in the bottom half --- and it is the one axis
   MYC does not drive.

Suggested wording: **mitochondria-led**, not "almost entirely mitochondrial". What the axis
tracks is a coordinated MYC anabolic--proliferative--mitochondrial growth state.

**And "revealing the leading role of mitochondrial remodeling" does not follow.** Loading is
**covariation, not primacy**, and script 37 says so in its own saved verdict
(`pathway_loading.rds$enrichment_verdict`). At n = 24 nothing on these panels can order cause;
the primacy claim belongs to the perturbations.
:::

```{r}
#| label: fig-1d
#| echo: false
#| fig-width: !expr panel_w("fig1_nes_ranking")
#| fig-height: !expr panel_h("fig1_nes_ranking")
invisible(draw_here("fig1_nes_ranking"))
print(panel_p("fig1_nes_ranking"))
```

```{r}
#| label: leg-1d
#| echo: false
#| output: asis
cat(panel_legend_md("fig1_nes_ranking"), sep = "\n")
```

::: {.callout-warning}
## The tier order is right at the top and inverted in the middle {#txt-tiers}

OXPHOS **+2.42**, MYC signatures **+2.38** and mitochondrial biogenesis **+2.37** are a
three-way tie at the top, exactly as written. But the sentence continues "followed by
biosynthetic metabolic pathways, E2F and overall proliferation signaling", and in the data
**E2F / cell cycle (+2.08) sits above biosynthetic metabolism (+1.84)**. One-word fix: swap
them.

Two more things the panel shows and the text does not use. The **curated-mitochondrial row
tops the panel at +2.47**, above OXPHOS --- it is drawn as its own row in its own colour
rather than deleted, because deleting it would flatter every row beneath it, and the rows to
read as programmes start below it. And **apoptosis is the one row that changes sign** (+0.74
at 6 weeks, −0.94 at 12, while every other row's median moves *up*). That is **not a result**:
the row is bimodal with its median on zero, so five weak, non-significant sets crossing zero
move it, and three of the five are non-apoptotic death modalities. The members that *are*
significant do not move.
:::

## Two mechanisms: content, and reprioritisation {#sec-s1-two}

> MYC-driven mitochondrial alterations manifested through two mechanisms. First, MYC drove a
> quantitative expansion by systemically upregulating 94% of 143 nuclear-encoded mitochondrial
> pathways, increasing the mitochondrial transcriptomic fraction by 21--27% (padj \<0.001),
> demonstrating robust mitochondrial biogenesis (**Fig. 1E**). Second, MYC reorganized how
> mitochondria prioritize its different pathways, measured using MitoPPS analysis (Monzel et
> al. 2025) (**Fig. 1F**). At 6 weeks MYC significantly promoted protein import/homeostasis,
> translation, OXPHOS subunits (all complexes), and biosynthetic pathways including glycine
> cleavage, pyruvate, and serine metabolism. Conversely, it demoted dynamics and surveillance,
> including fission, mitophagy, and apoptosis alongside calcium signalling.

**Two rulers, and the pair is the claim.** Content is the set-average raw log2 fold change ---
what the compartment *has*. mitoPPS (Monzel et al. 2025) is a pairwise ratio within the
mitochondrial compartment, **blind to content** --- what the compartment *spends its budget
on*. On content **95.1 %** of the 143 MitoPathways sit above zero; on mitoPPS **56.6 %**. The
difference between those two numbers **is** "two mechanisms".

```{r}
#| label: fig-1e
#| echo: false
#| fig-width: !expr panel_w("fig1_mito_content")
#| fig-height: !expr panel_h("fig1_mito_content")
invisible(draw_here("fig1_mito_content"))
print(panel_p("fig1_mito_content"))
```

```{r}
#| label: leg-1e
#| echo: false
#| output: asis
cat(panel_legend_md("fig1_mito_content"), sep = "\n")
```

::: {.callout-important}
## This is the one solid new result in the mouse dataset

Script 32 was written because **"biogenesis" had no content referent**: AP-abundance is
dispersion, mitoPPS is content-blind by construction, and the total mitochondrial score had
never been tested. It is now: **MYC raises mitochondrial content by 21--27 %, p = 0.0006
adjusted** for prep-stress and contamination, and it is a **lower bound**, because MYC also
amplifies total RNA per cell so a *share* understates it.
:::

::: {.callout-warning}
## Three numbers in this sentence need fixing {#txt-content}

- **"94 % of 143" mixes two numbers.** 94.4 % is the fraction above zero over **all 144** rows
  of script 40's ruler (the synthetic mtDNA-encoded pathway included); **95.1 %** is the
  fraction over the **143** that every regression, null and panel actually uses. Pick one and
  say which.
- **"21--27 %" is narrower than what the panel draws.** Those are the median per-gene rise
  across the nuclear compartment (**+21.4 %**) and the mass markers (**+26.5 %**). The five
  *set* arms drawn run **+20 % to +33 %**. Either widen the sentence or name the two
  quantities the range comes from.
- **"(padj \< 0.001)" is `BIOGENESIS_FULL`, p = 0.00049, and it is a POOLED genotype main
  effect**, not a per-age one. Split by age at n = 6 per cell, four nuclear arms clear p \<
  0.05 at **twelve** weeks and only biogenesis clears it at **six**. The effect *size* is the
  same at both ages (every interaction p 0.35--0.81), so the pooled test is the right one ---
  but the text should not imply a six-week significance the six-week data alone do not carry.
:::

```{r}
#| label: fig-1f
#| echo: false
#| fig-width: !expr panel_w("fig1_reallocation_ranked")
#| fig-height: !expr panel_h("fig1_reallocation_ranked")
invisible(draw_here("fig1_reallocation_ranked"))
print(panel_p("fig1_reallocation_ranked"))
```

```{r}
#| label: leg-1f
#| echo: false
#| output: asis
cat(panel_legend_md("fig1_reallocation_ranked"), sep = "\n")
```

::: {.callout-warning}
## "MYC significantly promoted ... OXPHOS subunits (all complexes)" --- the word does not reach that far {#txt-oxphos-sig}

Checked against script 08's own BH-adjusted p across all 144 pathways at 6W, of which 33 clear
0.05:

| in the sentence | mitoPPS at 6W | padj | verdict |
|---|---|---|---|
| translation | +0.117 | **0.042** | supported |
| pyruvate | +0.328 | **0.042** | supported |
| serine | +0.212 | **0.016** | supported |
| protein import / homeostasis | +0.117 | 0.053 | marginal at tier level; SAM 0.016, chaperones 0.039 clear it |
| glycine cleavage | +0.405 | 0.118 | **not significant** --- and it is the *largest* promotion on the list |
| **OXPHOS subunits** | **+0.112** | **0.274** | **not significant** |
| — CI / CII / CIII / CIV / CV | +0.055 / +0.021 / +0.160 / +0.017 / +0.165 | 0.37 / 0.54 / 0.16 / 0.86 / 0.15 | **not one complex is significant** |
| dynamics & surveillance | −0.156 | **0.028** | supported |
| fission | −0.202 | **0.016** | supported |
| apoptosis | −0.247 | **0.043** | supported |
| mitophagy | −0.182 | 0.084 | **not significant** |
| calcium signalling | −0.075 | **0.026** | supported |

No part of the OXPHOS arm is significant on this ruler, and **the one OXPHOS row that is, is
`CIII assembly factors` at −0.141 (padj 0.022) --- demoted.**

**Suggested repair, and it is a better sentence:** move OXPHOS out of the "significantly
promoted" clause into its own --- *"the OXPHOS subunits moved in the same direction (+0.112)
while their assembly factors did not (−0.028)"* --- because that split is exactly what
returns, much larger, in the wild-type timeline of @sec-s3.

Also worth knowing: glycine cleavage, pyruvate and serine are the *largest single* promotions
(+0.41 / +0.33 / +0.21), larger than OXPHOS, which leads as a **tier** and not as a leaf. Do
not let a later draft turn this into "OXPHOS leads".
:::

::: {.callout-note}
## MitoCarta sets are membership-loose, and the panel prints n for that reason

Several of the largest effects rest on four genes (glycine cleavage 4, serine 4). A big effect
on a four-gene set is **one gene, not a module** --- `Mrpl12` at +116 % carries
`MITOCARTA_TRANSCRIPTION`; `Atad3a` at +113 % carries `NUCLEOID`. No set-level mechanism claim
should be made without resolving it gene by gene.
:::

# Section 2 --- the paradoxical de-dosing {#sec-s2}

> ## A paradoxical reduction of MYC signalling is associated with the initial tumour cell expansion in the adult mammary gland
>
> While the MYC program remained unchanged and its relative ranking was preserved in the whole
> transcriptome (***see Figs 1D, E***), the amplitude of the MYC induced changes significantly
> declined during the 6W→12W puberty-adult transition. At 12 weeks the entire structure was
> present at half amplitude and unchanged in shape both in the whole and mitochondrial
> transcriptome (R2=0.51, 0.8, p\< 0.002, respectively; **Fig. 1G**) and a similar effect on
> the mitopathway priority scores was observed (***see Fig. 1F***). Interestingly, this ranking
> was not just maintained but enhanced at 12W in comparison with 6W: the normalized enrichment
> score for the mitochondrial OXPHOS set and MYC integrative signatures increased substantially
> (**Fig. 1H**). Thus, the adult MYC program represented the pubertal program operating at a
> lower intensity, but higher priority across the whole transcriptome.

## Rescaled, not reshaped {#sec-s2-rescale}

```{r}
#| label: fig-1g
#| echo: false
#| fig-width: !expr panel_w("fig1_rescaled_not_reshaped")
#| fig-height: !expr panel_h("fig1_rescaled_not_reshaped")
invisible(draw_here("fig1_rescaled_not_reshaped"))
print(panel_p("fig1_rescaled_not_reshaped"))
```

```{r}
#| label: leg-1g
#| echo: false
#| output: asis
cat(panel_legend_md("fig1_rescaled_not_reshaped"), sep = "\n")
```

::: {.callout-important}
## The distinction the sentence turns on: "half amplitude" is the SLOPE, "unchanged in shape" is the R². Only one of them beats its null.

| | observed | null median | null max | draws ≥ observed | p |
|---|---|---|---|---|---|
| slope | 0.552 | 0.419 | --- | **64 of 500** | 0.13 |
| **R²** | **0.801** | 0.298 | **0.719** | **0 of 500** | **\< 0.002** |

A shuffled set of the same size and expression **can** halve an effect. What it cannot do is
halve it *coherently*. **So the p belongs to the R², and the sentence should not attach a p to
the slope.**
:::

The `R2=***` placeholders the draft carried are resolved, and two of them must not be crossed:

| in the sentence | number | what it is |
|---|---|---|
| whole transcriptome, slope | **0.487** | through the origin, over the 2,648 genes MYC moves at 6W --- script 44's `global_rate_fitted` |
| whole transcriptome, "overall R2" | **0.51** | over the same 2,648. Over **all 8,774** reported genes the slope barely moves (0.450) but R² falls to **0.381** --- regression dilution from six thousand genes MYC does not move |
| mitochondrial, slope | **0.552** | intercept 0.0001, so a pure rescaling and not a shift; bootstrap 0.514--0.605 |
| mitochondrial, "overall R2" | **0.801** | |
| **"OXPHOS R2"** | **0.761** | the 19 OXPHOS pathways, **content** ruler, slope 0.525. **Not 0.945** --- that is the *mitoPPS* ruler, and mixing rulers inside one parenthesis would be wrong. The mitoPPS figures (slope 0.644, R² 0.789; OXPHOS 0.945) belong with the "see Fig. 1F" clause |
| "p \<" | **0.002** | empirical, 0 of 500 expression-matched shuffles reach R² 0.801 |

::: {.callout-warning}
## "Whole transcriptome" needs one word {#txt-whole}

The panel's left facet is labelled *MYC-responsive genes (2,648)*, because that is the set the
rate of record is fitted on. If the sentence keeps "whole transcriptome" it should quote
**0.450 / 0.381**; if it quotes **0.487 / 0.51** it should say "the genes MYC moves". Either
is fine. They must not be crossed.
:::

```{r}
#| label: fig-1h
#| echo: false
#| fig-width: !expr panel_w("fig1_nes_preserved_sharpened")
#| fig-height: !expr panel_h("fig1_nes_preserved_sharpened")
invisible(draw_here("fig1_nes_preserved_sharpened"))
print(panel_p("fig1_nes_preserved_sharpened"))
```

```{r}
#| label: leg-1h
#| echo: false
#| output: asis
cat(panel_legend_md("fig1_nes_preserved_sharpened"), sep = "\n")
```

::: {.callout-warning}
## "Not just maintained but enhanced" does not survive, for two independent reasons {#txt-enhanced}

**"Maintained" is strongly true and is the panel's real result** --- Spearman **0.933** over
866 sets while the effect size halves. Better still: the 92 sets that cross sign between the
ages are **84 % mammary-development and TF-target sets**; set them aside and Spearman over the
remaining 774 rises to **0.967**. *What holds its order is the metabolic, mitochondrial and
MYC core; what reshuffles is lineage identity.*

**"Enhanced" does not.**

1. **The rise is not specific.** Every point is above the line, not just the named ones ---
   **81 % of all 866 sets rise** (median +0.312), and **93 % of the 403 already enriched at
   six weeks**. As families the two named rise by the *median* amount: OXPHOS **+0.308, the
   49th percentile**; MYC signatures **+0.386, the 60th**. The individual sets quoted in
   earlier drafts are the best members of their families, not typical of them.
2. **A global NES rise is what a weaker ranking mechanically produces.** fGSEA normalises each
   enrichment score against a permutation null built from *that same ranked list*. The
   twelve-week Wald list is much flatter --- **SD 1.221 vs 1.744, 9.5 % of genes past
   \|stat\| = 2 vs 22.5 %** --- so random sets reach smaller enrichment scores, the normaliser
   shrinks, and the same relative enrichment scores a higher NES. **No claim of increased MYC
   activity can rest on this panel.**

**What is true, and is a better sentence:** NES is enrichment *relative to the rest of the
transcriptome*, so these programmes **hold their position at the top of the ranking while the
amplitude halves**. That is a statement about **position, not magnitude** --- which is exactly
what the author's own closing sentence ("lower intensity, but higher priority") already says.
It sits beside Fig. 1G rather than against it.
:::

## The mechanism: dose {#sec-s2-dose}

> Mechanistically, this attenuation was driven by a \~50% reduction in MYC protein levels at
> 12W compared to 6W (**Fig. S1D**). This decline occurred despite stable transcript levels;
> the *Myc* transcript level gap remained constant between 6W\_myc and 12W\_myc contrasts and
> the proximal MYC/MAX/MXD network (**Fig. S1E, F**). The MYC protein halving predicted the
> overall 0.55-fold transcriptomic rescaling (***see Fig. 1G***). However, MYC dosage reduction
> acts non-linearly due to differential promoter affinity: high-affinity targets are less
> sensitive to protein fluctuations than low-affinity sites (Lorenzin et al. 2016).
> Consistently, the core MYC signature retained 74--77% of its pubertal effect size, whereas
> lower-affinity mitochondrial biosynthetic and OXPHOS arms declined to 50--58%.

**Fig. S1D is the western blot --- bench material, and it is what closed this question.**
Before it, "why does the MYC effect halve" had four live hypotheses: (H1) the programme simply
decays; (H2) a background programme antagonises it; (H3) reduced *effective* MYC at constant
message; (H4) compositional dilution. Extensive negative scans are on the record --- **every
E-box binder is flat**, and a 111-gene nuclear-receptor / corepressor / DREAM scan gives
**zero significant interactions** --- so H2's selective form fails. The blot settles it on H3:
protein −50 %, and the transcriptomic rate of record is **×0.55**.

```{r}
#| label: fig-s1e
#| echo: false
#| fig-width: !expr panel_w("figS1_myc_transcript_stable")
#| fig-height: !expr panel_h("figS1_myc_transcript_stable")
invisible(draw_here("figS1_myc_transcript_stable"))
print(panel_p("figS1_myc_transcript_stable"))
```

```{r}
#| label: leg-s1e
#| echo: false
#| output: asis
cat(panel_legend_md("figS1_myc_transcript_stable"), sep = "\n")
```

**The control the blot needs.** If the message had fallen too, the attenuation would need no
further explanation and the dose argument would collapse into "the transgene was silenced". It
does not fall: the genotype gap is **+1.68 → +1.82 log2** and the Myc+ gland does not decline
across the window (**−0.10, padj 0.76**). Measured as retention, `Myc` is the **most retained
gene in the transcriptome** --- 1.085, the 99.99th percentile of 8,774.

::: {.callout-warning}
## "the expression gap remained constant between 6W_myc and 12W_myc" --- say which gap {#txt-gap}

A gap is between **genotypes**, not between ages. Both readings happen to be true and the
panel draws both (genotype gap +1.68 → +1.82; no decline within Myc+, −0.10 at padj 0.76), but
the sentence should say which it means.
:::

```{r}
#| label: fig-s1f
#| echo: false
#| fig-width: !expr panel_w("figS1_myc_network")
#| fig-height: !expr panel_h("figS1_myc_network")
invisible(draw_here("figS1_myc_network"))
print(panel_p("figS1_myc_network"))
```

```{r}
#| label: leg-s1f
#| echo: false
#| output: asis
cat(panel_legend_md("figS1_myc_network"), sep = "\n")
```

**A clean negative, and the most obvious alternative to dose.** MYC cannot bind an E-box
without MAX and competes for the same sites with the MXD/MNT family, so a fall in MAX or a
rise in the repressors would attenuate its output at constant MYC. Neither happens: nothing
clears significance at either age, and no interaction is below **padj 0.83**.

::: {.callout-note}
## The error-bar equality is worth more than the negative

The standard errors of the two ages agree to within **0.024 log2 across all ten genes**,
median ratio 1.01. That is the objection every attenuation result in this paper has to survive
--- *twelve weeks is not the noisier cohort* --- and it is asserted in the panel script rather
than assumed.
:::

## The paradox this sets up {#sec-s2-paradox}

> These results raised the question of how MYC-driven hyperplasia is established at 12W in the
> adult mammary gland when the oncogenic program exhibits attenuated amplitude and the driver's
> transcriptional output is halved? This paradoxical observation suggested that early
> tumourigenesis may not depend on the absolute intensity of the MYC program, but rather on the
> balance between different MYC-induced programmes, i.e. proliferation and cell death. Indeed,
> as shown in **Fig. 1A**, the ratio of the Ki-67 and CC3 positive cells increased significantly
> at 12W as compared to 6W, mainly due to a reduction of the number of cells undergoing
> apoptosis, suggesting that reduced apoptotic sensitivity with maintained proliferation
> signalling underlies early stages of MYC-driven tumourigenesis following the puberty-adult
> transition.

::: {.callout-note}
## Two things the section decided not to draw, and both were right

**The DE counts.** 1,967 → 135 genes at FDR 5 %, 2,777 → 239 at 10 % --- while the median
\|LFC\| of the survivors *rises*, 0.678 → 0.869. Drawn, that would flatter a −93 % that is
really a −45 %. It is threshold arithmetic, and belongs in a trailing italic sentence.
(For the record: the `interaction` contrast has **0** genes at either threshold.)

**The interaction standard-error point** (0.333 vs 0.233) --- same reason. It is the power
statement of @sec-constraints, and the place it earns its keep is beside the `Foxo3` / `Bbc3`
interaction in @sec-s3, not here.
:::

::: {.callout-important}
## Dose vs competence --- the spine that carries into Section 3

The MMTV-Myc timeline measures **MYC DOSE**: protein halves, output per unit MYC is unchanged,
and that single fact explains most of the negative results in this corpus. The MYC-ER model of
@sec-s3 measures something else --- **COMPETENCE**: 80 % less death at *equal* dose, in animals
with no prior MYC exposure, so the change is **developmental, not selection**.

The molecular hinge is that MYC builds a death-ready mitochondrion at **both** ages (`Htra2`
+1.49 padj 2.6e-8, `Bax` up, `Bcl-xL` down) and loses the **trigger**: `Bax`:`Bcl-xL` retains
**0.55**, the global rate, while `PUMA`:`Bcl-xL` retains **−0.09**. The non-linearity is
Lorenzin's affinity-ordered occupancy --- core 0.75 \> global 0.55 \> mito 0.5 \>\> PUMA
−0.09 --- which is why "the programme fades" does not, on its own, explain the death arm.
:::

# Section 3 --- OXPHOS repression, and the loss of the trigger {#sec-s3}

> ## OXPHOS repression in the developing mammary gland is coupled to tumourigenesis via the loss of the apoptotic trigger PUMA
>
> To evaluate the proliferative effects and apoptotic sensitivity of MYC at constant dosage of
> the transgene at different stages of mammary gland development, we crossed R26-LSL-CAG-MYCERT2
> mice with MMTV-Cre mice to generate a model with mammary-specific switchable MYC (MYCERT2)
> (**Fig. 2A**) (Kreuzaler et al. 2019; Bywater et al. 2020; Kreuzaler et al. 2023). We
> administered tamoxifen (tam) or vehicle for 48 hours to 6W and 12W old MYCERT2+/- mice and
> evaluated proliferation and apoptotic rates using IHC (**Fig. 2B, C**). Importantly, the
> apoptotic response robustly induced by MYC at 6W was almost completely lost in the 12W group,
> whereas MYC-mediated proliferation remained at approximately 70% in the adult mammary gland,
> resulting in a markedly elevated proliferation-to-apoptosis ratio. Crucially, these
> alterations were not a consequence of decreased MYC expression at 12W (**Fig. S2A**).
> Furthermore, on the longer term (13 days follow up), MYC triggered substantial MEC expansion
> exclusively when activated in the 12W adult mammary gland (**Fig. 2D**). E-cadherin staining
> revealed significant ductal expansion and early hyperplasia in 12W old mice but not at 6W,
> demonstrating that the adult gland provides a permissive temporal window for initial tumour
> growth.

**Figs. 2A--2D and S2A are bench material.** They are also the reason the rest of this chapter
matters: they establish that the change between six and twelve weeks is a change in
**competence at constant dose**, which the MMTV-Myc transcriptome --- a *dose* experiment ---
cannot show on its own. Everything below asks what in the wild-type gland could account for it.

## The grammar four of these panels share: two timelines {#sec-s3-grammar}

The reworded text puts **both timelines side by side in every claim**, so four of the
alternatives below use one plane:

- **x = `6>12W_wt`** --- what the normal gland does between six and twelve weeks
- **y = `6>12W_myc`** --- what the Myc+ gland does over the same window
- **the identity line `y = x` = *development alone***

**The vertical drop from that diagonal IS the interaction, exactly.** Script 40's ruler
satisfies `c_tp = c_tn + c_int` to floating point (1.2e-16 on content, bit-identical on
mitoPPS, asserted in every panel that uses the plane). So a point below the line has lost more
than development explains, and the distance is the genotype × time term itself --- not an
approximation of it.

`coord_equal()` is not cosmetic here: both axes are the same quantity in the same units, so
equal scaling is honest *and* it pins the diagonal at exactly 45°.

::: {.callout-warning}
## And this is where the batch confound bites hardest

**Both axes of this plane are batch-confounded** (batch = timepoint, @sec-constraints). What
is batch-clean is the **distance from the diagonal**, because the shared 6W→12W offset
cancels in the difference. Read the plane as *"how far below the line"*, never as *"how far
along the x axis"*.
:::

The device is declared once as `two_timeline_base()` in `figures/panels/_panel_common.R`, so
the four panels cannot drift apart.

## Fig. 2E --- the substrate: TEB lost, proliferation not {#sec-s3-2e}

> Further analysis of the transcriptomic basis underlying the tumour expansion in the adult
> gland identified a striking metabolic shift in the WT mitochondrial transcriptome and a
> complex interaction between the WT and MYC-transformed developmental trajectories determining
> the cell death outcome. First, the 6\>12W\_wt comparison showed that while the TEB signature
> was lost as expected in the puberty-adult transition, the overall proliferation signalling
> remained relatively stable (**Fig. 2E**), enabling MYC driven proliferation
> (***see Fig. 2B-D***).

```{r}
#| label: fig-2e
#| echo: false
#| fig-width: !expr panel_w("fig2_wt_teb_proliferation")
#| fig-height: !expr panel_h("fig2_wt_teb_proliferation")
invisible(draw_here("fig2_wt_teb_proliferation"))
print(panel_p("fig2_wt_teb_proliferation"))
```

```{r}
#| label: leg-2e
#| echo: false
#| output: asis
cat(panel_legend_md("fig2_wt_teb_proliferation"), sep = "\n")
```

**This is the negative limb that keeps Fig. 2F honest.** Had proliferation collapsed here, the
respiratory withdrawal would be its shadow rather than a finding.

::: {.callout-warning}
## "relatively stable" is right, and it needs one qualifier {#txt-prolif}

On the enrichment ruler it is as clean as it gets: **not one of the 14 curated proliferation
sets is significant** on `6>12W_wt`, and the family median (−0.89) sits inside the ±1 band an
unmoved set occupies --- against the TEB programmes' **−1.58 with 29 of 38 significant**.

But on the gene-level ruler, pooled `PROLIF_*` (731 genes) moves **−0.047 log2, the 1.3rd
percentile of its own expression-matched null, p = 0.013**. It is *small, not immobile* ---
script 43's own phrase. Suggested: keep "relatively stable" and make it quantitative ---
**the TEB arm moves 8.9 times as far**.

**A second observation the text should use.** The generic proliferation regulons agree with
the curated sets and the *lineage-context lanes of the same transcription factors do not*:
`TFT_E2F1_DOROTHEA` −1.18 and `TFT_E2F1_CHIPATLAS` +0.86 are both n.s., while the Gray `_LE`
lanes fall hard (FOXM1_HS_LE −2.10, TFDP1_HS_LE −2.08, MYBL2_HS_LE −2.04, E2F1_HS_LE −1.72,
all padj \< 1e−3). Those are lineage-identity lanes, not cell-cycle sets. **What the normal
gland withdraws from is lineage and morphogenesis; the cell cycle itself does not move.**
:::

## Fig. 2F --- the respiratory withdrawal {#sec-s3-2f}

> Second, the normal gland development showed a specific mitochondrial transcriptomic decline
> pattern. All OXPHOS subunit logFC and MitoPPS dropped significantly, while the overall
> mitochondrial transcriptome abundance remained relatively stable. Similar reduction of the
> OXPHOS complexes were observed at the protein level (*see* ***Fig. S1D***) At the same time,
> the maturing gland upregulated amino-acid and lipid catabolism. Specifically, fatty-acid
> oxidation, the carnitine shuttle and glycine cleavage all increased, shifting metabolism to
> feed electrons into a repressed respiratory chain, likely underlying a catabolic, highly
> reductive state (**Fig. 2F**). Importantly, the WT temporal program operated independently of
> the MYC-driven changes (**Fig. S2B**). While OXPHOS subunits were downregulated in the MYC
> 6→12W timeline in accordance with the reduced Myc effect, the normal maturation of the
> mammary gland added a further reduction in OXPHOS in the WT 6→12W timeline, suggesting a
> critical role for MYC-driven oncogenesis.

::: {.callout-warning}
## Two candidates for this slot --- not yet chosen (2026-08-09)

|  | **original** `fig2_wt_mito_contraction` | **alternative** `fig2_two_timelines_mito_alt` |
|---|---|---|
| **what it draws** | all 143 MitoPathways on the **content × mitoPPS plane**, one timeline (`6>12W_wt`) | all 143 on the **two-timeline plane**, one ruler (content) |
| **what it shows that the other cannot** | the **two rulers agree** (Spearman 0.82) --- a content fall alone could be normalisation, a content-blind ratio fall alone could be a reshuffle inside a growing compartment | the Myc-specific **extra** loss: OXPHOS ends deepest because the two terms **add** (−0.255 in development, −0.150 more from MYC, −0.405 under MYC) |
| **what it costs** | says nothing about the Myc timeline, which is half the sentence | drops the second ruler, so "logFC **and** MitoPPS" has to move to the legend |

**They are not variants of one picture.** Each drops what the other keeps. Drawing both rulers
*and* both timelines at 89 mm is what made the first attempt unreadable.
:::

### Fig. 2F --- original

```{r}
#| label: fig-2f
#| echo: false
#| fig-width: !expr panel_w("fig2_wt_mito_contraction")
#| fig-height: !expr panel_h("fig2_wt_mito_contraction")
invisible(draw_here("fig2_wt_mito_contraction"))
print(panel_p("fig2_wt_mito_contraction"))
```

```{r}
#| label: leg-2f
#| echo: false
#| output: asis
cat(panel_legend_md("fig2_wt_mito_contraction"), sep = "\n")
```

### Fig. 2F (alt) --- the two timelines

```{r}
#| label: fig-2f-alt
#| echo: false
#| fig-width: !expr panel_w("fig2_two_timelines_mito_alt")
#| fig-height: !expr panel_h("fig2_two_timelines_mito_alt")
invisible(draw_here("fig2_two_timelines_mito_alt"))
print(panel_p("fig2_two_timelines_mito_alt"))
```

```{r}
#| label: leg-2f-alt
#| echo: false
#| output: asis
cat(panel_legend_md("fig2_two_timelines_mito_alt"), sep = "\n")
```

::: {.callout-important}
## The statistic that belongs in this sentence is the PAIRED null, not either arm's own

"The respiratory chain falls **while** biogenesis does not" is a *comparative* claim, and
script 43 tests it by redrawing both sets together and testing the difference:

- **OXPHOS subunits − mitoribosome = −0.230, percentile 0.05, p = 0.0005**
- OXPHOS subunits − nucleotide metabolism = −0.255, p = 0.0085
- OXPHOS subunits − pooled proliferation = −0.208, p \< 0.0005

**Per pathway, nothing survives BH on either ruler --- 0 of 143** (smallest adjusted p 0.12 on
mitoPPS). So the text must not attach a p-value to any single MitoPathway. What the per-pathway
level *does* show is **coherence**: 23 pathways clear an unadjusted 0.05 against 7.2 expected,
binomial **p = 8e−7** --- the compartment moves together and no single pathway carries it.
:::

::: {.callout-warning}
## Four wording items in this paragraph {#txt-2f}

- **"across all complexes" has one exception.** CI, CIII, CIV and CV subunits all fall on both
  rulers; **CII subunits rise** (+0.088 content, +0.026 mitoPPS). Complex II is the only
  respiratory complex with no mtDNA-encoded subunit and the only one outside the proton
  circuit --- but the set is **four genes**, so it is a direction to note, not a mechanism.
  One-word fix: *"across the four complexes that carry mtDNA-encoded subunits"*.
- **The missing clause is the background.** The median MitoPathway **gains** content over this
  window (+0.041, 72 % above zero), so the respiratory arm falls *against a rising
  compartment*. That is what licenses the verb "withdraw", and it is one clause.
- **The assembly-factor control deserves a half-sentence**, not just the panel: the assembly
  factors of the same complexes sit at **+0.001 / −0.002**, percentile 50.2 of the matched
  null. *The gland withdraws the stoichiometry of the chain, not the machinery that builds it.*
- **The mtDNA arm is not on the panel and the text should not quote it either.** It is the
  largest movement in the compartment (+0.633 content, +0.525 mitoPPS) and it is the one
  quantity that is time-associated rather than genotype-associated --- so a temporal contrast
  is exactly where it cannot be read. See @sec-died.
:::

## Fig. S2B --- and the independence claim needed a control {#sec-s3-s2b}

```{r}
#| label: fig-s2b
#| echo: false
#| fig-width: !expr panel_w("figS2_reallocation_independence")
#| fig-height: !expr panel_h("figS2_reallocation_independence")
invisible(draw_here("figS2_reallocation_independence"))
print(panel_p("figS2_reallocation_independence"))
```

```{r}
#| label: leg-s2b
#| echo: false
#| output: asis
cat(panel_legend_md("figS2_reallocation_independence"), sep = "\n")
```

**The sentence is right on both scopes --- but only after a control.** On the 143 MitoPathways
the correlation is **−0.022** (mitoPPS) and −0.038 (content). On 15,191 genes the raw
correlation is **+0.228**, and that number is a property of the design, not of the biology:
`myc_6W` = 6W_myc − 6W_wt and `6>12W_wt` = 12W_wt − 6W_wt share the **same six wild-type
animals** in the subtraction.

| | value |
|---|---|
| observed | **+0.228** |
| predicted from the standard errors (Cov = Var(6W_wt) = SE²/2) | **+0.198** |
| **measured, split baseline** --- MYC effect against one half of the six WT animals, temporal effect against the other | **−0.025** |

The split-baseline control is **exhaustive, not sampled** --- all `choose(6,3) = 20`
assignments --- so the panel has no RNG in it.

::: {.callout-important}
## One bound that must not be dropped

This is **not** the statement that the Myc+ gland's own temporal change is independent of the
wild-type one. That regression (Myc+time ~ WTtime) has slope **0.749** --- much of the Myc+
gland's drift *is* the wild-type drift. Different pairs of vectors; they must never be run
together in one sentence.

And **never flip the axes**: the shared-vector regression gives 0.749 / −0.172 one way and
0.559 / +0.118 the other. It is not symmetric.
:::

## Fig. S2C --- the WT death substrate does not move {#sec-s3-s2c}

> Next, we asked whether these interactions between the WT and MYC driven trajectories could
> contribute to the time-dependent decline in apoptotic sensitivity. While the transcriptomic
> basis of the  overall apoptotic priming remained stable throughout the WT timeline
> (**Fig. S2C**), we detected a selective loss of a single trigger gene in the MYC 6 →12W
> timeline.

```{r}
#| label: fig-s2c
#| echo: false
#| fig-width: !expr panel_w("figS2_priming_balance")
#| fig-height: !expr panel_h("figS2_priming_balance")
invisible(draw_here("figS2_priming_balance"))
print(panel_p("figS2_priming_balance"))
```

```{r}
#| label: leg-s2c
#| echo: false
#| output: asis
cat(panel_legend_md("figS2_priming_balance"), sep = "\n")
```

**"Overall priming" is a balance, and both arms move together, so the balance does not move.**
Across the window the pro arm shifts **+0.025** and the anti arm **+0.045** on content
(difference **−0.020**); on the content-blind mitoPPS ruler, +0.054 against +0.046 (difference
+0.008). For contrast, **MYC moves the two arms apart**: +0.156 pro against −0.024 anti at six
weeks, a difference of **+0.180** --- an order of magnitude larger, and in the opposite
geometry. *The balance is something MYC changes and development does not.*

**One of 37 moves, and it moves the wrong way for a loss of priming:** `Bnip3` **+0.63, padj
0.021** --- a pro-apoptotic gene going *up*.

::: {.callout-note}
## A negative at n = 6 needs a power control, and this one has it

On the same transcripts, in the same libraries, at the same n, the **genotype** contrast
reaches padj \< 0.05 for **6 of 35** mapped genes, against **1** across the window. The
measurement can see movement in these genes; there is none to see here.

**And the bound that travels with it:** a negative on *transcripts* is not a negative on
*priming*. Priming is a property of the protein complement and of how close the mitochondrion
sits to the threshold. **BH3 profiling is the measurement**, and this panel is a reason to do
it rather than a substitute for it (@sec-open).
:::

## Fig. 2G --- the ratios, and the one that reverses {#sec-s3-2g}

> Specifically, most apoptotic priming ratios established by MYC at 6W remained stable since
> both pro- and anti-apoptotic proteins diminished in accordance with the global rescaling of
> the MYC effect (**Fig. 2G**), but the PUMA (*Bbc3*)/BCL-XL ratio showed a striking reversal,
> deviating significantly from the expected pattern. This reduction was highly specific to the
> *Bbc3* transcript, as other BH3-only sensors did not show similar trends. Notably, *Bbc3*
> mRNA levels remained constant in the WT gland but decreased significantly within the
> 6\>12W\_myc context (-0.48, padj = 0.016), demonstrating a specific interaction between
> genotype and timeline effects (interaction p \< 0.01).

::: {.callout-warning}
## Two candidates for this slot --- not yet chosen (2026-08-09)

|  | **original** `fig2_priming_ratios` | **alternative** `fig2_death_two_timelines_alt` |
|---|---|---|
| **what it draws** | seven pro:Bcl-xL **ratios**, MYC effect at 6W against 12W, with the **global rescaling line** (slope 0.487) as the expected pattern | the death machinery **per gene** on the two-timeline plane |
| **the claim it makes** | *the ratio deviates from the rescaling* --- and it refutes "it is just the global attenuation" **from inside the panel**, because every ratio shares the Bcl-xL denominator | *the numerator moves and the denominator does not* --- `Bbc3` +0.06 → −0.48 while `Bcl2l1` sits at the origin, so the reversal is **all numerator** |
| **what it costs** | cannot show that `Bcl2l1` is flat, which is the cleanest form of "specific" | loses the rescaling line, so "in accordance with the global rescaling" has no drawn reference |
| **a bonus only the alt has** | --- | `Bmf` (+1.02 → +0.84, significant on both) sits **on** the diagonal: the one death gene the window itself moves, and the panel's own negative control |

Different claims, not different drawings of one claim.
:::

### Fig. 2G --- original

```{r}
#| label: fig-2g
#| echo: false
#| fig-width: !expr panel_w("fig2_priming_ratios")
#| fig-height: !expr panel_h("fig2_priming_ratios")
invisible(draw_here("fig2_priming_ratios"))
print(panel_p("fig2_priming_ratios"))
```

```{r}
#| label: leg-2g
#| echo: false
#| output: asis
cat(panel_legend_md("fig2_priming_ratios"), sep = "\n")
```

| ratio | 6W | 12W | retention | |
|---|---|---|---|---|
| `Bax` | +0.901 | +0.495 | **0.55** | on the line |
| `Bid` | +0.946 | +0.396 | 0.42 | on the line |
| `Bak1` | +0.688 | +0.185 | 0.27 | a little under it |
| **`Bbc3`** | **+0.674** | **−0.061** | **−0.09** | **crosses zero** |

### Fig. 2G (alt) --- the death machinery on the two timelines

```{r}
#| label: fig-2g-alt
#| echo: false
#| fig-width: !expr panel_w("fig2_death_two_timelines_alt")
#| fig-height: !expr panel_h("fig2_death_two_timelines_alt")
invisible(draw_here("fig2_death_two_timelines_alt"))
print(panel_p("fig2_death_two_timelines_alt"))
```

```{r}
#| label: leg-2g-alt
#| echo: false
#| output: asis
cat(panel_legend_md("fig2_death_two_timelines_alt"), sep = "\n")
```

::: {.callout-warning}
## "Significantly" needs its scope, and "specific" needs one word {#txt-2g}

**Significance.** The PUMA:Bcl-xL **ratio** interaction is p = 0.037 raw, 0.017
purity-adjusted --- but its **BH across the nine pairs is 0.33**, and against script 42's
matched-pair null (random pro-like/anti-like pairs matched on expression, conditioned on a
six-week effect at least as large) the **empirical p is 0.082**. It is a nominal result with a
**pre-specified licence** --- PUMA was named in advance from the PGC1a cell experiments --- not
a multiplicity-surviving one. The honest counterweight is that on the empirical null the most
extreme pair is `Bmf` (p = 0.025), a ratio MYC never established (p6 = 0.40), which is exactly
why the *conditional* null is the one to read.

**Two numbers must not be interchanged.** The gene-level `Bbc3` interaction (**p = 0.0081**)
is a different statistic from the *ratio* interaction (**p = 0.037**). The text's
"(interaction p \< 0.01)" is the gene-level one and is correct as written --- but the
**BH-adjusted value is 0.84**, and the text should say once that pre-specification is what
licenses the test.

**"Highly specific to the Bbc3 transcript" needs a qualifier.** `Bax` **also falls
significantly under MYC** (−0.376, padj 0.008) --- it is an effector, not a sensor, so
"specific" holds **among the BH3-only sensors** and the sentence should say so. And `Bmf`
rises in **both** timelines, which belongs in the text as the contrast to `Bbc3`.
:::

## Figs. 2H and 2I --- Foxo3, and the coupling {#sec-s3-2hi}

> Crucially, the expression of *Bbc3* closely mirrored that of *Foxo3* (**Fig. 2H**), its
> recognized p53-independent activator, which yielded the lowest interaction p-value (0.0074)
> among all biogenesis- or cell-death-related genes. Although *Foxo3* was upregulated in the
> maturing WT gland (6\>12W\_wt padj \> 0.01), its expression was reduced along the 6\>12W\_myc
> timeline. Meanwhile, no transcriptional alterations were detected among other established
> upstream inducers of PUMA (**Fig. S2D**).

> Altogether, the transcriptomic analysis unequivocally identified the apoptotic arm as the
> primary driver of the shifting oncogenic balance in the adult gland. The analysis suggested
> that apoptotic sensitivity declined substantially due to loss of PUMA (*Bbc3*) in 12W\_myc
> mice. The decline could not be explained by the halving of MYC dosage, and was coupled to a
> metabolic shift driven by the loss of OXPHOS subunits, coupled to the transcriptional
> repression of *Foxo3*. Indeed, the interaction between MYC and OXPHOS subunit coupling to the
> PUMA/Bcl-XL ratio is highly significant (p = 0.0052) (**Fig. 2I**), whereas no such link
> exists for other redox and metabolic axes. These results led to the hypothesis that high
> respiratory state is required for MYC-PUMA mediated mammary epithelial cell death, and the
> reduction of respiratory capacity (OXPHOS subunits) in the adult gland is crucial for
> MYC-induced oncogenesis.

::: {.callout-warning}
## Two candidates for these slots --- and the pairing is TWO-TO-ONE

The alternative **merges Fig. 2H and Fig. 2I into a single panel**. Choosing it frees a slot.

|  | **originals** `fig2_departure_from_dose` (2H) **+** `fig2_oxphos_puma_coupling` (2I) | **alternative** `fig2_puma_chain_alt` (2H+I) |
|---|---|---|
| **the framing of `Foxo3`** | the **8,774-gene** departure-from-dose scan: `Foxo3` z −2.53, **percentile 0.46**; `Bbc3` −2.49, 0.51; only **four genes lie between them** | **rank 1 and rank 2 of the 26** biogenesis/cell-death mechanism genes by interaction p (0.0074, 0.0081; gap to third `Pgs1` at 0.066) --- the framing the reworded text already uses |
| **what it draws** | 2H a distribution with three genes marked; 2I one point per animal, mitoPPS against the ratio, a line per genotype | the chain as **three distributions in the order it runs** (OXPHOS mitoPPS → `Foxo3` → PUMA:Bcl-xL) with six within-genotype contrasts and three interactions, a 2 × 4 median heat map of the ratio's two members, and the coupling beneath |
| **what it costs** | two slots, and 2H's percentile framing is a **rank statement, not a test** | loses the genome-wide context that makes `Foxo3`'s position striking |
| **what only the alt shows** | --- | that the respiratory arm falls on **both** timelines, `Foxo3` rises **only** in WT, and the ratio collapses **only** under MYC --- the text's own sentence, in one picture |

`Bcl2l11` is on the 2H original because the scan was **pre-registered with two genes** and
**they split**: `Bbc3` at percentile 0.51, `Bcl2l11` in the ordinary middle at 91.6 with an
interaction p of 0.86. Drawing only the half that worked would be the wrong panel.
:::

### Fig. 2H --- original

```{r}
#| label: fig-2h
#| echo: false
#| fig-width: !expr panel_w("fig2_departure_from_dose")
#| fig-height: !expr panel_h("fig2_departure_from_dose")
invisible(draw_here("fig2_departure_from_dose"))
print(panel_p("fig2_departure_from_dose"))
```

```{r}
#| label: leg-2h
#| echo: false
#| output: asis
cat(panel_legend_md("fig2_departure_from_dose"), sep = "\n")
```

### Fig. 2I --- original

```{r}
#| label: fig-2i
#| echo: false
#| fig-width: !expr panel_w("fig2_oxphos_puma_coupling")
#| fig-height: !expr panel_h("fig2_oxphos_puma_coupling")
invisible(draw_here("fig2_oxphos_puma_coupling"))
print(panel_p("fig2_oxphos_puma_coupling"))
```

```{r}
#| label: leg-2i
#| echo: false
#| output: asis
cat(panel_legend_md("fig2_oxphos_puma_coupling"), sep = "\n")
```

| axis | wild type | Myc+ | script 43's interaction |
|---|---|---|---|
| **OXPHOS subunits** | **−4.80** | **+4.35** | **+6.09, p = 0.0052** |
| redox | −7.70 | −6.72 | +2.16, p = 0.72 |

**Redox is the right control precisely because it is not a null axis.** Both genotypes couple
to it, and strongly. What it does not do is couple *differently*. An axis nothing couples to
would not have excluded the possibility that the PUMA ratio simply tracks mitochondrial scores
in general.

### Fig. 2H+I (alt) --- the chain, and the coupling beneath it

```{r}
#| label: fig-2hi-alt
#| echo: false
#| fig-width: !expr panel_w("fig2_puma_chain_alt")
#| fig-height: !expr panel_h("fig2_puma_chain_alt")
invisible(draw_here("fig2_puma_chain_alt"))
print(panel_p("fig2_puma_chain_alt"))
```

```{r}
#| label: leg-2hi-alt
#| echo: false
#| output: asis
cat(panel_legend_md("fig2_puma_chain_alt"), sep = "\n")
```

| | wild type 6\>12W | Myc+ 6\>12W | difference (interaction) |
|---|---|---|---|
| OXPHOS mitoPPS | −0.11, p 0.108 | −0.15, **p 0.027** | p 0.615 |
| `Foxo3` | +0.46, **p 0.0044** | −0.00, p 0.982 | **p 0.013** |
| PUMA:Bcl-xL | +0.17, p 0.517 | −0.50, **p 0.019** | **p 0.042** |

::: {.callout-important}
## Asked and answered: is `Foxo3` 12W_wt vs 12W_myc significant? No --- and the reason IS the result

Drawn values −0.89 z, **p = 0.092**; DESeq2 −0.242 log2, p = 0.059, **padj = 0.53**. The 6W
genotype contrast is its **mirror**: +0.92 z (p = 0.076), DESeq2 +0.243 (p = 0.057).

MYC raises `Foxo3` at six weeks and lowers it at twelve by almost exactly the same amount.
**Neither half clears 0.05 alone, and that is precisely why the interaction is the strongest
term in the roster** (−0.486, p = 0.0074, rank 1 of 26). Labelling either genotype contrast on
a panel would print a failed test beside the claim. Neither is drawn.
:::

::: {.callout-important}
## Fig. 2I is ONE HALF of an interaction, and drawing one line misleads

The wild-type animals trend the **opposite way just as strongly**: R² 0.328, p 0.052, against
Myc+ R² 0.354, p 0.041. The result is a **genotype-by-slope interaction**, not a Myc-specific
correlation. That is why the alternative prints `interaction p = 0.0052` beside the R², in
neutral ink, while the R² takes the Myc+ colour.

**And read it as a lead.** The permutation null in the same object puts the interaction at the
**91.7th percentile of 5,000 draws (empirical p = 0.083)**, and adding a timepoint term moves
p from 0.0052 to **0.088**. It is the one axis-by-genotype interaction in the corpus that
reaches nominal significance --- which is a statement about the corpus as much as about the
axis.

**Do not quote the raw correlations.** Pooled across genotypes the PUMA ratio correlates +0.04
with the OXPHOS axis and −0.48 with redox --- the pooling artefact two opposite slopes produce.
:::

::: {.callout-warning}
## Four wording items in the closing paragraphs {#txt-2hi}

- **"whereas no such link exists for other redox and metabolic axes"** --- script 43's
  trade-off analysis carries **exactly two axes**, OXPHOS subunits and redox. The sentence
  should say *"for the redox axis"* unless a wider panel is run. What *is* broader is the
  outcome side: two axes against five outcomes, and only this one reaches p \< 0.05.
- **"6\>12W\_wt padj \> 0.01"** is a typo for **padj \< 0.01** --- `Foxo3` rises in the WT
  window at **+0.473, padj 0.0078**.
- **"unequivocally identified"** is stronger than the evidence. Nothing in the death arm
  survives genome-wide multiplicity; the licence throughout is **pre-specification**. Suggested:
  *"identified the apoptotic arm as the candidate driver"*.
- **The arrow is not the obvious one, and nothing here establishes direction.** PUMA restrains
  the mitochondrial pyruvate carrier (Kim, *Cancer Cell* 2019), so "PUMA falls, therefore
  respiration falls" is backwards; respiration sits upstream (Dey & Moraes), and FOXO3 → BBC3
  closes it into a **negative-feedback circuit** rather than a linear chain.
:::

::: {.callout-note}
## `Foxo3` was NOT pre-specified --- it was found here, and that is the main caveat

Forty-four genes sit below the 0.5th percentile of the departure scan with it, and they are
mostly unrelated (`Adh7`, `Agpat2`, `Aldh3a2`, `Ccnjl`, `Cntnap2`, `Inhba`, `Wfdc2`,
`Vtcn1`…). **Position alone is not evidence.** What makes `Foxo3` a lead rather than one of
forty-four names is the **prior** --- FOXO3 is PUMA's canonical p53-independent activator, with
direct ChIP evidence --- and the text should introduce it that way round.

The two routes to the same departure are worth keeping distinct: `Foxo3` is a **failure to
follow the normal developmental rise**; `Bbc3` is an **active loss**.
:::

## Fig. S2D --- the inputs that did not move {#sec-s3-s2d}

::: {.callout-warning}
## Two candidates for this slot --- and they draw DIFFERENT ROSTERS

This is not a redraw of one control. It is a choice between two different negative controls.

|  | **original** `figS2_puma_inducers` | **alternative** `figS2_p53_arm_alt` |
|---|---|---|
| **roster** | `exclusions$puma_inputs` --- 12 genes: p53 family (`Trp73`), E2F (`E2f1`), the integrated stress response (`Atf4`, `Ddit3`, `Trib3`, `Chac1`, `Eif2ak3`, `Nupr1`, `Sesn2`) and the FOXO family (`Foxo1`, `Foxo3`, `Foxo4`) | `exclusions$p53_axis` --- 9 genes: `Trp53`, `Mdm2`, `Cdkn1a`, `Cdkn2a`, `Trp53inp1`, `Zmat3`, `Eda2r`, `Phlda3`, `Ccng1` |
| **the sentence it answers** | *"no changes among other established upstream inducers of PUMA"* --- literally the written sentence | *"the p53-**dependent** arm is not what moves PUMA"* --- which is what makes `Foxo3` the candidate |
| **axes** | the MYC effect at each age, joined | the two-timeline plane, so `Bbc3` and `Foxo3` can be drawn **as the two that do move** |
| **what only it shows** | that `Foxo1` and `Foxo4` do **not** do it, so `Foxo3` is not a family-wide effect | that `Bbc3` **leaves the band** and `Foxo3` **leaves the diagonal** --- two different ways of moving |

The original's roster **contains `Foxo3`**, so the panel is a negative *with one exception*,
and the exception is the preceding sentence. The alternative's roster excludes it by
construction and draws it as a reference point instead.
:::

### Fig. S2D --- original

```{r}
#| label: fig-s2d
#| echo: false
#| fig-width: !expr panel_w("figS2_puma_inducers")
#| fig-height: !expr panel_h("figS2_puma_inducers")
invisible(draw_here("figS2_puma_inducers"))
print(panel_p("figS2_puma_inducers"))
```

```{r}
#| label: leg-s2d
#| echo: false
#| output: asis
cat(panel_legend_md("figS2_puma_inducers"), sep = "\n")
```

### Fig. S2D (alt) --- the p53 arm on the two timelines

```{r}
#| label: fig-s2d-alt
#| echo: false
#| fig-width: !expr panel_w("figS2_p53_arm_alt")
#| fig-height: !expr panel_h("figS2_p53_arm_alt")
invisible(draw_here("figS2_p53_arm_alt"))
print(panel_p("figS2_p53_arm_alt"))
```

```{r}
#| label: leg-s2d-alt
#| echo: false
#| output: asis
cat(panel_legend_md("figS2_p53_arm_alt"), sep = "\n")
```

::: {.callout-warning}
## Two bounds, and one gene that must be named whichever panel is chosen {#txt-s2d}

- **`Trp73` rises +2.14 across the wild-type window at padj 0.0039** --- on a mean expression
  of **12 counts**, at the floor of what this design can measure, and on the batch-confounded
  contrast. Not a MYC effect and not on either panel's axis, but it should be named so the
  panel is not read as flatter than it is. (`Eda2r` −0.99 at padj 0.019 is the equivalent in
  the p53-arm roster.)
- **A transcript-level negative is not a pathway-level one.** The ISR and the p53 axis act
  substantially through protein stability, phosphorylation and localisation --- ATF4 in
  particular is translationally controlled and can be fully active with an unchanged message.
  What this excludes is a **transcriptional** re-routing of PUMA's inputs, which is the
  alternative the text raises, and no more than that.
:::

# What did not survive {#sec-died}

This chapter is the reason the record exists. Roughly a year of analysis produced a large
number of results that **do not appear in the manuscript**, and they are not absent because
they were uninteresting --- they are absent because they were **tested and failed**. Several
of them were in earlier drafts. Anyone re-reading this project will re-derive them unless the
failures are written down with their reasons.

They are grouped by **why** they failed, because the failure modes recur and the modes are
more useful than the individual results.

## Circular by construction {#sec-died-circular}

::: {.callout-important}
## The mito → death bridge correlated mitochondrial genes with mitochondrial genes

`pro_comp` --- the pro-apoptotic composite that every death coupling in this corpus was built
on --- **is a MitoCarta set**: `MITOCARTA_APOPTOSIS_PRO`, 25 genes, defined at `scripts/08`.
So is the mitonuclear imbalance. "The mitochondrial state couples to death priming" therefore
correlated **mito genes with mito genes**.

**Every death coupling in the corpus is exposed to this.** It is the single most important
thing on this page. The couplings were not weak; they were not measuring what they were said
to measure.
:::

**And "MYC raises apoptotic priming" (genotype p = 0.037 / 0.038) is the mitochondrial CONTENT
effect wearing a death label.** Against an expression-matched **MitoCarta** background,
`APOPTOSIS_PRO` rises **+0.196 against a background of +0.364 (z = −2.42, p = 0.012)** --- i.e.
*below average for a mitochondrial gene*. On the **404 non-mitochondrial** pro-death genes the
effect does not reproduce at all (**−0.106, p = 0.078**, wrong sign). This independently
reproduces script 23's mitoPPS de-prioritisation of apoptosis (−0.21, padj 0.022).

**A related construction trap in the loading analysis.** The dominant-axis "top loaders" are
**94 % `_MITO` lanes** --- MitoCarta subsets *by build* --- so their loading is a property of
how the set was made. `MYC_LE_MITO` loads 0.993 at rank 6; `MYC_AP_HE` loads 0.013 at rank
881. Same regulon, different construction. The axis is **mitochondria-led, not
mitochondria-specific**, and loading is covariation, not primacy (@sec-s1-axis).

## At the correlation ceiling --- or untestable {#sec-died-ceiling}

**Script 25's per-sample death couplings fail an empirical null.** Couplings to `pro_comp`
(bio_comp 0.83, MASC 0.78, imbalance 0.67) sit at the **54th--90th percentile** of couplings
to all 884 library sets, permutation **p 0.10--0.50**. The null median coupling is **0.79**.
At n = 12 within a timepoint, everything correlates with everything at 0.6--0.8.

**Script 36 found where the ceiling comes from, and it is one axis.**
`cor(mito_oxphos, mean of all 884 scores) = 0.968`; regress that common mode out and the
ambient coupling drops **0.80 → 0.13**, which is the chance floor. Not low n (chance floor
0.13), not curation (random pairs 0.54), not the genotype split (0.83 within `6W_pos` alone).

::: {.callout-important}
## The interpretability trap, and it re-ranked everything

When an axis **is** the global factor, removing the global factor leaves noise. `mito_oxphos`
has **r² = 0.94** with the global mean, so its global-adjusted coupling of −0.62 is
**noise-on-noise, not antagonism**. A global-adjusted coupling may only be read where the
residual fraction is **≥ ~0.10**.

**Consequence: "OXPHOS is central" is not false --- it is UNTESTABLE.** There is no separable
variance in which to test it. That is the cleanest possible statement of its failure, and it
is a stronger sentence than "it did not replicate".
:::

Three results were re-ranked by that finding:

- **`cholesterol/mevalonate → MB2 fork` is DEMOTED.** Its script-35 headline (+0.40 against a
  partial ceiling of 0.13, 96th percentile, p = 0.043) was a **raw-ambient artefact**;
  properly adjusted it is **+0.14**, "separable but weak".
- **`biogenesis is a bystander` HOLDS, and harder than published** --- +0.11 against its own
  ceiling of 0.22 (27th percentile); to the TEB axis −0.03, the 7th percentile, i.e. *below* an
  arbitrary programme.
- **`redox → MB2 fork` is the one robust testable lead** --- global-adjusted −0.63, same sign
  across all three quantifiers, axis r² = 0.03, and MYC-independent. Redox is the control axis
  precisely because MYC does not drive it (d = 0.08), which is why it has readable variance
  left.

**Of 96 partial-coupling tests, exactly 1 beats BH \< 0.05.** Nothing in the coupling layer is
significant at n = 24.

**And the MYC-vs-PGC1a contrast is untestable for the same reason** (script 38 PART D). PGC1a
and MYC biogenesis are **not separable**: raw r = 0.93, both strongly MYC-induced (β +0.57 /
+0.75), and the global-adjusted −0.03 is the over-removal trap again (resid_frac 0.01 / 0.06).
All biogenesis axes couple to priming identically. **That is the answer: it needs an
ESRRA/PGC1a perturbation, not n = 24 bulk.**

## The wrong null {#sec-died-nulls}

::: {.callout-warning}
## Four wrong nulls, and each one produced a published-looking number

1. **`binom.test(..., p = 0.5)`** on the directional cell-death test. Under a **global
   attenuation every MYC-induced gene has ΔLFC \> 0 by construction**, so 0.5 is not the null.
   The correct comparator is the genome-wide supporting fraction.
2. **`mu = 0`** for Gate 2's p = 0.023. Under the same global attenuation the null is not zero:
   against a matched background, **z = −1.27, p = 0.21**.
3. **Effective n.** Gate 2's 23 genes have a measured inter-gene correlation of 0.08, giving an
   **effective n of 8.4, not 23**. A correlation-aware test of the same question (CAMERA) gives
   **p = 0.298, FDR = 0.926**.
4. **A raw ambient** as the null for a *partial* coupling. Script 35's Q4 claim is about the
   partial, so the null has to be a **partial ceiling** --- and against it, "OXPHOS is central"
   sits at the 68th percentile (p = 0.32), i.e. at chance.
:::

## A shared-baseline or shared-vector artefact {#sec-died-shared}

**Issue #6's "34 % of the attenuation is wild-type convergence" is a shared-baseline
artefact.** On the matched control --- the same three-mouse 6W wild-type baseline shared
between the two contrasts, versus split --- the global term **collapses to chance**, with an
artefact of **+0.24 ± 0.14 in 19 of 20 partitions**. Only the biosynthetic arms survive, and
the divergent arms are untouched (artefact ~0).

**The same mechanism, drawn constructively, is Fig. S2B** (@sec-s3-s2b): +0.228 raw, −0.025
once the baseline is broken.

::: {.callout-warning}
## Three artefacts of this class, recorded so they are not re-derived

- `cor(WTtime, ΔMyc-effect)` is **algebraically forced** --- script 40's `$artifact_ledger`
  marks it "STRUCTURAL --- do not cite".
- **Never flip the axes of the shared-vector regression**: 0.749 / −0.172 one way, 0.559 /
  +0.118 the other. "Uniform offset" is exact only at slope 1.
- **Cosine is not correlation.** On the content ruler the uncentred cosine is +0.210 and the
  correlation is −0.038; both vectors have a large positive mean and the cosine picks up the
  shared offset. The sentence's word is *correlation*.
:::

## Confounded beyond repair by this dataset {#sec-died-confounded}

**`mt%` --- the mtDNA-encoded read fraction, 3.4--40 % across animals --- is three-way
confounded** and the three cannot be separated here: real mitochondrial content × a
proliferation denominator (r −0.61) × dissociation leak (r with the IEG signature +0.74).

The consequences are structural, because two other quantities **are** mt%:

| quantity | correlation with mt% |
|---|---|
| the mitonuclear imbalance | **−0.92** |
| the mtDNA arm | **+0.95** |

So their couplings to priming collapse on mt%-adjustment (+0.70 → +0.34; −0.71 → −0.20). **Only
nuclear-OXPHOS ↔ priming is mt%-robust** (+0.46 → +0.48) --- which is why that is the coupling
that reached Fig. 2I and the others did not.

**And the sign kills the obvious alternative story:** mt% ↔ priming is *negative*, so
prep-stress → death does not work.

::: {.callout-important}
## The mitonuclear imbalance has no supported contrast at all

Time p = 0.31, genotype p = 0.52, interaction p = 0.92. Its death coupling fails a null. The
bridge from it to death is **retracted as a mechanism**.

**And the "direction test" that appeared to support it was circular and was withdrawn on
2026-07-17.** That nuclear mitochondrial genes anticorrelate with mt% is *equally* what a real
mitonuclear imbalance predicts and what a shared-denominator artefact predicts. It is evidence
for neither. Do not resurrect it.
:::

**The pure developmental axis is confounded with batch** and no QC metadata exists to separate
them (@sec-constraints). **Technical versus biological in the global axis is likewise not
adjudicable** --- the axis is *mixed*: contamination cannot be biological, proliferation can.
Script 37's technical correction uses **RNA-derived surrogates**, and they absorb **r² = 0.67
of design-residual proliferation variance**, so the corrected lens removes real biology. That
is why the design-only lens stays primary and the correction is a second lens, not the answer.

## Fitted, saved, and never read {#sec-died-unread}

::: {.callout-warning}
## The priming interaction had been on disk all along

`death_timing_substrate.rds$h1$state_tests` has carried **PRO interaction p = 0.177** and
**priming interaction p = 0.263** since it was written. The narrative instead quoted **group
means** --- "~3× more at 6W", +0.48 against +0.15 --- which is the same data with the test
removed.

**The lesson is procedural, not statistical.** An object that is fitted and saved but never
read is worse than one that was never fitted, because its existence implies it was checked.
:::

**A related, still-live trap:** `results/death_timing_substrate.rds` **on disk is stale**.
Script 23 as committed builds an 11-row `convergence` table; the saved object has 9, so Parts
2b/5b have never been run. **Re-source script 23 before reading it.**

## It doesn't transfer {#sec-died-transfer}

Lesner et al. (bioRxiv 2026.07.13.738248) propose a hypoxia → MXI1/JUN route by which MYC
suppresses mitochondrial biogenesis. Script 39 tested it in these contrasts:

- **The hypoxia route does not transfer.** `HYPOXIA` is *lower* in Myc+, does not rise over the
  window, `Hif1a` is low, `Mxi1` and `Tfam` are flat, `Jun` is not suppressed. It is
  liver-specific.
- **Nuclear OXPHOS does decline MYC-specifically** (interaction NES −2.21, padj 7e−11) but
  **not hypoxia-linked** --- it is the general MYC-programme fade.
- **MYC raises mitochondrial turnover machinery** (`DYNAMICS` +2.23, padj 1e−8) **but not
  mitophagy specifically**: against a matched MitoCarta background, z −1.4 to −2.6. That is
  the content effect again.

**Net: the same MYC → mito-turnover relationship by a non-hypoxic route.** What Lesner
validates is the **question**, not our contrast --- and usefully so, since their TCGA
(flash-frozen) and mouse HCC data show the nuclear-up / mtDNA-down discordance is not a
dissociation artefact.

## Retracted claims from script 32 itself {#sec-died-s32}

The content result survived its own audit; three claims around it did not. Recorded because
they were written down before they were checked:

- "commissioned but unbuilt" --- withdrawn.
- the absolute-share form of the mitonuclear imbalance --- withdrawn.
- **the depth confound --- withdrawn.** Shares and DESeq2 both cancel depth by construction;
  the within-cohort correlation is ~0, and the pooled −0.47 is a **pooling artefact**. 12W is
  **not** the degraded cohort.

::: {.callout-note}
## And one methodological fix that silently changed set-level numbers everywhere

MitoCarta and GMT files carry **old gene symbols**; the expression matrix carries current ones.
Renamed genes were silently dropped during **set aggregation** (ATP synthase: 7 of 24 subunits;
OXPHOS ~12 %). Per-gene DESeq2 results were complete throughout, so nothing reverses --- the
effect was a **conservative under-count**. Fixed by `functions/reconcile_gene_symbols.R` and
wired into scripts 32/08/15/20/29/31 and the figure layer on 2026-07-24.

**Any new set-based analysis must use the reconciler.**
:::

# Where the written sentences and the data disagree {#sec-disagree}

The checklist to cut against. Every row was found while building the panel that draws it, and
each links to the full argument above. **None of these is a result changing** --- they are
words.

| # | the sentence | what the data say | where |
|---|---|---|---|
| 1 | "no major changes in the overall composition" | true and non-significant, but the WT change is **luminal decline**, not BMYO expansion | @txt-teb |
| 2 | "a clear TEB-ductal shift" | sample-level p = 0.24 in WT; "clear" rests on the matched-null ruler (0th percentile) | @txt-teb |
| 3 | "suppressed BMYO in favor of differentiation to LHS" | BMYO holds; the LHS half is a **reversal**, true at 12W and false at 6W | @txt-teb |
| 4 | "aligned almost entirely with mitochondria related terms" | **mitochondria-led**: 94 of the top 100 are build-tautological, non-mito growth programmes reach 0.98, redox loads 0.67 | @txt-axis |
| 5 | "revealing the leading role of mitochondrial remodeling" | loading is **covariation, not primacy**; script 37 says so in its own saved verdict | @txt-axis |
| 6 | "followed by biosynthetic metabolic pathways, E2F …" | **E2F (+2.08) sits above biosynthetic metabolism (+1.84)** --- swap them | @txt-tiers |
| 7 | "94% of 143" | mixes the 144-row (94.4 %) and 143-row (95.1 %) fractions --- pick one | @txt-content |
| 8 | "21--27%" | the *set* arms drawn run **+20 % to +33 %** --- widen it or name the two quantities | @txt-content |
| 9 | "(padj \< 0.001)" | a **pooled** genotype main effect; only biogenesis clears 0.05 at six weeks alone | @txt-content |
| 10 | "MYC significantly promoted … OXPHOS subunits (all complexes)" | **not significant** (padj 0.274), and not one complex clears 0.05. The only significant OXPHOS row is CIII assembly factors, **demoted** | @txt-oxphos-sig |
| 11 | "R2=0.51, 0.8, p \< 0.002" | correct --- but the **p attaches to the R², not the slope**, and "whole transcriptome" must pick 0.450/0.381 or 0.487/0.51 | @txt-whole |
| 12 | "not just maintained but enhanced" | **"enhanced" does not survive**: 81 % of all sets rise, the two named rise by the median amount, and a flatter ranking mechanically raises NES | @txt-enhanced |
| 13 | "the expression gap remained constant between 6W_myc and 12W_myc" | a gap is between **genotypes**; both readings are true, say which | @txt-gap |
| 14 | "the overall proliferation signalling remained relatively stable" | right, and worth quantifying: **the TEB arm moves 8.9× as far** | @txt-prolif |
| 15 | "All OXPHOS subunit logFC and MitoPPS dropped significantly" | **0 of 143 pathways survive BH on either ruler.** The statistic that carries the claim is script 43's **paired null** (−0.230, p = 0.0005) | @txt-2f |
| 16 | "across all complexes" | **CII rises** (+0.088) --- say "the four complexes that carry mtDNA-encoded subunits" | @txt-2f |
| 17 | *(missing clause)* | the compartment **gains** content over the window (+0.041, 72 % up), which is what licenses "withdraw" | @txt-2f |
| 18 | *(missing clause)* | the **assembly factors do not move** (+0.001 / −0.002, percentile 50.2) --- the gland withdraws the chain, not the machinery that builds it | @txt-2f |
| 19 | "highly specific to the *Bbc3* transcript" | holds **among the BH3-only sensors**; `Bax` also falls significantly (−0.376, padj 0.008), and `Bmf` rises on both timelines | @txt-2g |
| 20 | "(interaction p \< 0.01)" | correct (0.0081) --- but **BH is 0.84**, and the text should say once that **pre-specification** is the licence | @txt-2g |
| 21 | "6\>12W_wt padj \> 0.01" | typo for **padj \< 0.01** (`Foxo3` +0.473, padj 0.0078) | @txt-2hi |
| 22 | "no such link exists for other redox and metabolic axes" | script 43 carries **exactly two axes** --- say "for the redox axis" | @txt-2hi |
| 23 | "unequivocally identified" | nothing in the death arm survives multiplicity; suggested: "identified … as the **candidate** driver" | @txt-2hi |
| 24 | "no changes … among other established upstream inducers" | true, **with one exception, and the exception is `Foxo3` itself**, which is in the roster | @txt-s2d |
| 25 | "~3K DEGs" | **2,777** at FDR 10 %, **1,967** at FDR 5 %; the interaction contrast has **0** at either | @sec-s2-paradox |
| 26 | the 986 / 885 / 866 set counts | each panel uses a different one; 986 is right where it stands, the other two need naming | @txt-counts |

# Open questions, and the experiment that settles each {#sec-open}

Everything below is a question this dataset **cannot** answer, paired with the experiment that
can. This is the list the reduction should protect, because a halved Results still has to hand
the reader to the perturbations.

::: {.callout-important}
## The framing that makes this list useful

A number that fails its null still nominates a **testable proposition**. The value of this
analysis is that it **pre-specifies** what the perturbation experiments should measure --- and
pre-specification, not circularity, is the strongest position available at n = 24. Present the
competing readings side by side with the experiment that adjudicates each; the bioinformatics
sets up the perturbations, it does not stand alone.
:::

| # | question | why this dataset cannot answer it | the experiment |
|---|---|---|---|
| 1 | **Is the transcript-level PUMA loss a loss of PRIMING?** | priming is a property of the protein complement and of how close the mitochondrion sits to the threshold; transcripts are its substrate, not its measurement | **BH3 profiling**, 6W vs 12W, ± MYC. Pre-specified readout: PUMA-peptide sensitivity |
| 2 | **Is respiration upstream or downstream of PUMA?** | the two are not separable here --- PUMA restrains the mitochondrial pyruvate carrier (Kim 2019), so the naive arrow is backwards, and FOXO3 → BBC3 closes it into a feedback circuit | **ESRRA/PGC1a perturbation** against the MYC substrate. Script 38 PART D states explicitly that n = 24 bulk cannot do it |
| 3 | **Is `Foxo3` causal, or one of forty-four?** | it was **found**, not pre-specified; 44 genes share its position and are mostly unrelated. The prior (canonical p53-independent PUMA activator, direct ChIP) is what makes it a lead | `Foxo3` knockdown / constitutive-active in the iMMEC system; does PUMA follow? |
| 4 | **Does mitochondrial content actually fall, or only its transcripts?** | the mtDNA arm is the one quantity that is time- rather than genotype-associated, and mt% is three-way confounded (@sec-died-confounded) | **mtDNA qPCR** (and citrate synthase / TOM20 blot). Lesner et al. independently validate the discordance in flash-frozen TCGA and mouse HCC |
| 5 | **Is the protective route priming or MITOPHAGY?** | script 39 finds MYC raises turnover machinery but **not mitophagy-specifically** (that is the content effect); the two predict the same transcriptome | NRF1 / DRP1 / FUNDC1 at protein level; mitophagy flux reporter |
| 6 | **How much of the attenuation is compositional dilution?** | E2F and proliferation co-decline, and marker support exists (`Oxtr`, `Myh11`, `Trp63`) --- but bulk cannot separate a per-cell effect from a change in cell mixture | **deconvolution or single-cell** on the same window (deferred subproject, plan on disk) |
| 7 | **Is the developmental (time) reading real, or batch?** | **batch = timepoint, perfectly.** No dissociation batch, viability or RIN metadata exists on disk | a cohort processed in one batch, or archived QC metadata if it can be recovered |
| 8 | **Is the global axis technical or biological?** | it is **mixed** --- contamination cannot be biological, proliferation can --- and the RNA-derived correction absorbs r² = 0.67 of real proliferation variance | measured technical covariates (dissociation batch, viability, RIN), which do not exist for this cohort |

::: {.callout-note}
## Three negatives that travel with the death story and should not be lost

They are negatives, and they are load-bearing:

1. **The gland de-respires without de-proliferating** (−0.255 against −0.047, assembly factors
   flat). The withdrawal is **morphogenetic** --- TEB lumen clearance --- not proliferative.
   This *contradicts* the proliferation hypothesis and is the negative limb of the
   PGC1a/Bcl-xL sign reversal, not a control.
2. **There is no co-regulated OXPHOS/apoptosis module.** `CORE_MITO` is **depleted** 0.76× for
   apoptotic genes; `MYC_MITO` is **enriched** 2.68× and holds `Bax`, `Casp9`, `Diablo`,
   `Pmaip1`. The PGC1a regulon is flat while only its OXPHOS arm falls. The relationship is
   **dependence, not co-transcription**.
3. **PUMA is a solo.** It belongs to no module in any pre-registered question. What collapses
   with it is **luminal lineage identity** (`MG_HS_GRAY` −2.79, padj 1.7e−17); what is retained
   is the MYC programme. `Bcl2l11`, its co-pre-specified partner, sits at the 91.6th percentile
   --- the pair splits.
:::

# Recommendation --- what a halved paper can afford {#sec-recommend}

::: {.callout-warning}
## This chapter is a RECOMMENDATION, not a record

Everything above states what exists and what it supports. This chapter is a proposal, written
so the reduction does not start from a blank page. Overrule it freely --- and note that the
figure follows the text, so a claim the author cuts takes its panel with it.
:::

## The claim chain the closing conclusion actually needs

The section ends on: *"high respiratory state is required for MYC-PUMA mediated mammary
epithelial cell death, and the reduction of respiratory capacity (OXPHOS subunits) in the adult
gland is crucial for MYC-induced oncogenesis."* Working backwards, that needs six links and no
more:

1. **MYC drives a mitochondrial programme in this tissue** --- otherwise the paper is not about
   mitochondria. *Fig. 1D (ranking) + Fig. 1E (content).*
2. **MYC reprioritises within the compartment**, which is also what introduces the mitoPPS
   instrument that Section 3 depends on. *Fig. 1F.*
3. **The MYC effect halves, and it is dose** --- needed because the closing paragraph says the
   PUMA decline *cannot* be explained by it. *Fig. 1G + Fig. S1D (blot) + Fig. S1E.*
4. **The adult gland is permissive at constant dose** --- the whole point. *Figs. 2A--2D, bench.*
5. **The wild-type substrate changed: proliferation stable, OXPHOS withdrawn.** *Fig. 2E +
   Fig. 2F.*
6. **The trigger was lost, specifically, and it tracks respiration.** *Fig. 2G + Fig. 2H+I.*

## The per-slot call on the four alternatives

| slot | recommendation | why |
|---|---|---|
| **2F** | **the alternative** | the sentence *"the normal maturation added a further reduction in OXPHOS … suggesting a critical role for MYC-driven oncogenesis"* **is** the additive claim, and only the two-timeline plane draws it. The two-ruler agreement (Spearman 0.82) moves to the legend, where a robustness check belongs |
| **2G** | **the original** | the global-rescaling line refutes *"it is just the dose"* **from inside the panel**, and that refutation is load-bearing in the closing paragraph. Its weakness --- that it cannot show `Bcl2l1` is flat --- is covered by the 2H+I alternative's heat map |
| **2H + 2I** | **the alternative** | it **frees a slot** in a figure that is being halved; it uses the rank-1-of-26 framing the reworded text already uses, which is smaller and more meaningful than an 8,774-gene percentile; and it is the only panel that shows the three-part chain as one picture |
| **S2D** | **text-first, not figure-first** | the original matches the sentence **as written** ("other established upstream inducers"). The alternative answers *"the p53-dependent arm does not move"*, which makes `Foxo3` the candidate but **is not currently a written sentence**. Decide the sentence, then the panel follows |

**Choosing 2F(alt) + 2G(orig) + 2H+I(alt) is internally consistent**: the two-timeline plane
appears once for the metabolic claim and once for the death claim, and 2G's rescaling line is
the one panel on a different grammar --- which is right, because it is making a different kind
of argument.

## What that costs, panel by panel

| panel | recommendation |
|---|---|
| **1B** lineage / TEB | **keep** --- it is the developmental frame the whole paper rests on |
| **1C** pathway PCA | **to supplementary.** Its sentence is bounded three ways (@txt-axis) and the claim it is used for ("leading role") does not follow from a loading |
| **1D** NES ranking | **keep** --- with the E2F / biosynthetic order swapped |
| **1E** content | **keep** --- the one solid new result |
| **1F** mitoPPS reallocation | **keep** --- and rewrite the sentence per @txt-oxphos-sig; it also introduces the instrument 2F needs |
| **1G** rescaled not reshaped | **keep** --- it supplies the ×0.55 rate that 2G's expected line uses |
| **1H** NES preserved | **cut, or demote to supplementary.** "Enhanced" does not survive (@txt-enhanced), and what *is* true ("position held while amplitude halves") is a sentence, not a panel |
| **S1A** design | **keep** |
| **S1B** library | **keep** --- Methods-adjacent but it is where the 986 comes from |
| **S1C** axis loadings | **keep only if 1C stays**; they are one analysis |
| **S1E / S1F** transcript / network | **keep S1E** (the blot's control); **S1F to this document** --- a clean negative on an alternative nobody has raised in print |
| **2E** | **keep** |
| **S2B** independence | **keep** --- it supports a written sentence, and its split-baseline control is the only defence against an obvious objection |
| **S2C** WT priming balance | **keep** --- it is the negative that makes the Myc-specific loss specific |
| **S2D** | see the table above |
| *(uncited)* MB-fork specificity | **this document only**, unless the human-resemblance row of Fig. 1B is cited in the text |

**Net: Figure 1 goes from A--H (8) to A--G (7 with 1H cut, or 6 if 1C also moves).
Figure 2 goes from A--I (9) to A--H (8), of which four are bench.** Supplementary 1 loses one
or two panels; Supplementary 2 keeps three.

# Provenance {#sec-provenance}

## The panels this document drew

```{r}
#| label: tbl-panels
#| echo: false
# slug_slot_order() returns the slugs already IN slot order (main figures, then
# supplementary, then number, then letter), so build the table over its output
# directly -- reordering the table by it afterwards inverts the permutation.
legs <- stats::setNames(lapply(ALL_SLUGS, function(s) panel_env(s)$LEGEND), ALL_SLUGS)
ord  <- slug_slot_order(ALL_SLUGS, legs)

knitr::kable(
  data.frame(
    slot   = vapply(ord, panel_slot, character(1)),
    script = paste0("figures/panels/", ord, ".R"),
    size   = vapply(ord, function(s)
               sprintf("%g x %g mm", panel_env(s)$SIZE_MM[["width"]],
                                     panel_env(s)$SIZE_MM[["height"]]), character(1)),
    stringsAsFactors = FALSE),
  row.names = FALSE, col.names = c("slot", "script", "designed size"))
```

## The integrity checks {#sec-checks}

The document asserts its own completeness rather than claiming it. **If any of these fail the
render stops**, so a rendered document is a passed one.

```{r}
#| label: checks
#| echo: true
drawn <- unlist(DRAWN, use.names = FALSE)

# 1. Every panel script on disk appears exactly once.
stopifnot(!anyDuplicated(drawn), setequal(drawn, ALL_SLUGS))

# 2. Every alternative is paired with the original(s) it competes with, and the
#    pairing map matches what is actually on disk -- so a NEW alternative added
#    to figures/panels/ cannot land in this document unnoticed.
PAIRS <- list("Fig. 2F (alt)"   = "Fig. 2F",
              "Fig. 2G (alt)"   = "Fig. 2G",
              "Fig. 2H+I (alt)" = c("Fig. 2H", "Fig. 2I"),
              "Fig. S2D (alt)"  = "Fig. S2D")
slots <- vapply(ALL_SLUGS, panel_slot, character(1))
stopifnot(setequal(grep("\\(alt\\)$", slots, value = TRUE), names(PAIRS)),
          all(unlist(PAIRS) %in% slots))

# 3. ... and each pair is CONTIGUOUS in draw order, i.e. drawn in one section.
for (a in names(PAIRS)) {
  grp <- names(slots)[slots %in% c(a, PAIRS[[a]])]
  pos <- match(grp, drawn)
  stopifnot(max(pos) - min(pos) + 1L == length(pos))
}

# 4. Every drawn size is the panel's own save_panel_p() size, not a copy.
for (s in ALL_SLUGS)
  stopifnot(isTRUE(all.equal(panel_w(s) * 25.4, unname(panel_env(s)$SIZE_MM[["width"]]))),
            isTRUE(all.equal(panel_h(s) * 25.4, unname(panel_env(s)$SIZE_MM[["height"]]))))

cat(sprintf("%d panels, %d alternatives, all paired and contiguous.\n",
            length(ALL_SLUGS), length(PAIRS)))
```

## The analysis objects these panels read

```{r}
#| label: tbl-results
#| echo: false
rds <- list.files(here::here("results"), pattern = "\\.rds$", full.names = TRUE)
info <- file.info(rds)
o   <- order(info$mtime, decreasing = TRUE)
knitr::kable(data.frame(object  = basename(rds)[o],
                        size    = sprintf("%.1f MB", info$size[o] / 1024^2),
                        written = format(info$mtime[o], "%Y-%m-%d %H:%M"),
                        stringsAsFactors = FALSE),
             row.names = FALSE)
```

::: {.callout-warning}
## One freshness trap that is deliberately not guarded

`results/mitopps_scores.rds` is timestamped **one minute older** than
`results/gsva_scores.rds` although both came out of the same post-reconciliation re-run. The
project's `require_fresher_than()` guard would therefore wrongly stop **Fig. 1F**. That panel
is gated on `interaction_results.rds` instead, and asserts script 08's mitoPPS difference
**exactly** (max difference 0) --- an identity check, which is the stronger guard.
:::

## The pipeline

```{r}
#| label: tbl-scripts
#| echo: false
sc <- sort(list.files(here::here("scripts"), pattern = "^[0-9]{2}_.*\\.R$"))
knitr::kable(data.frame(script = sc), col.names = "scripts/")
```

**Scripts 00--12** are the trunk: load, DESeq2 (basic + interaction), QC, fGSEA, mitoPPS,
GSVA, cell death. **13--31** are Block A --- the broad exploratory arc and the step-by-step
revision (Issues #1--6), reviewed and tagged `block-a-reviewed`. **32--44** are the arc this
document reports: content proxies (32), the mt-axis and the coupling null (33), the death-priming
reassessment (34), the correlation ceiling (35), the coupling method of record (36), pathway
loading (37), mitoPPS priming and PGC1a (38), the Lesner test (39), the background/MYC
decomposition (40), MYC stability (41), the priming arm and TEB substrate (42), substrate
specificity and the trade-off (43), and the collapse module (44).

::: {.callout-note}
## Option A --- who runs what

Numbered analysis scripts under `scripts/` are **hand-run by the author in Positron**. The
figure layer and this document read what they wrote and never re-run them. That separation is
why a panel that needs a refreshed object **stops** rather than drawing a stale number.
:::

## Repository state

```{r}
#| label: git-state
#| echo: false
#| output: asis
# system2() pastes the arguments into a shell string, so ANY argument containing
# a space must be shQuote()d -- an unquoted --format=%h %ad splits and git exits
# 128 with an "ambiguous argument" that never reaches the document.
gi <- function(...) suppressWarnings(system2("git", c(...), stdout = TRUE, stderr = FALSE))
cat("```\n")
cat("branch  ", gi("rev-parse", "--abbrev-ref", "HEAD"), "\n", sep = "")
cat("commit  ", gi("log", "-1", shQuote("--format=%h  %ad  %s"), "--date=short"), "\n", sep = "")
cat("tags    ", paste(gi("tag"), collapse = ", "), "\n", sep = "")
cat("```\n")
```

## Session

```{r}
#| label: session
#| echo: false
sessionInfo()
```
