---
title: "Is the developmental OXPHOS decline a PGC1a change?"
subtitle: "Six panels for the collaborator discussion, and what they license in the experimental part"
author: "Szabadkai lab"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    toc-expand: 2
    number-sections: true
    code-fold: true
    embed-resources: true
    theme: cosmo
    fig-format: png
    fig-dpi: 300
---

```{r}
#| label: setup
#| include: false

library(here)
here::i_am("paper/biogenesis_axis_discussion.qmd")
source(here::here("figures", "panels", "_panel_common.R"))

PANEL_DIR <- here::here("figures", "panels")

# The panel-embedding mechanism is analysis_record.qmd's, unchanged, so the two
# documents cannot drift apart in how they render a panel. The ONLY difference is
# the glob: this document collects `^biogax.*\.R$` -- the discussion panels, which
# hold no manuscript slot and are deliberately outside rebuild_panels.R and
# panels_to_pdf.R.
.panel_cache <- new.env(parent = emptyenv())

panel_env <- function(slug) {
  if (!is.null(.panel_cache[[slug]])) return(.panel_cache[[slug]])
  f <- file.path(PANEL_DIR, paste0(slug, ".R"))
  if (!file.exists(f)) stop("no panel script: ", slug, ".R", call. = FALSE)
  old <- options(myc.fig.capture = TRUE)
  on.exit(options(old), add = TRUE)
  env <- new.env(parent = globalenv())
  source(f, local = env)
  if (!exists("p", envir = env, inherits = FALSE))
    stop(slug, " built no `p`", call. = FALSE)
  if (!exists("LEGEND", envir = env, inherits = FALSE))
    stop(slug, " carries no panel_legend() block", call. = FALSE)
  reg <- get0(".myc_panel_sizes", envir = globalenv())
  if (is.null(reg) || !exists(slug, envir = reg, inherits = FALSE))
    stop(slug, " did not register a size", call. = FALSE)
  sz <- get(slug, envir = reg)
  assign("SIZE_MM", c(width = sz$width, height = sz$height), envir = env)
  .panel_cache[[slug]] <- env
  env
}
panel_w <- function(slug) unname(panel_env(slug)$SIZE_MM[["width"]]  / 25.4)
panel_h <- function(slug) unname(panel_env(slug)$SIZE_MM[["height"]] / 25.4)
panel_p <- function(slug) panel_env(slug)$p
panel_legend_md <- function(slug) {
  md <- legend_md(panel_env(slug)$LEGEND, slug = slug)
  c("::::: {.callout-note collapse=\"true\"}",
    paste0("## Legend block --- ", sub("^### ", "", md[1])),
    md[-1], ":::::")
}

ALL_SLUGS <- sub("\\.R$", "", sort(list.files(PANEL_DIR, pattern = "^biogax.*\\.R$")))
DRAWN <- character(0)
draw_here <- function(s) { DRAWN <<- c(DRAWN, s); invisible(s) }

# Numbers quoted in the prose are READ FROM THE OBJECT, never transcribed.
BA <- readRDS(here::here("results", "biogenesis_axis_developmental.rds"))
if (is.null(BA$foxo_reach))
  stop("results/biogenesis_axis_developmental.rds predates PART I. Re-source ",
       "scripts/47_biogenesis_axis_and_the_developmental_oxphos_decline.R ",
       "in Positron, then render again.", call. = FALSE)
num <- function(x, d = 2) formatC(x, format = "f", digits = d)
sp  <- as.data.frame(BA$split)
core <- function(part, col) sp[[col]][sp$regulon == "CORE_MITO" & sp$part == part]
```

::: {.callout-important}
## What this document is

A question raised repeatedly across Block A and never closed, now closed, with the
six panels that close it. **It is a discussion document, not a manuscript section.**
None of these panels holds a figure slot; they sit outside `rebuild_panels.R` and
`panels_to_pdf.R` and do not appear in `paper/analysis_record.qmd`. Nothing here
re-runs an analysis --- each panel is sourced from `figures/panels/` and renders its
own legend block, so this document cannot drift from the scripts.

Full write-up with every number and its source:
`docs/2026-08-17_developmental_oxphos_decline_and_the_biogenesis_axis.md`.
Analysis: `scripts/47_biogenesis_axis_and_the_developmental_oxphos_decline.R`.
:::

# The question, and why it stayed open

The Results describe what changes in the gland between six and twelve weeks. That
description then carries the anti-apoptotic window: respiration falls, the PUMA
trigger is lost. The paper models the fall experimentally with **PGC-1a and NRF1**.

So a reviewer will ask, and we should ask first: **is the developmental change a
PGC1a-axis change at all?**

The prior says it should not be. The 6-to-12-week loss is OXPHOS-specific rather
than a general loss of mitochondrial content, and PGC-1a governs far more of the
mitochondrion than the respiratory chain. If the axis had turned down, much more
than the chain should have gone with it.

**The question stayed open for a whole analysis block because the two instruments
that look like they should answer it are both broken for this contrast**, and
nobody had measured how badly. The TF-target sets built by intersecting a factor's
programme with MitoCarta fall whenever OXPHOS falls --- by construction. And the
mammary-context TF sets report which cell state the gland is in, not which factor
is active. Panel D3 is that failure, measured.

# The answer

## The regulon splits in two directions {#sec-d1}

```{r}
#| label: fig-d1
#| echo: false
#| fig-width: !expr panel_w(draw_here("biogax_regulon_split"))
#| fig-height: !expr panel_h("biogax_regulon_split")
panel_p("biogax_regulon_split")
```

```{r}
#| output: asis
#| echo: false
cat(panel_legend_md("biogax_regulon_split"), sep = "\n")
```

**A change in a transcription factor's activity acts on its regulon. It cannot act
on a functional subset of the regulon.** Inside the ERRa/NRF1/GABP regulon the
`r core("n OXPHOS subunits", "n_genes")` respiratory subunits fall to the
**`r num(core("n OXPHOS subunits", "wt_null_pct"), 1)`th percentile** of a matched
null while the other `r core("rest", "n_genes")` members rise to the
**`r num(core("rest", "wt_null_pct"), 1)`th**, and the regulon taken whole does not
move (`r num(core("all", "wt_null_pct"), 1)`th).

All four regulons that actually contain the respiratory chain split the same way.
The three that do not split hold between two and nine subunits, so the test cannot
run --- read it as four of four. And the panel shows the test can detect an axis
change when there is one: the ER-associated mitochondrial regulon moves as a unit,
at the `r num(sp$wt_null_pct[sp$regulon == "DEVELOPMENTAL_MITO" & sp$part == "all"], 1)`th
percentile, upward.

## The gland withdraws from the chain, not from the organelle {#sec-d2}

```{r}
#| label: fig-d2
#| echo: false
#| fig-width: !expr panel_w(draw_here("biogax_arm_map"))
#| fig-height: !expr panel_h("biogax_arm_map")
panel_p("biogax_arm_map")
```

```{r}
#| output: asis
#| echo: false
cat(panel_legend_md("biogax_arm_map"), sep = "\n")
```

A withdrawal of mitochondrial biogenesis predicts that the subunits, the factors
that assemble them, the mitochondrial ribosome and the import channels all move
together --- they are one programme. **The structural subunits sit at the 1.75th
percentile as a class; the assembly factors that build those same complexes sit at
the 53rd.** Meanwhile every catabolic arm rises.

Two details are worth arguing over. **Complex II does not fall** --- the one
respiratory complex that neither pumps protons nor carries an mtDNA-encoded
subunit, and also a TCA enzyme. And **complex IV leads the withdrawal**. If the
withdrawal is specifically from proton-pumping capacity rather than from
"mitochondria", that is a sharper claim than the one currently in the text and it
is worth deciding whether to make it.

## The factors themselves do not move {#sec-d3}

```{r}
#| label: fig-d4
#| echo: false
#| fig-width: !expr panel_w(draw_here("biogax_factor_roster"))
#| fig-height: !expr panel_h("biogax_factor_roster")
panel_p("biogax_factor_roster")
```

```{r}
#| output: asis
#| echo: false
cat(panel_legend_md("biogax_factor_roster"), sep = "\n")
```

Two facts here bear directly on the experimental part.

**PGC-1a is not expressed in these cells.** `Ppargc1a` sits at baseMean
`r num(BA$roster$baseMean[BA$roster$sym == "Ppargc1a"], 0)`, against PRC's
`r num(BA$roster$baseMean[BA$roster$sym == "Pprc1"], 0)`. Whatever coactivator the
mammary epithelium uses, it is not PGC-1a. That does not stop PGC-1a being a good
instrument --- if anything it makes the gain-of-function cleaner --- but it means the
transgene level is not a calibration, and it is a question a reviewer may raise.

**PRC is a lead and not a result.** `Pprc1` is the only PGC-1 family member that
moves and it does so in both genotypes with an adjusted p of
`r formatC(BA$roster$wt_padj[BA$roster$sym == "Pprc1"], format = "e", digits = 1)`
--- but against genes of its own abundance it sits at the
`r num(BA$roster$wt_pct[BA$roster$sym == "Pprc1"], 1)`th percentile, inside the band.
The small p is precision at high expression, not a large effect. PRC has never been
examined in this project; it is worth an antibody, not a mechanism.

### The ERR family, and a candidate that had to be tested {#sec-errb}

`Esrrb` is the **highest riser in the whole roster** ---
`r num(BA$roster$wt_lfc[BA$roster$sym == "Esrrb"], 2)` log2, padj
`r num(BA$roster$wt_padj[BA$roster$sym == "Esrrb"], 3)`, the
`r num(BA$roster$wt_pct[BA$roster$sym == "Esrrb"], 1)`th expression-matched percentile --- and it
was missing from the first version of this ledger. The reason is worth stating: **the ledger was
built around what could have *lowered* the chain, and ERRb goes up**, so it could only qualify
through a repression or competition model, which was never considered.

The model is coherent. ERRa, ERRb and ERRg share the same response element, so a rising ERRb could
displace ERRa from its own sites --- and unlike FOXO3, **reach does not disqualify it**: through the
ERRE it would touch every gene ERRa touches, which is 49 of the 89 that fell.

It fails three tests, each already applied to the other candidates.

| test | result |
|---|---|
| within-animal coupling to OXPHOS | **+0.36 in wild type, −0.29 in Myc+** --- mid-range, and opposite signs. Displacement predicts a consistent negative. |
| the sibling control | `Esrrg` rises too (`r num(BA$roster$wt_lfc[BA$roster$sym == "Esrrg"], 2)`, `r num(BA$roster$wt_pct[BA$roster$sym == "Esrrg"], 1)`th pct) and couples **+0.75** to OXPHOS. The oxidative ERR points the wrong way for a repression story. |
| per-animal distribution | 12W wild types: **58, 85, 101, 312, 332, 696** --- 12-fold spread inside one group, against `Esrra`'s 1.2-fold. The median shift is real (82 → 207); the dispersion is a minority-signal profile, not a programme. |

There is also **no ESRRB target set in the library**, so its regulon has no direct read-out here,
and it tracks none of the state composites (AP_LE +0.13, TEB +0.25, basal +0.09 in wild type).

**What survives is a good experiment rather than a conclusion.** ERRb is a real, uncharacterised
developmental change that Myc suppresses at six weeks, and the test is sharper than anything the
PGC1a arm offers: **knock ERRb down in the twelve-week-like state and the respiratory subunits
should rise while the rest of the ERRa regulon stays put** --- the same split, run backwards. Worth
first checking whether the cell system expresses ERRb at all.

::: {.callout-note}
## The general lesson

A **rising** factor can only cause a **falling** programme through repression or competition, and
that whole class is easy to skip when a candidate list is organised by direction. Screen the risers
as well as the fallers, then discriminate with the within-animal coupling --- which is what demoted
this one, and what demoted PRC.
:::

# Why the TF layer could not answer it {#sec-d4}

```{r}
#| label: fig-d3
#| echo: false
#| fig-width: !expr panel_w(draw_here("biogax_tf_context"))
#| fig-height: !expr panel_h("biogax_tf_context")
panel_p("biogax_tf_context")
```

```{r}
#| output: asis
#| echo: false
cat(panel_legend_md("biogax_tf_context"), sep = "\n")
```

Context explains `r num(100 * BA$tf_layer$adj_r2_context, 0)` per cent of the
variance in lane score and the identity of the transcription factor
`r num(100 * BA$tf_layer$adj_r2_tf, 0)` per cent. Inside one context, a MICOS
structural protein and general transcription factor IIIA both outrank ERRa. **A
layer in which CHCHD3 beats ERRa cannot be used to nominate ERRa** --- or to
exonerate it, which is why the argument above rests on the regulon split and the
arm map instead.

# Is FOXO3 the strongest hypothesis? {#sec-foxo3}

```{r}
#| label: fig-d5
#| echo: false
#| fig-width: !expr panel_w(draw_here("biogax_foxo3"))
#| fig-height: !expr panel_h("biogax_foxo3")
panel_p("biogax_foxo3")
```

```{r}
#| output: asis
#| echo: false
cat(panel_legend_md("biogax_foxo3"), sep = "\n")
```

**FOXO3 is the strongest readable SIGNAL. It is not the strongest hypothesis for
the OXPHOS decline, and the distinction is worth holding onto.**

What makes it the strongest signal is that it survives both filters that destroy
everything else in the TF layer. Its target set contains **no OXPHOS subunit at
all**, so its rise cannot be mitochondrial content; and it carries no Gray context,
so it cannot be the cell-state artifact. It is the top of the 69 context-free lanes.
Gene level agrees: `Foxo3` at the
`r num(BA$roster$wt_pct[BA$roster$sym == "Foxo3"], 1)`th expression-matched
percentile, `Sirt1` at the
`r num(BA$roster$wt_pct[BA$roster$sym == "Sirt1"], 1)`th, both significant.

What stops it being the driver is the same test that disqualified the PGC1a axis as
the cause, applied consistently: **reach**. FOXO3's regulon contains
**`r BA$foxo_reach$covers_ox[BA$foxo_reach$set == "TFT_FOXO3_CHUNG"]` of the 89
subunits that fell**, against 59 for the ERRa/NRF1/GABP regulon. A factor that
touches none of the affected genes cannot be their proximal cause, however clean
its own signal is. **FOXO3 is the readable marker of the adult state in which
respiration is lower --- which corroborates the state hypothesis rather than
competing with it.**

Three further readings the panel forecloses, each of which is a story someone will
otherwise tell:

1. **"Myc blocks the developmental FOXO3 rise."** No --- the programme rises with
   age in *both* genotypes and the interaction is not significant. What Myc does is
   suppress it by a constant offset at both ages, which is a clean genotype effect
   and the part worth keeping.
2. **"The adult gland arms PUMA through FOXO3."** No --- `Bbc3` is flat in the
   wild-type timeline. PUMA falls in the *Myc+* timeline, which is a different
   statement.
3. **"The gland becomes quiescent, hence less respiration."** Only the atrophy and
   turnover arm of the programme moves; the cell-cycle-arrest arm does not move at
   all. Whatever this is, it is not a simple arrest programme.

::: {.callout-warning}
## Two things not to say about this panel

`Foxo3`'s **interaction adjusted p is 1**. The supported sentence is "the wild type
rises, the Myc+ gland does not" --- never "Myc blocks it". Two halves that differ in
significance do not differ significantly, which is the trap already recorded for
the Foxo3/PUMA pair.

And the two genotype rows must not be read as "the FOXO3 programme does not
attenuate". **NES is scale-free and cannot see an amplitude change**; equal NES at
both ages is not equal effect size.
:::

# What licenses the PGC1a/NRF1 rescue {#sec-rescue}

```{r}
#| label: fig-d6
#| echo: false
#| fig-width: !expr panel_w(draw_here("biogax_rescue_licence"))
#| fig-height: !expr panel_h("biogax_rescue_licence")
panel_p("biogax_rescue_licence")
```

```{r}
#| output: asis
#| echo: false
cat(panel_legend_md("biogax_rescue_licence"), sep = "\n")
```

If the axis did not cause the decline, the experiment needs its justification
stated. The strongest form is not that PGC-1a is a plausible mechanism --- it is not
the mechanism --- but that **the mouse cannot answer the question and the
intervention can.**

In the gland, respiratory expression is inseparable from the epithelial state that
carries it: carrier and cargo correlate at
**r = `r num(BA$carrier$cor_ox_apLE_nonmito, 2)`** with every mitochondrial gene
stripped from the carrier, a variance inflation factor of
`r num(BA$carrier$vif_apLE_nonmito, 1)`. No observational contrast in this animal
separates "respiration fell" from "the state that respires became rarer". An
intervention that moves respiration while cell state is held fixed is the only
instrument that breaks that. **The limitation is the licence.**

Three properties make the axis the right instrument, and all three are measured:

- **Reach.** It did not lower these genes but it reaches
  `r BA$rescue$reach$covers_ox[BA$rescue$reach$regulon == "CORE_MITO"]` of the 89
  that fell.
- **Convergence.** ERRa and NRF1 regulons share only
  `r BA$rescue$convergence_test$n_shared` genes and half of that overlap is
  respiratory subunits ---
  `r num(BA$rescue$convergence_test$fold, 2)`-fold over their union,
  p = `r num(BA$rescue$convergence_test$p_enrich, 4)`. Two interventions whose reach
  barely overlaps, giving one phenotype, localise the effect to what they share.
  **Using both factors is a control, not a redundancy.**
- **Non-circularity.** The death effectors are not in the axis: five of the 25
  pro-apoptotic MitoCarta genes, below expectation, with **no PUMA, no BAX, no
  caspase**. They are in MYC's regulon at 2.7-fold. So a restored death phenotype
  cannot be direct transcription of the machinery --- it has to run through the
  respiratory state, which is the claim under test.

::: {.callout-tip}
## The sentence for the paper

> Because respiratory gene expression in the gland is inseparable from the
> epithelial state that carries it, the developmental data cannot establish whether
> respiration itself is the operative variable. We therefore raised it directly.
> PGC-1a and NRF1 were chosen not to reproduce the developmental mechanism --- which
> is not an axis-activity change --- but because their regulons reach two thirds of
> the subunits that declined while excluding the apoptotic effectors, so a restored
> death phenotype cannot be direct transcription of the death machinery.
:::

# Decisions this puts to the group

1. **Does the Results text change?** Any sentence implying the developmental loss
   is a decline in PGC1a-driven biogenesis has to go. The replacement is stronger,
   not weaker: the axis is unchanged and only its respiratory arm falls.
2. **Do we make the sharper claim?** Complex II does not fall and the electron
   carriers rise. "The gland withdraws from proton-pumping capacity" is more
   specific than "from respiration" --- and more falsifiable.
3. **How is the rescue described?** As an instrument, with the reach and
   convergence numbers, or as a mimic. The first survives review; the second
   invites the objection this document exists to pre-empt.
4. **Do we pre-specify the overshoot in writing?** The developmental change was
   arm-selective and the rescue will not be: assembly factors, mitoribosome and
   central dogma never fell yet are 29 to 35 per cent inside the regulon. Naming
   that before the experiment keeps the claim to "raising respiratory capacity
   restores death competence".
5. **Is the carrier question worth an experiment?** Separating cell state from
   respiration needs sorted AP/HS/BA or single-cell. It is the one question bulk
   cannot answer, and it is the question a hostile reviewer will find.
6. **Is PRC worth a Western?** It is the only PGC-1 family coactivator this tissue
   expresses. It does not clear its own null, so it is not a result --- but if asked
   "which coactivator does the gland use", we currently cannot answer.
7. **Is ERRb worth following?** It is the sharpest developmental marker in the roster
   and the project has never looked at it. It is not a supported cause of the
   respiratory decline, but the knockdown prediction (@sec-errb) is the most
   specific one this analysis generates.

# Scope, stated once {#sec-scope}

```{r}
#| output: asis
#| echo: false
cat(BA$notes[grepl("^(SCOPE|What|and PART E|MAGNITUDE)", BA$notes)], sep = "\n")
```

**BATCH = TIMEPOINT.** Every between-age number here is confounded with cohort.
What survives that is the arm-selectivity --- a shared batch effect cannot move one
arm of a regulon to the floor and the rest to the ceiling --- and the within-timepoint
couplings, which remove the timepoint means entirely. What does not survive is the
*magnitude* of the developmental decline. n = 24, twelve per genotype: this is
ranking plus a set of negatives, not confirmatory inference.

One correction that belongs in the open: **the within-age coupling between
respiration and the cell cycle (r = 0.89) must not be quoted on its own.** In the
same twelve samples the OXPHOS composite correlates with the whole PGC1a regulon at
0.98 and with the mitochondrial ribosome at 0.97. Everything correlates with
everything at this n; only the gene-level percentile against all expressed genes is
readable.

```{r}
#| label: completeness
#| echo: false
#| output: asis

missing <- setdiff(ALL_SLUGS, DRAWN)
extra   <- setdiff(DRAWN, ALL_SLUGS)
dup     <- DRAWN[duplicated(DRAWN)]
if (length(missing) || length(extra) || length(dup))
  stop("panel bookkeeping: missing [", paste(missing, collapse = ", "),
       "] extra [", paste(extra, collapse = ", "),
       "] duplicated [", paste(dup, collapse = ", "), "]")
cat(sprintf("\n*%d discussion panels, each drawn once. Analysis run %s.*\n",
            length(ALL_SLUGS), as.character(BA$analysis_date)))
```
