---
title: "Mitochondria integrate oncogenic and metabolic transcriptional programs to shape breast cancer progression"
subtitle: "MMTV-Myc mammary timecourse -- exploratory synthesis and figure narrative"
author: "Szabadkai lab"
date: today
---

::: {.callout-note}
**Skeleton, not the finished document.** This is the container the reflective write-up
will grow into as the publication panels are finalised. It mirrors the exploratory arc of
scripts 32-39 and reuses the same `results/*.rds` objects and the `figures/` panel layer.
Each section below is a stub: prose to be written, panels to be embedded. The organising
principle, settled 2026-07: **genotype CONTRASTS survived; cross-sectional COUPLINGS sit
at a correlation ceiling and mostly did not.**
:::

```{r}
#| label: setup
#| include: false
library(here)
library(ggplot2)

# Pin here() to the repo root: this .qmd lives under paper/, and a knitr render
# starts the session there, so anchor explicitly (sourced panel scripts inherit it).
here::i_am("paper/myc_mito.qmd")

# Display panels inline without re-exporting their canonical PDFs to outputs/.
options(myc.fig.nosave = TRUE)

# Read-only load of the analysis objects. The figure layer never re-runs analysis;
# it reads what scripts 00-39 already wrote to results/.
res <- function(x) readRDS(here::here("results", paste0(x, ".rds")))

# Loaded lazily per section as the prose is written; listed here as the manifest:
#   content   <- res("mito_content_proxies")        # script 32 -- Fig content
#   mtaxis    <- res("mtdna_axis_and_coupling_null") # script 33 -- adjusted +21%
#   loading   <- res("pathway_loading")              # script 37 -- global axis / dim
#   coupling  <- res("linear_pathway_coupling")      # script 36 -- ceiling origin
#   lesner    <- res("lesner_mechanism_test")        # script 39 -- external validation
```

# Overview {#sec-overview}

*(stub)* One-paragraph orientation: bulk RNA-seq of enzymatically purified MMTV-Myc MECs,
four groups x six replicates, `~timepoint*myc_status` interaction design. Batch = timepoint,
so genotype/interaction contrasts are clean and the pure developmental reading is confounded.
What this document does: walk the exploratory process and land the defensible results.

# Myc raises and reprioritises the mitochondrial compartment {#sec-content}

The strongest new result: constant Myc (@fig-myc-dose) raises the **nuclear** mitochondrial arms (mass
markers, nuclear MitoCarta, nuclear OXPHOS, mito biogenesis) while mtDNA-encoded output stays
flat -- a content increase of ~+21% (p=0.0006) after adjusting for prep-stress and
contamination, and a *lower bound* because Myc amplifies total RNA per cell.

The biogenesis arm is **MitoCarta central dogma + protein import/sorting/homeostasis** (314
genes: mtDNA replication and transcription, mtRNA metabolism, mitochondrial translation and
ribosome, import, chaperones, proteases). It replaces the earlier mitoribosome facet for two
reasons: only **7 of its 314 genes** are in nuclear OXPHOS, so *OXPHOS vs general biogenesis*
is a genuinely disjoint comparison rather than a whole-vs-part one; and it is the same
`mito_biogenesis` axis the coupling analyses use (script 28), so figure and analysis name one
thing. Deliberately **not** the `Biogenesis_discrimination` panel (`CORE_MITO`, `MYC_MITO`,
`DEVELOPMENTAL_MITO`): those are regulons, and `MYC_MITO` is Myc targets by construction --
circular on a genotype axis.

```{r}
#| label: fig-content
#| fig-cap: "Myc raises the nuclear mitochondrial arms; mtDNA-encoded output stays flat. Genotype is the clean axis; the time axis is cohort-confounded. n=6/group."
#| fig-width: 9
#| fig-height: 4.4
#| echo: false
# Reuse the panel script: it builds `p` and (as a side effect) writes the PDF to
# outputs/figures/. Here we just display the same object inline.
source(here::here("figures", "fig01_mito_content.R"))
print(p)
```

The compositional critique of the share panel -- "the fraction can rise because *other*
genes fall" -- is answered by the same four comparisons in per-gene log2FC space
(`figures/fig01b_mito_content_lfc.R`, from `interaction_results.rds`): the arm genes rise
in **absolute** terms and **coordinately** (Myc@6W: 83-95% of each nuclear arm's genes up,
biogenesis highest at 95%; mtDNA-encoded genes are *not* induced -- 0-15% up), so the share
rises because the genes rise, not by subtraction.

::: {.callout-important}
## Two cross-panel tensions the narrative must own (both real, not artifacts)

1. **6W: LFC says strong, share says marginal -- except on the biogenesis arm.** fig01b shows
   369 individually significant up-genes in Nuclear MitoCarta at 6W, yet fig01's *share*
   bracket at 6W is only p=0.11 there (mass markers 0.085, nuclear OXPHOS 0.12). Per-gene
   induction is large; the n=6 aggregate *fraction* is noisy at 6W, so the genes clearly move
   where the share is underpowered. **Mito biogenesis is the exception and the arm to lead
   with**: p=**0.035** at 6W and p=**0.0015** at 12W -- the only arm significant at both, and
   the same arm fig01b independently ranks most coordinate (94.7% of genes up at 6W). Share
   space and LFC space agree once the arm is a real programme rather than a sub-arm; part of
   the old tension was arm choice, not only sample size.

2. **12W: significant counts undersell the coordinate shift.** OXPHOS@12W has a clearly
   positive box (87% of genes up) but only **1** gene clears FDR, because attenuation shrinks
   per-gene magnitude and n=6 limits power at 12W -- exactly where fig01's share is *firmest*
   (p 0.003). The `% up` line on fig01b (power-independent) is what keeps 12W from looking
   artificially weak; read it alongside the significant-gene counts. The biogenesis arm makes
   the point explicit: at 12W it sits at essentially the same 88% up as OXPHOS yet retains
   **14** FDR-significant genes to OXPHOS's 1 -- the counts are tracking set size and power,
   not a difference in biology. This previews @sec-attenuation.

Integrated reading: **6W = strong per-gene induction, noisy aggregate; 12W = attenuated
per-gene signal, consolidated compartment share.**

Not a tension but the mirror of the claim: the only significant *time* bracket in fig01 is
mtDNA-encoded within Myc+ (p=0.0079, 6W->12W), matching fig01b's 85% up. It is on the
batch-confounded axis (batch = timepoint) and stays a described observation, not a result --
see @sec-lesner for why the mtDNA arm needs qPCR to settle.
:::

*(stub: the content narrative -- mass-panel coherence, where the imbalance is NOT
[the mtDNA machinery is not the transcriptional bottleneck], the three ceilings, and why
mtDNA qPCR / blot is the settling experiment. Possible future panel: the four contrasts as
the sides of a square [Myc@6W / Myc@12W / WT-time / Myc+-time] -- a 2x2 contrast diagram.)*

# The mitochondrial-group survey: what rises when OXPHOS falls {#sec-mito-survey}

fig01 shows OXPHOS is the mitochondrial arm that *falls* across the WT 6->12W timeline,
yet the total nuclear MitoCarta share is flat -- so other groups must rise to hold it
there. This supplementary survey resolves which. It takes the seven canonical MitoCarta
MitoPathway top-level categories -- **with Metabolism split into its nine level-2 children**
(hierarchy read from MitoCarta Sheet 4) -- and reads them the same two ways as fig01/fig01b:
compartment share across the four groups (@fig-survey-share) and per-gene log2FC across the
four contrasts (@fig-survey-lfc). **OXPHOS is shown as its two halves** -- nuclear OXPHOS
(`MITOCARTA_OXPHOS_NU`, 155 genes, no mt-* genes) and mtDNA OXPHOS (the 13 mtDNA-encoded
subunits, which *are* `MITOCARTA_MTDNA_ENCODED`) -- so the nuclear category carries no mtDNA
bias and there is no redundant "mtDNA reference" facet. `share_nomt` (denominator drops only
the 13 mt-* genes) is the consistent scale.

```{r}
#| label: fig-survey-share
#| fig-cap: "Compartment share of all 16 main MitoCarta groups across the four groups (% of the non-mtDNA transcriptome, free y). Genotype is the clean axis; the time axis is batch-confounded. n=6/group."
#| fig-width: 9
#| fig-height: 8
#| echo: false
source(here::here("figures", "figS1_mitocarta_survey_share.R"))
print(p)
```

```{r}
#| label: fig-survey-lfc
#| fig-cap: "Per-gene raw log2FC across the four fig01 contrasts for the 16 groups. A: median LFC per group x contrast (n = genes at padj<0.05; -n = mostly down). B: the LFC distributions. Read the WT 6->12W column."
#| fig-width: 9
#| fig-height: 8.5
#| echo: false
source(here::here("figures", "figS1b_mitocarta_survey_lfc.R"))
print(p)
```

The answer: in the WT timeline **nuclear OXPHOS is the outlier** -- only ~38% of its genes up
(6 individually significant *down*) -- while the biosynthetic and substrate-metabolism groups
*rise* (sulfur 80%, carbohydrate 77%, amino-acid 76%, lipid 68% of genes up; metals &
cofactors, detoxification, small-molecule transport and electron carriers all >55%). The flat
total is therefore a **reallocation**, not stasis: as the developmental nuclear-OXPHOS
programme winds down over the timeline, substrate-metabolism groups take up the share. Note
that **mtDNA OXPHOS moves opposite to nuclear OXPHOS** -- it rises over the timeline (@fig-survey-lfc
panel A) -- the same nuclear-down / mtDNA-up mitonuclear discordance discussed in @sec-lesner,
though here on the batch-confounded time axis and so described, not claimed.

Two guards on reading this. (1) The WT-timeline columns sit on the batch-confounded axis
(batch = timepoint), so the rise is **described, not claimed** -- the clean read is the
genotype columns, where Myc induces essentially *every* group at 6W and that induction fades
by 12W (attenuation, @sec-attenuation). (2) The median-LFC fills and the significant-gene
counts track set size and power; the direction (% up) is the power-independent read.

The companion table lists, per group, the genes whose LFC varies most across the four
contrasts (`spread = max - min`) -- the context-dependent movers behind each group's summary
(full CSV at `outputs/figures/figS1b_diverse_genes.csv`).

```{r}
#| label: tbl-diverse-genes
#| tbl-cap: "Most divergent genes per MitoCarta group (top by LFC spread across the four contrasts)."
#| echo: false
knitr::kable(diverse_genes, digits = 2,
             col.names = c("group", "gene", "Myc@6W", "Myc@12W", "WT 6->12W",
                           "Myc+ 6->12W", "spread", "sd"))
```

# OXPHOS, complex by complex {#sec-oxphos-complexes}

The survey isolates **nuclear OXPHOS** as the arm that rises with Myc and falls across the
WT timeline. Does one respiratory complex carry that, or all five together? This drills into
the five nuclear-encoded complexes (Complex I-V; the 13 mtDNA subunits are the separate mtDNA
OXPHOS facet) the same two ways -- share (@fig-cplx-share) and per-gene LFC (@fig-cplx-lfc).

```{r}
#| label: fig-cplx-share
#| fig-cap: "Transcriptome share of each respiratory complex across the four groups (% of the non-mtDNA transcriptome, free y). Brackets = genotype/ time p<0.05 only. n=6/group."
#| fig-width: 9
#| fig-height: 4.2
#| echo: false
source(here::here("figures", "figS2_oxphos_complex_share.R"))
print(p)
```

```{r}
#| label: fig-cplx-lfc
#| fig-cap: "Per-gene raw log2FC of the five complexes across the four contrasts. A: median LFC per complex x contrast (n = genes at padj<0.05). B: LFC distributions."
#| fig-width: 9
#| fig-height: 5.8
#| echo: false
source(here::here("figures", "figS2b_oxphos_complex_lfc.R"))
print(p)
```

The reading is **uniformity, not selectivity**: Myc raises all five complexes coordinately at
6W (the share is genotype-significant for CI, CIII, CIV and CV; CII -- 8 nuclear genes, the one
complex with no mtDNA subunit -- is the underpowered exception), and every complex attenuates in
step by 12W. No complex is singled out; the oncogene drives the respiratory chain as one block,
consistent with the coordinate biogenesis reading of @sec-content.

::: {.callout-note}
## Mapping note (why Complex V needed care)
MitoCarta's complex sets carry the **old ATP-synthase gene names** (`Atp5a1/b/c1/...`), while
the count matrix uses the **current** ones (`Atp5f1a..`, `Atp5mc*`, `Atp5pd/pf/po`). A
current-symbol match silently drops 17 of Complex V's 24 genes. These figures therefore resolve
membership through MitoCarta's **own Ensembl IDs** (Sheet 2) rather than by symbol, recovering
CV 24/24. They are computed self-contained (not via script 32), so fig01/figS1 are untouched.
:::

```{r}
#| label: tbl-cplx-diverse
#| tbl-cap: "Most divergent genes per complex (top by LFC spread across the four contrasts)."
#| echo: false
knitr::kable(diverse_genes, digits = 2,
             col.names = c("complex", "gene", "Myc@6W", "Myc@12W", "WT 6->12W",
                           "Myc+ 6->12W", "spread", "sd"))
```

# Myc reprioritises the compartment {#sec-reallocation}

The content figures use one ruler -- absolute abundance. mitoPPS (Monzel 2025) adds the
second: it is **content-blind**, reading *relative* resource reallocation *within* the
compartment (each MitoPathway is ratio-normalized to 1; the pairwise ratio cancels the
per-sample common mode, mitoPPS ambient ~0.42 vs GSVA's ~0.80). The main point is simply that
**reprioritisation happens** -- Myc does not merely raise the compartment, it reshuffles
priority within it, coherently by pathway tier and at both ages (@fig-reallocation).

```{r}
#| label: fig-reallocation
#| fig-cap: "MitoPathways ranked by their Myc mitoPPS effect (Myc+ - WT, 6W), promoted (left) to demoted (right); one dot per pathway coloured by tier, * = padj<0.05, the 7 top-level categories labelled, and 12W shown as a small point joined to 6W (the fade). Reprioritisation is structured by function. n=6/group."
#| fig-width: 9
#| fig-height: 4.7
#| echo: false
source(here::here("figures", "fig02_reallocation_ranked.R"))
print(p)
```

The ranking is **structured by pathway function**, not random: the **demoted** end is dominated
by dynamics & surveillance, apoptosis and signalling (and the mtDNA-encoded OXPHOS arm at the
very bottom), while the **promoted** end gathers biosynthetic metabolism and the import
machinery (SAM, chaperones). The seven top-level categories (labelled) span the whole range --
dynamics most demoted, import/central-dogma most promoted. **OXPHOS is visibly *split***: its
structural subunits sit high (promoted) and its assembly factors low (demoted), which is why
OXPHOS lands ~neutral as a whole tier. This is the **clean genotype axis**, but **descriptive /
exploratory** (n=6/group): the claim is the ordered, tier-coherent pattern, not any single pathway.

The fully-labelled ranking -- every pathway named, with 6W and 12W dots and their connectors --
is @fig-reallocation-full.

```{r}
#| label: fig-reallocation-full
#| fig-cap: "Supplementary (figS4): every MitoPathway (~144) ranked by Myc mitoPPS effect, fully labelled; filled = 6W, open = 12W, line = the 6W->12W change. The full version of @fig-reallocation. n=6/group."
#| fig-width: 8.5
#| fig-height: 12
#| echo: false
source(here::here("figures", "figS4_reallocation_full.R"))
print(p)
```

The pathway-level detail behind this -- and the one place the two rulers *disagree* -- is in
@fig-reallocation-detail: across all ~148 pathways, absolute content (DESeq set log2FC) is
positive almost everywhere, yet within **OXPHOS** the structural *subunits* are promoted in
priority (CIII-subunits mitoPPS +0.20, CV-subunits +0.17) while the *assembly factors* are
demoted (CII −0.26, CIII −0.14 padj 0.022, respirasome −0.11) and the mtDNA arm falls (−0.34) --
all on positive content. That subunit-over-assembly split is why OXPHOS reads ~neutral at tier
level, and it only surfaces at all-pathways granularity.

```{r}
#| label: fig-reallocation-detail
#| fig-cap: "Supplementary: content (absolute set log2FC) vs priority (relative mitoPPS) Myc effect across all ~148 MitoPathways, tier-blocked. The colour flip -- red on the left, blue on the right -- is the reallocation, sharpest within OXPHOS (subunits vs assembly). n=6/group."
#| fig-width: 8.5
#| fig-height: 13
#| echo: false
source(here::here("figures", "figS3_reallocation_two_rulers.R"))
print(p)
```

Collapsing the two rulers into one scatter with marginal distributions
(@fig-reallocation-scatter) makes the logic plain. **Content is one-sided**: ~94% of pathways
rise (the top marginal is a single hump right of zero) -- Myc raises nearly the whole
compartment. **Priority is wide**: it straddles zero, only ~56% up (the right marginal). That
wide priority spread, sitting on a near-uniform content increase, *is* the reprioritisation.
Two features stand out. (1) The reallocation is **directional**: the content-down / priority-up
quadrant is **empty** -- nothing that falls in content gains priority, so the demoted pathways
are not switched off, they simply rise *slower than the biogenesis machinery* and lose relative
share. (2) The priority extremes name candidates: **most promoted** are the import chaperones and
SAM, the immune response and anaplerotic metabolism (creatine, pyruvate); **most demoted (raised
yet)** are the OXPHOS assembly factors, cAMP-PKA signalling, the dynamics/apoptosis pathways and
the mtDNA-encoded arm. No trend line is drawn: mitoPPS is a ratio (~content relative to the
compartment mean), so a content/priority slope is partly built in. Descriptive/exploratory
(n=6/group).

```{r}
#| label: fig-reallocation-scatter
#| fig-cap: "Supplementary (figS5): priority (relative mitoPPS effect, x) vs content (set-average log2FC, y) per pathway at 6W, with marginal densities; the 7 top-level categories labelled. Content is one-sided-positive (~94% up), priority wide both ways (~56% up) -- that spread is the reallocation. Ringed = padj<0.05. n=6/group."
#| fig-width: 9
#| fig-height: 6.6
#| echo: false
source(here::here("figures", "figS5_reallocation_scatter.R"))
print(p)
```

## Reallocation over the timeline {#sec-reallocation-time}

The reallocation above is the Myc *genotype* effect within a timepoint. A **separate**
phenomenon plays out over the timeline: the **WT background itself reprioritises** as the tissue
develops. Ranking the pathways by their WT 6->12W mitoPPS change (@fig-reallocation-time) shows a
full developmental reallocation in the WT alone -- and the dropline to the Myc+ 6->12W change
shows where Myc **bends** that trajectory. The two are largely shared (WT-time and Myc+-time
mitoPPS correlate r=0.75, 70% same direction, @fig-reallocation-time-div): most of the temporal
reallocation is developmental, common to both genotypes. Where Myc departs is the **fade** -- the
biogenesis / biosynthesis and OXPHOS arms it promoted at 6W lose relative priority hardest by 12W
in Myc+ (they sit *below* the y=x line), the mitoPPS reading of the attenuation (@sec-attenuation).
The apoptosis / mitophagy pathways move with the developmental background rather than diverging --
i.e. no Myc-specific temporal priming signal in this lens.

::: {.callout-warning}
The **time axis is batch-confounded** (batch = timepoint), so the individual WT-time and Myc+-time
changes are **descriptive, not claims**. The clean quantity is their **divergence** -- the dropline
in @fig-reallocation-time, the deviation from y=x in @fig-reallocation-time-div -- because the
batch effect cancels in the difference. Both are exploratory (n=6/group).
:::

```{r}
#| label: fig-reallocation-time
#| fig-cap: "Supplementary (figS6): MitoPathways ranked by WT 6->12W mitoPPS change (filled), with a dropline to Myc+ 6->12W (open) -- the Myc-specific bend. The 7 top-level categories labelled. n=6/group."
#| fig-width: 9
#| fig-height: 5
#| echo: false
source(here::here("figures", "figS6_reallocation_temporal_ranked.R"))
print(p)
```

```{r}
#| label: fig-reallocation-time-div
#| fig-cap: "Supplementary (figS7): WT 6->12W vs Myc+ 6->12W mitoPPS change per pathway. On the y=x line = shared/developmental (r=0.75); off it = where Myc bends the WT trajectory; below-diagonal = the biogenesis arms Myc fades. n=6/group."
#| fig-width: 6.5
#| fig-height: 6.8
#| echo: false
source(here::here("figures", "figS7_reallocation_temporal_divergence.R"))
print(p)
```

# The global axis and the correlation ceiling {#sec-axis}

*(stub)* Why cross-sectional couplings are unreliable: the 884 pathway scores collapse onto
one global per-sample axis (`pathway_loading`: PC1=76% var, eff. dim 1.7; OXPHOS r=0.968 with
the global mean). A mito axis's median coupling to an *arbitrary* library set is ~0.80.
"OXPHOS is central" is therefore untestable (no separable variance); mito is mito-LED, not
mito-SPECIFIC. Panels: `pathway_loading/A_scree`, `A3_gene_axis_covariates`, `linear_pathway_coupling/A_ceiling_origin`.

# The oncogenic programme attenuates across progression {#sec-attenuation}

Foreshadowed already in fig01b (@sec-content, tension 2): the per-gene induction shrinks
6W->12W (OXPHOS still 87% up at 12W but only 1 gene clears FDR), and the Time-Myc+ contrast
is net **down** (Nuclear MitoCarta 68 down vs 20 up) -- the fade, in the same panel. The
question this section settles is *why*: does the Myc effect fade because the **background**
the oncogene acts on has moved, or because the **programme itself** decays? @fig-background
answers it in four panels, all built on one exact identity -- the change in the genotype gap
equals (Myc+ temporal change) - (WT temporal change), which holds to machine precision on
both the content and the priority ruler.

**First, the trivial explanation: the driver does not fade** (@fig-myc-dose). Myc's own
genotype gap *widens* over the window, **+1.64 log2 at 6W (p=2.5e-4) to +1.80 at 12W
(p=9.6e-6)**, and no within-genotype decline could carry the fade (Myc+ 8142 -> 7578
normalised counts, p=0.65). The WT baseline itself drifts down borderline (2543 -> 2143,
p=0.052), but that sits on the batch-confounded axis -- which is precisely why the gap, the
batch-clean quantity, is the one read. The proximal MYC/MAX/MXD network is flat: across MAX,
the MXD1/3/4-MXI1-MNT-MGA repressors and the MLX arm, 40 simple-effect tests give 5 at nominal
p<0.05 and **none survive BH** (min 0.052), with no genome-wide interaction padj<0.05 for any
of them -- no repressor rises to titrate MAX away.

The contrast that settles it is driver against output. **Every** measure that would register a
falling dose widens over the window -- Myc **+0.17**, Myc:MAX **+0.32** (+1.42 -> +1.74),
Myc:repressor-sum **+0.25** (+1.65 -> +1.90) -- while Myc's *output*, the genotype gap in
Hallmark MYC targets V2, **narrows by -0.36** (+1.32 -> +0.95). Driver up, output down: the
attenuation is **downstream of dose**, and the rest of this section is the decomposition of
that downstream part. Two bounds on the claim. Reads map to the mouse *Myc* locus, so total
message in Myc+ is transgene plus endogenous and the two are not separable here (script 27
decomposes them at programme level) -- "constant dose" is a statement about total message,
which is what the substrate sees. And transcript is not protein: the 31-gene post-translational
stability panel (`outputs/myc_stability_panel/`, script 41) finds nothing Myc-specific (every
interaction padj>0.05), no excess movement over an expression-matched background, and
directions that predict a *more* stable MYC (Fbxw7 -0.35 and Huwe1 -0.24, the two established
degradative E3 ligases, both down) -- but that is absence of transcriptional evidence, not evidence of
absence. The settling experiments are protein-level: a MYC blot at 6W vs 12W, phospho-T58/S62,
and a cycloheximide chase.

```{r}
#| label: fig-myc-dose
#| fig-cap: "figS8: is the Myc dose constant? (A) Myc normalised counts across the four groups. (B) The proximal MYC/MAX/MXD network -- MAX, the MXD/MNT/MGA repressors and the MLX arm; brackets unlabelled, red = p<0.05. (C) Everything that would register a falling dose (Myc, Myc:MAX, Myc:repressor sum) against Myc's output (Hallmark MYC targets V2), as the genotype gap at each age; open = 6W, arrow to 12W. The driver gap widens while the output gap narrows -- the attenuation is downstream of dose. n=6/group."
#| fig-width: 9
#| fig-height: 7.4
#| echo: false
source(here::here("figures", "figS8_myc_network_levels.R"))
print(p)
```

**Myc's effect is rescaled, not reshaped.** Regressing the 12W Myc effect on the 6W effect
across the 144 MitoPathways gives **slope 0.55** (bootstrap 0.51-0.60), R2 0.80, rho 0.84 on
the content ruler and slope 0.64, R2 0.79, rho 0.93 on priority, with per-tier slopes
0.53-0.83 -- uniform (@fig-background B). The coherence sits at the **100th percentile** of
gene-label shuffles matched on set size, expression and pathway overlap. The 12W profile *is*
the 6W profile at 55% of the amplitude: the same pathways, in the same order, less of it.
This is the pathway-level form of the Issue #4 result that magnitude, not rank, shrinks.

**What Myc adds over the timeline is a uniform offset, not a different set of pathways.**
Regressing the Myc+ 6->12W change on the WT 6->12W change (@fig-background C) gives slope
0.75 and intercept -0.17. The **slope** -- the part both genotypes share -- is not
distinguishable from the null (null median 0.63, 82nd percentile). The **intercept** is, at
the 0th percentile. It is also the only batch-clean quantity in the comparison, since an
offset common to both temporal contrasts cancels in their difference.

**The background is not moving toward the Myc state.** The WT 6->12W vector projects onto the
Myc(6W) vector at just **0.064** -- against a null median of 0.24, the 9.8th percentile, i.e.
*less* than an arbitrary matched panel. Restricted to the biogenesis and OXPHOS arms the cos
angle is **-0.34** (content) / -0.39 (priority): the background moves *away* from where Myc
pushes. That aggregate is not homogeneous, and the per-tier breakdown is informative -- protein
import (**-0.63** content, **-0.77** priority) and OXPHOS (**-0.65**) move against Myc, while
mitochondrial central dogma (**+0.74**) and metabolism (+0.48) move with it. The Myc+
trajectory, by contrast, has cos angle **-0.84** with the Myc effect (0th percentile), and
-0.80 to -0.98 in every substantial tier: over the timeline the Myc+ compartment retraces the
Myc programme backwards. @fig-background A shows the same thing per tier -- a red Myc arrow
that keeps its direction and shortens, and a grey background arrow that points elsewhere.

**And the WT-convergence term does not survive its control.** The published decomposition
aligns each gene to Myc's induction direction and reads the WT temporal change along it -- but
the genotype contrast and the WT temporal contrast **share the 6W WT baseline with opposite
signs**, so noise in that baseline manufactures apparent convergence. The six 6W_neg mice are
therefore split over all 20 three-versus-three partitions -- and, because halving the baseline
also adds noise to the alignment direction (which pulls the statistic toward chance from either
side), each partition is run twice off the *same* half baseline, once shared between the two
contrasts and once split. Identical direction, identical universe, identical noise; the only
difference is the sharing.

Globally, sharing the baseline adds **+0.24** (sd 0.138) of apparent convergence, in 19 of the
20 partitions: `frac_wt_toward` falls from 0.720 to **0.478** -- chance -- and the percentage
from +61% to -12% (@fig-background D). And the artifact is **absent exactly where the published
analysis reported divergence**: +0.01 on the MYC-target core, -0.01 on OXPHOS subunits, so
their divergence stands as published. What survives on the other side is programme-specific:
the biosynthetic arms genuinely converge (amino acid 0.74 -> **0.65**, lipid 0.77 -> **0.63**,
TCA and nucleotide ~0.63-0.64). The programme-specificity was the whole result; the global
number was the artifact.

So the picture is **additive, not causal**: a reallocation of the mitochondrial compartment
that both genotypes undergo, plus a Myc programme that decays proportionally without changing
shape. The background's own move is 57-62% as large as the Myc effect and points in a
direction the Myc effect does not.

One number the section deliberately does not give is a single size for the attenuation. On
the same gene sets, the share ruler retains 77-83% of the genotype gap at 12W while the
set-average LFC rulers retain 54-63% -- the sign and direction agree, the magnitude is
normalisation-dependent (denominator: per-sample transcriptome sum vs median-of-ratios size
factors). The manuscript quotes the range.

::: {.callout-warning}
**Batch = timepoint.** Both temporal contrasts carry the same batch offset. It cancels in the
interaction -- @fig-background C's intercept, and the whole rescaling result -- but **not** in
the convergence/fade split, so calling the shared temporal vector "developmental" rather than
"batch" remains an interpretation. The sample-split removes the shared-*baseline* artifact; it
cannot remove the batch confound. Separately, a family of tempting statistics here is
**algebraically forced** and carries no information: cor(WT-time, change in Myc effect) is
-0.19 (content) / -0.33 (priority) and the value the marginals force is *exactly* the same,
because the change is the difference of the two temporal contrasts. The full list is in
`results/background_vs_myc.rds$artifact_ledger`. None of them may be cited.
:::

```{r}
#| label: fig-background
#| fig-cap: "fig03: is Myc's mitochondrial effect shaped by a changing background? (A) Each Level-1 MitoPathway tier as four states in priority x content; red arrows = the Myc effect at 6W (solid) and 12W (dashed), grey = each genotype's temporal move. (B) The 12W Myc effect is the 6W effect x0.55 -- rescaled, not reshaped. (C) Myc+ vs WT temporal change: the shared slope is not beyond the null, the Myc-specific intercept is. (D) WT-convergence on the matched control -- the same three-mouse 6W WT baseline shared between the two contrasts (open) vs split (filled), 20 partitions; tick = the published six-mouse value. The global term collapses to chance; the biosynthetic arms survive and OXPHOS/MYC-core divergence is untouched. n=6/group, exploratory."
#| fig-width: 9
#| fig-height: 8.6
#| echo: false
source(here::here("figures", "fig03_background_vs_myc.R"))
print(p)
```

Full numbers, nulls and the artifact ledger: `docs/2026-07-24_background_vs_myc_interpretation.md`
(script 40). The earlier per-gene decomposition (script 31) is unchanged on disk; script 40
tests it and reports the correction.

# The trade-off: what Myc's mitochondria cost the cell {#sec-death}

*(stub -- DO NOT write this section from the text below. It is a scope note, and the
argument order is still under discussion.)*

**The framing here changed completely on 2026-07-25/26 and this heading is the only part of
the old stub that was worth keeping in place.** This was "Death priming: what did not
survive", a negative-results section. It is now the positive spine of the paper. Write it
from, in this order:

- `docs/2026-07-25_death_narrative_dose_vs_competence.md` -- section 0 (the epistemic
  contract: the in-vivo transcriptome GENERATES, the cell perturbations PROVE), section 3
  (Myc builds a death-ready mitochondrion and loses the trigger: Bax:Bcl-xL retains 0.550,
  the global rate, while PUMA:Bcl-xL retains -0.090), section 5 (the six negatives),
  section 6.5 (the coincidence test), section 9 (argument order).
- `docs/2026-07-26_introduction_alignment_and_the_question.md` -- section 0 (the trade-off
  and the three solutions), 0.5a (the asymmetry, scoped to the ONE outcome with a working
  control), 0.7 (the gland de-respires without de-proliferating).

**What survives from the old stub, and must appear here as the bounds** -- these are not
optional, they are the reason the section is defensible: the GSVA-space mito->death bridge is
**circular** (priming built from `MITOCARTA_APOPTOSIS_PRO`/`_ANTI`, so it correlated mito
genes with mito genes) and fails every null; "Myc raises priming" as a **composite** is the
mito-content effect, below average for a mitochondrial gene; and the attenuation is
**global**, so a composite priming signal fades because the Myc programme fades. The finding
is gene-level, not composite-level -- which is why the composite analyses found nothing and
is exactly what the cell work predicts.

**Added 2026-07-27, and the argument order is now decided: SUBSTRATE FIRST.** Write from the
nine-step order in section 9 of the death narrative (rewritten) -- what the gland does to
itself, then what Myc does on top of it. Four things from script 44 (author-run) belong here:

- `Bbc3` sits at the **0.51st percentile of 8774 Myc-responsive genes** on a genome-wide
  collapse statistic that never touches the Bcl-xL denominator, pre-specified from the PGC1a
  westerns rather than found by scanning -- an independent reproduction of the priming result.
- **`Bcl2l11` (BIM) sits at the 91.6th percentile**, so the westerns' PUMA+BIM pair does
  **not** transfer as a pair; the in-vivo collapse is PUMA-specific. Report both.
- **PUMA is a solo.** No pre-registered question returns a module -- not OXPHOS, not
  apoptosis on any lens, not AP_TEB, not ISR. What loses Myc-inducibility genome-wide is
  **luminal/hormone-sensing lineage identity** (`MG_HS_GRAY` NES -2.79, padj 1.7e-17) and
  what keeps it is **the MYC programme itself** (all MYC signatures NES > +2.5), which
  independently reproduces Issue #6's "MYC-core protected".
- **An unplanned internal control worth one sentence:** `Myc` itself is at the 100th
  percentile, retention 1.09 (+1.679 -> +1.822). The transgene message does not attenuate,
  so the x0.55 fade downstream cannot be transcript-level dose -- the blot's premise
  confirmed from inside the RNA.

Sources on disk: `results/priming_arm_teb.rds` (42), `results/substrate_specificity_tradeoff.rds`
(43), `results/death_priming_reassessment.rds` (34, for the bounds above),
`results/collapse_module_ownership.rds` (44).
Panels: unbuilt on purpose -- fig04 follows the agreed argument order. Substrate-first
promotes **script 43 PART A to an early main figure**.

::: {.callout-note title="Planned experiment: MYAZ + PGC1a RNA-seq"}
Not yet done; the author intends to do it. It is **the only design that can define the
OXPHOS/apoptosis module by perturbation** rather than by correlation -- whatever PGC1a
co-induces that is both mitochondrial and apoptotic *is* the module, with no null to fight
and no correlation ceiling. In this dataset the module cannot be defined by co-expression:
the wild-type gland does not move its apoptotic transcripts at all, the PGC1a-axis regulon is
*depleted* of pro-apoptotic genes (fold 0.76) while MYC's is enriched (2.68x), and the
per-sample couplings sit at the ambient. See section 0.8 of
`docs/2026-07-26_introduction_alignment_and_the_question.md`. When the data exists it should
settle: whether the payload/trigger division of labour holds under direct PGC1a manipulation,
and why PUMA and BIM behave as a pair in cells but split in vivo.
:::

# External validation: Lesner et al. {#sec-lesner}

*(stub)* The MYC nuclear-OXPHOS-up / mtDNA-down discordance is independently established
(Lesner et al., bioRxiv 2026.07.13.738248) in TCGA, mouse HCC, proteomics, mtDNA qPCR --
killing the "our enzymatic-dissociation artifact" worry. Lesner validates the *question*; our
mtDNA-down arm is genotype-independent (p~0.49), i.e. underpowered, not demonstrated. Their
hypoxia->MXI1/JUN Myc-suppression route does NOT transfer to our model (script 39; liver-specific).
Panels: `lesner/A_mechanism_genes`, `lesner/B_fgsea_contrasts`, `lesner/C_mitophagy_specificity_null`.

# Leads for the bench {#sec-leads}

*(stub)* Hypothesis-generating, not results (n=24): `redox -> MB2 tumorigenic fork` (the one
robust Myc-independent coupling; test: redox/GSH assay) and `nuclear OXPHOS -> death-priming`
(test: **BH3 profiling** -- priming is an occupancy state at the outer membrane, and no
amount of transcriptomics substitutes for measuring it).

**`Foxo3`, new 2026-07-27 and the one live mechanistic lead** -- it collapses like PUMA
(0.87th percentile of its matched stratum; +0.243 at 6W -> -0.242 at 12W, interaction
p 0.0074, against `Bbc3`'s +0.258 -> -0.283, p 0.0081) and it is PUMA's canonical
p53-independent activator. Test: **FOXO3 western across the two ages, and in the PGC1a
arm.** Caveats that must travel with it: both 6W effects are non-significant, both
interactions are genome-wide null, two genes are not a module, and `Foxo3` *rises* in the
wild-type gland. Note also that our direction is **opposite** to the canonical
PGC-1a-sequesters-FOXO3a axis in `docs/library_reference/PUMA-and-its-relationships.md`. Not
for the Results on the RNA alone.

**Closed, do not propose it again:** the retrograde/ISR route to PUMA is tested and negative
(script 44 PART D -- the ATF4 target set and Hallmark UPR are null in the collapse ranking,
and `Chac1`, which looked collapsed, has no 6W effect to lose at p 0.47).

The OXPHOS->priming lead is **gene-level, not composite-level** (corrected 2026-07-27, it was
described here in its pre-script-43 composite form). The composite version is dead twice
over: script 34 showed it is circular, and script 43 PART B showed that of five outcomes only
the **gene-level `Bbc3:Bcl2l1` ratio** has a `redox` control that behaves (`myc x OXPHOS-priority`
+6.09, p 0.005, 91.7th percentile against a control at the 51.2nd). The claim is a **sign
contrast on one outcome**, not a magnitude comparison across five. Two further tests the
mechanism menu (section 0.9 of the alignment doc) names, neither answerable from RNA:
cardiolipin by lipidomics and cristae architecture by EM.

# Provenance {#sec-provenance}

```{r}
#| label: session
#| echo: false
sessionInfo()
```
