---
title: "Mitochondria integrate oncogenic and developmental signals to time Myc-driven mammary tumourigenesis"
subtitle: "MMTV-Myc mammary timecourse -- exploratory synthesis and figure narrative"
author: "Szabadkai lab"
date: today
---

::: {.callout-note}
**Skeleton, not the finished document.** This is the container the reflective write-up
will grow into as the publication panels are finalised. It mirrors the exploratory arc of
scripts 32-39 and reuses the same `results/*.rds` objects and the `figures/` panel layer.
Each section below is a stub: prose to be written, panels to be embedded. The organising
principle, settled 2026-07: **genotype CONTRASTS survived; cross-sectional COUPLINGS sit
at a correlation ceiling and mostly did not.**
:::

```{r}
#| label: setup
#| include: false
library(here)
library(ggplot2)

# Pin here() to the repo root: this .qmd lives under paper/, and a knitr render
# starts the session there, so anchor explicitly (sourced panel scripts inherit it).
here::i_am("paper/myc_mito.qmd")

# Display panels inline without re-exporting their canonical PDFs to outputs/.
options(myc.fig.nosave = TRUE)

# Read-only load of the analysis objects. The figure layer never re-runs analysis;
# it reads what scripts 00-39 already wrote to results/.
res <- function(x) readRDS(here::here("results", paste0(x, ".rds")))

# Loaded lazily per section as the prose is written; listed here as the manifest:
#   content   <- res("mito_content_proxies")        # script 32 -- Fig content
#   mtaxis    <- res("mtdna_axis_and_coupling_null") # script 33 -- adjusted +21%
#   loading   <- res("pathway_loading")              # script 37 -- global axis / dim
#   coupling  <- res("linear_pathway_coupling")      # script 36 -- ceiling origin
#   lesner    <- res("lesner_mechanism_test")        # script 39 -- external validation
```

::: {.callout-tip collapse="true"}
## Figure manifest -- build number vs manuscript figure

**The bioinformatics is TWO main figures of up to six panels plus two supplementary**
(author, 2026-07-27), and they follow the order of the Results paragraph: what MYC builds,
then what development takes away and what that hands to the cell experiments.

| paper figure | script | panels | source analysis | section |
|--------------|--------|--------|-----------------|---------|
| **Figure 1** | `figure1_myc_mitochondrion.R` | A-D | 32, 08, 42 | @sec-content |
| **Figure 2** | `figure2_developmental_window.R` | A-F | 43, 44, 42 | @sec-window |
| Supp 1 | `figureS1_compartment_detail.R` | A-C | 32, 08 | @sec-mito-survey |
| Supp 2 | `figureS2_controls.R` | A-D | 40, 41, 44 | @sec-attenuation |

Everything else under `figures/` -- `fig01`-`fig05`, `figS1`-`figS8` -- stays as the
**exploratory layer** this document walks through. Those scripts are not renamed: their
numbers are what the commits, the dated docs and `outputs/figures/` cite.

:::

# Overview {#sec-overview}

*(stub)* One-paragraph orientation: bulk RNA-seq of enzymatically purified MMTV-Myc MECs,
four groups x six replicates, `~timepoint*myc_status` interaction design. Batch = timepoint,
so genotype/interaction contrasts are clean and the pure developmental reading is confounded.
What this document does: walk the exploratory process and land the defensible results.

# MYC builds a death-competent mitochondrion {#sec-content}

**MANUSCRIPT FIGURE 1.** At constant dose (@fig-myc-dose), MYC raises the **nuclear**
mitochondrial arms -- mass markers, nuclear MitoCarta and nuclear OXPHOS -- while the 13
mtDNA-encoded transcripts, the reference arm, stay flat: a content increase of ~**+21%**
(p=0.0006) after adjusting for prep-stress and contamination, and a *lower bound*, because MYC
amplifies total RNA per cell (@fig-myc-mito A). The biogenesis arm behaves the same way and is
the exploratory detail below.

**But the compartment is reprioritised, not simply enlarged.** Ranking every MitoPathway by
MYC's effect on mitoPPS -- a within-compartment ratio, blind to how many mitochondria a cell
has -- separates promoted from demoted arms in an order structured by function
(@fig-myc-mito B). Plotting the two rulers against each other shows what that means: the
content-down / priority-up quadrant is **empty** (@fig-myc-mito C). Demoted arms are not
switched off; they rise more slowly than the rest and lose share. Reprioritisation here is a
skew on a broad rise, not a redistribution of a fixed budget.

**And the organelle MYC builds can execute its own death.** The same genotype contrast raises
`Htra2` (**+1.49**, padj 2.6e-08) and `Bax` (**+0.48**, padj 1.4e-04) alongside the import,
cristae and OXPHOS arms, while lowering `Bcl2l1`/BCL-xL (**-0.42**, padj 0.040)
(@fig-myc-mito D). Asset and liability, built in one move -- the trade-off written as
molecules rather than as a model.

```{r}
#| label: fig-myc-mito
#| fig-cap: "MANUSCRIPT FIGURE 1 (figure1). MYC builds a death-competent mitochondrion. (A) Compartment share across the four groups for three nuclear arms, with the 13 mtDNA-encoded transcripts as the reference arm that does not move; brackets are the genotype comparison at each age, which is the batch-clean axis. (B) All MitoPathways ranked by MYC's effect on mitoPPS. (C) Content against priority, one point per pathway; the content-down / priority-up quadrant is empty. (D) The curated death-and-mitochondrion roster on the same genotype contrast. Genotype contrast throughout, n=6/group, raw log2FC."
#| fig-width: 9
#| fig-height: 7.7
#| echo: false
source(here::here("figures", "figure1_myc_mitochondrion.R"))
print(p)
```

The rest of this section is the exploratory detail behind panel A.

The biogenesis arm is **MitoCarta central dogma + protein import/sorting/homeostasis** (314
genes: mtDNA replication and transcription, mtRNA metabolism, mitochondrial translation and
ribosome, import, chaperones, proteases). It replaces the earlier mitoribosome facet for two
reasons: only **7 of its 314 genes** are in nuclear OXPHOS, so *OXPHOS vs general biogenesis*
is a genuinely disjoint comparison rather than a whole-vs-part one; and it is the same
`mito_biogenesis` axis the coupling analyses use (script 28), so figure and analysis name one
thing. Deliberately **not** the `Biogenesis_discrimination` panel (`CORE_MITO`, `MYC_MITO`,
`DEVELOPMENTAL_MITO`): those are regulons, and `MYC_MITO` is Myc targets by construction --
circular on a genotype axis.

```{r}
#| label: fig-content
#| fig-cap: "Myc raises the nuclear mitochondrial arms; mtDNA-encoded output stays flat. Genotype is the clean axis; the time axis is cohort-confounded. n=6/group."
#| fig-width: 9
#| fig-height: 4.4
#| echo: false
# Reuse the panel script: it builds `p` and (as a side effect) writes the PDF to
# outputs/figures/. Here we just display the same object inline.
source(here::here("figures", "fig01_mito_content.R"))
print(p)
```

The compositional critique of the share panel -- "the fraction can rise because *other*
genes fall" -- is answered by the same four comparisons in per-gene log2FC space
(`figures/fig01b_mito_content_lfc.R`, from `interaction_results.rds`): the arm genes rise
in **absolute** terms and **coordinately** (Myc@6W: 83-95% of each nuclear arm's genes up,
biogenesis highest at 95%; mtDNA-encoded genes are *not* induced -- 0-15% up), so the share
rises because the genes rise, not by subtraction.

::: {.callout-important}
## Two cross-panel tensions the narrative must own (both real, not artifacts)

1. **6W: LFC says strong, share says marginal -- except on the biogenesis arm.** fig01b shows
   369 individually significant up-genes in Nuclear MitoCarta at 6W, yet fig01's *share*
   bracket at 6W is only p=0.11 there (mass markers 0.085, nuclear OXPHOS 0.12). Per-gene
   induction is large; the n=6 aggregate *fraction* is noisy at 6W, so the genes clearly move
   where the share is underpowered. **Mito biogenesis is the exception and the arm to lead
   with**: p=**0.035** at 6W and p=**0.0015** at 12W -- the only arm significant at both, and
   the same arm fig01b independently ranks most coordinate (94.7% of genes up at 6W). Share
   space and LFC space agree once the arm is a real programme rather than a sub-arm; part of
   the old tension was arm choice, not only sample size.

2. **12W: significant counts undersell the coordinate shift.** OXPHOS@12W has a clearly
   positive box (87% of genes up) but only **1** gene clears FDR, because attenuation shrinks
   per-gene magnitude and n=6 limits power at 12W -- exactly where fig01's share is *firmest*
   (p 0.003). The `% up` line on fig01b (power-independent) is what keeps 12W from looking
   artificially weak; read it alongside the significant-gene counts. The biogenesis arm makes
   the point explicit: at 12W it sits at essentially the same 88% up as OXPHOS yet retains
   **14** FDR-significant genes to OXPHOS's 1 -- the counts are tracking set size and power,
   not a difference in biology. This previews @sec-attenuation.

Integrated reading: **6W = strong per-gene induction, noisy aggregate; 12W = attenuated
per-gene signal, consolidated compartment share.**

Not a tension but the mirror of the claim: the only significant *time* bracket in fig01 is
mtDNA-encoded within Myc+ (p=0.0079, 6W->12W), matching fig01b's 85% up. It is on the
batch-confounded axis (batch = timepoint) and stays a described observation, not a result --
see @sec-lesner for why the mtDNA arm needs qPCR to settle.
:::

*(stub: the content narrative -- mass-panel coherence, where the imbalance is NOT
[the mtDNA machinery is not the transcriptional bottleneck], the three ceilings, and why
mtDNA qPCR / blot is the settling experiment. Possible future panel: the four contrasts as
the sides of a square [Myc@6W / Myc@12W / WT-time / Myc+-time] -- a 2x2 contrast diagram.)*

# The mitochondrial-group survey: what rises when OXPHOS falls {#sec-mito-survey}

**SUPPLEMENTARY FIGURE 1** is the three controls behind @sec-content, in the order the
claims are made: the compartment share rises because the *genes* rise and not by subtraction
(the compositional objection to a fraction); something has to rise while OXPHOS falls, and
the sixteen-group survey names it as the substrate-metabolism arms, so a flat compartment
total is reallocation rather than stasis; and the respiratory chain moves as one block, so
"MYC raises OXPHOS" is the right sentence rather than a claim about one complex.

```{r}
#| label: fig-compartment-detail
#| fig-cap: "SUPPLEMENTARY FIGURE 1 (figureS1). The compartment in detail. (A) Per-gene direction across the four contrasts for the Figure 1 arms; numbers inside the bars are FDR-significant genes, which track set size and power, while the % up line does not. (B) Median log2FC across the 16 MitoCarta groups; read the WT column. (C) The five respiratory complexes. Set membership resolved with the vintage reconciler, which recovers 17 of Complex V's 24 subunits that a plain symbol match drops. Genotype columns are the batch-clean axis; the temporal columns are confounded and described, not claimed."
#| fig-width: 9
#| fig-height: 8.2
#| echo: false
source(here::here("figures", "figureS1_compartment_detail.R"))
print(p)
```

The rest of this section is the exploratory version of panels B and C.

fig01 shows OXPHOS is the mitochondrial arm that *falls* across the WT 6->12W timeline,
yet the total nuclear MitoCarta share is flat -- so other groups must rise to hold it
there. This supplementary survey resolves which. It takes the seven canonical MitoCarta
MitoPathway top-level categories -- **with Metabolism split into its nine level-2 children**
(hierarchy read from MitoCarta Sheet 4) -- and reads them the same two ways as fig01/fig01b:
compartment share across the four groups (@fig-survey-share) and per-gene log2FC across the
four contrasts (@fig-survey-lfc). **OXPHOS is shown as its two halves** -- nuclear OXPHOS
(`MITOCARTA_OXPHOS_NU`, 155 genes, no mt-* genes) and mtDNA OXPHOS (the 13 mtDNA-encoded
subunits, which *are* `MITOCARTA_MTDNA_ENCODED`) -- so the nuclear category carries no mtDNA
bias and there is no redundant "mtDNA reference" facet. `share_nomt` (denominator drops only
the 13 mt-* genes) is the consistent scale.

```{r}
#| label: fig-survey-share
#| fig-cap: "Compartment share of all 16 main MitoCarta groups across the four groups (% of the non-mtDNA transcriptome, free y). Genotype is the clean axis; the time axis is batch-confounded. n=6/group."
#| fig-width: 9
#| fig-height: 8
#| echo: false
source(here::here("figures", "figS1_mitocarta_survey_share.R"))
print(p)
```

```{r}
#| label: fig-survey-lfc
#| fig-cap: "Per-gene raw log2FC across the four fig01 contrasts for the 16 groups. A: median LFC per group x contrast (n = genes at padj<0.05; -n = mostly down). B: the LFC distributions. Read the WT 6->12W column."
#| fig-width: 9
#| fig-height: 8.5
#| echo: false
source(here::here("figures", "figS1b_mitocarta_survey_lfc.R"))
print(p)
```

The answer: in the WT timeline **nuclear OXPHOS is the outlier** -- only ~38% of its genes up
(6 individually significant *down*) -- while the biosynthetic and substrate-metabolism groups
*rise* (sulfur 80%, carbohydrate 77%, amino-acid 76%, lipid 68% of genes up; metals &
cofactors, detoxification, small-molecule transport and electron carriers all >55%). The flat
total is therefore a **reallocation**, not stasis: as the developmental nuclear-OXPHOS
programme winds down over the timeline, substrate-metabolism groups take up the share. Note
that **mtDNA OXPHOS moves opposite to nuclear OXPHOS** -- it rises over the timeline (@fig-survey-lfc
panel A) -- the same nuclear-down / mtDNA-up mitonuclear discordance discussed in @sec-lesner,
though here on the batch-confounded time axis and so described, not claimed.

Two guards on reading this. (1) The WT-timeline columns sit on the batch-confounded axis
(batch = timepoint), so the rise is **described, not claimed** -- the clean read is the
genotype columns, where Myc induces essentially *every* group at 6W and that induction fades
by 12W (attenuation, @sec-attenuation). (2) The median-LFC fills and the significant-gene
counts track set size and power; the direction (% up) is the power-independent read.

The companion table lists, per group, the genes whose LFC varies most across the four
contrasts (`spread = max - min`) -- the context-dependent movers behind each group's summary
(full CSV at `outputs/figures/figS1b_diverse_genes.csv`).

```{r}
#| label: tbl-diverse-genes
#| tbl-cap: "Most divergent genes per MitoCarta group (top by LFC spread across the four contrasts)."
#| echo: false
knitr::kable(diverse_genes, digits = 2,
             col.names = c("group", "gene", "Myc@6W", "Myc@12W", "WT 6->12W",
                           "Myc+ 6->12W", "spread", "sd"))
```

# OXPHOS, complex by complex {#sec-oxphos-complexes}

The survey isolates **nuclear OXPHOS** as the arm that rises with Myc and falls across the
WT timeline. Does one respiratory complex carry that, or all five together? This drills into
the five nuclear-encoded complexes (Complex I-V; the 13 mtDNA subunits are the separate mtDNA
OXPHOS facet) the same two ways -- share (@fig-cplx-share) and per-gene LFC (@fig-cplx-lfc).

```{r}
#| label: fig-cplx-share
#| fig-cap: "Transcriptome share of each respiratory complex across the four groups (% of the non-mtDNA transcriptome, free y). Brackets = genotype/ time p<0.05 only. n=6/group."
#| fig-width: 9
#| fig-height: 4.2
#| echo: false
source(here::here("figures", "figS2_oxphos_complex_share.R"))
print(p)
```

```{r}
#| label: fig-cplx-lfc
#| fig-cap: "Per-gene raw log2FC of the five complexes across the four contrasts. A: median LFC per complex x contrast (n = genes at padj<0.05). B: LFC distributions."
#| fig-width: 9
#| fig-height: 5.8
#| echo: false
source(here::here("figures", "figS2b_oxphos_complex_lfc.R"))
print(p)
```

The reading is **uniformity, not selectivity**: Myc raises all five complexes coordinately at
6W (the share is genotype-significant for CI, CIII, CIV and CV; CII -- 8 nuclear genes, the one
complex with no mtDNA subunit -- is the underpowered exception), and every complex attenuates in
step by 12W. No complex is singled out; the oncogene drives the respiratory chain as one block,
consistent with the coordinate biogenesis reading of @sec-content.

::: {.callout-note}
## Mapping note (why Complex V needed care)
MitoCarta's complex sets carry the **old ATP-synthase gene names** (`Atp5a1/b/c1/...`), while
the count matrix uses the **current** ones (`Atp5f1a..`, `Atp5mc*`, `Atp5pd/pf/po`). A
current-symbol match silently drops 17 of Complex V's 24 genes. These figures therefore resolve
membership through MitoCarta's **own Ensembl IDs** (Sheet 2) rather than by symbol, recovering
CV 24/24. They are computed self-contained (not via script 32), so fig01/figS1 are untouched.
:::

```{r}
#| label: tbl-cplx-diverse
#| tbl-cap: "Most divergent genes per complex (top by LFC spread across the four contrasts)."
#| echo: false
knitr::kable(diverse_genes, digits = 2,
             col.names = c("complex", "gene", "Myc@6W", "Myc@12W", "WT 6->12W",
                           "Myc+ 6->12W", "spread", "sd"))
```

# Myc reprioritises the compartment {#sec-reallocation}

The content figures use one ruler -- absolute abundance. mitoPPS (Monzel 2025) adds the
second: it is **content-blind**, reading *relative* resource reallocation *within* the
compartment (each MitoPathway is ratio-normalized to 1; the pairwise ratio cancels the
per-sample common mode, mitoPPS ambient ~0.42 vs GSVA's ~0.80). The main point is simply that
**reprioritisation happens** -- Myc does not merely raise the compartment, it reshuffles
priority within it, coherently by pathway tier and at both ages (@fig-reallocation).

```{r}
#| label: fig-reallocation
#| fig-cap: "MitoPathways ranked by their Myc mitoPPS effect (Myc+ - WT, 6W), promoted (left) to demoted (right); one dot per pathway coloured by tier, * = padj<0.05, the 7 top-level categories labelled, and 12W shown as a small point joined to 6W (the fade). Reprioritisation is structured by function. n=6/group."
#| fig-width: 9
#| fig-height: 4.7
#| echo: false
source(here::here("figures", "fig02_reallocation_ranked.R"))
print(p)
```

The ranking is **structured by pathway function**, not random: the **demoted** end is dominated
by dynamics & surveillance, apoptosis and signalling (and the mtDNA-encoded OXPHOS arm at the
very bottom), while the **promoted** end gathers biosynthetic metabolism and the import
machinery (SAM, chaperones). The seven top-level categories (labelled) span the whole range --
dynamics most demoted, import/central-dogma most promoted. **OXPHOS is visibly *split***: its
structural subunits sit high (promoted) and its assembly factors low (demoted), which is why
OXPHOS lands ~neutral as a whole tier. This is the **clean genotype axis**, but **descriptive /
exploratory** (n=6/group): the claim is the ordered, tier-coherent pattern, not any single pathway.

The fully-labelled ranking -- every pathway named, with 6W and 12W dots and their connectors --
is @fig-reallocation-full.

```{r}
#| label: fig-reallocation-full
#| fig-cap: "Supplementary (figS4): every MitoPathway (~144) ranked by Myc mitoPPS effect, fully labelled; filled = 6W, open = 12W, line = the 6W->12W change. The full version of @fig-reallocation. n=6/group."
#| fig-width: 8.5
#| fig-height: 12
#| echo: false
source(here::here("figures", "figS4_reallocation_full.R"))
print(p)
```

The pathway-level detail behind this -- and the one place the two rulers *disagree* -- is in
@fig-reallocation-detail: across all ~148 pathways, absolute content (DESeq set log2FC) is
positive almost everywhere, yet within **OXPHOS** the structural *subunits* are promoted in
priority (CIII-subunits mitoPPS +0.20, CV-subunits +0.17) while the *assembly factors* are
demoted (CII −0.26, CIII −0.14 padj 0.022, respirasome −0.11) and the mtDNA arm falls (−0.34) --
all on positive content. That subunit-over-assembly split is why OXPHOS reads ~neutral at tier
level, and it only surfaces at all-pathways granularity.

```{r}
#| label: fig-reallocation-detail
#| fig-cap: "Supplementary: content (absolute set log2FC) vs priority (relative mitoPPS) Myc effect across all ~148 MitoPathways, tier-blocked. The colour flip -- red on the left, blue on the right -- is the reallocation, sharpest within OXPHOS (subunits vs assembly). n=6/group."
#| fig-width: 8.5
#| fig-height: 13
#| echo: false
source(here::here("figures", "figS3_reallocation_two_rulers.R"))
print(p)
```

Collapsing the two rulers into one scatter with marginal distributions
(@fig-reallocation-scatter) makes the logic plain. **Content is one-sided**: ~94% of pathways
rise (the top marginal is a single hump right of zero) -- Myc raises nearly the whole
compartment. **Priority is wide**: it straddles zero, only ~56% up (the right marginal). That
wide priority spread, sitting on a near-uniform content increase, *is* the reprioritisation.
Two features stand out. (1) The reallocation is **directional**: the content-down / priority-up
quadrant is **empty** -- nothing that falls in content gains priority, so the demoted pathways
are not switched off, they simply rise *slower than the biogenesis machinery* and lose relative
share. (2) The priority extremes name candidates: **most promoted** are the import chaperones and
SAM, the immune response and anaplerotic metabolism (creatine, pyruvate); **most demoted (raised
yet)** are the OXPHOS assembly factors, cAMP-PKA signalling, the dynamics/apoptosis pathways and
the mtDNA-encoded arm. No trend line is drawn: mitoPPS is a ratio (~content relative to the
compartment mean), so a content/priority slope is partly built in. Descriptive/exploratory
(n=6/group).

```{r}
#| label: fig-reallocation-scatter
#| fig-cap: "Supplementary (figS5): priority (relative mitoPPS effect, x) vs content (set-average log2FC, y) per pathway at 6W, with marginal densities; the 7 top-level categories labelled. Content is one-sided-positive (~94% up), priority wide both ways (~56% up) -- that spread is the reallocation. Ringed = padj<0.05. n=6/group."
#| fig-width: 9
#| fig-height: 6.6
#| echo: false
source(here::here("figures", "figS5_reallocation_scatter.R"))
print(p)
```

## Reallocation over the timeline {#sec-reallocation-time}

The reallocation above is the Myc *genotype* effect within a timepoint. A **separate**
phenomenon plays out over the timeline: the **WT background itself reprioritises** as the tissue
develops. Ranking the pathways by their WT 6->12W mitoPPS change (@fig-reallocation-time) shows a
full developmental reallocation in the WT alone -- and the dropline to the Myc+ 6->12W change
shows where Myc **bends** that trajectory. The two are largely shared (WT-time and Myc+-time
mitoPPS correlate r=0.75, 70% same direction, @fig-reallocation-time-div): most of the temporal
reallocation is developmental, common to both genotypes. Where Myc departs is the **fade** -- the
biogenesis / biosynthesis and OXPHOS arms it promoted at 6W lose relative priority hardest by 12W
in Myc+ (they sit *below* the y=x line), the mitoPPS reading of the attenuation (@sec-attenuation).
The apoptosis / mitophagy pathways move with the developmental background rather than diverging --
i.e. no Myc-specific temporal priming signal in this lens.

::: {.callout-warning}
The **time axis is batch-confounded** (batch = timepoint), so the individual WT-time and Myc+-time
changes are **descriptive, not claims**. The clean quantity is their **divergence** -- the dropline
in @fig-reallocation-time, the deviation from y=x in @fig-reallocation-time-div -- because the
batch effect cancels in the difference. Both are exploratory (n=6/group).
:::

```{r}
#| label: fig-reallocation-time
#| fig-cap: "Supplementary (figS6): MitoPathways ranked by WT 6->12W mitoPPS change (filled), with a dropline to Myc+ 6->12W (open) -- the Myc-specific bend. The 7 top-level categories labelled. n=6/group."
#| fig-width: 9
#| fig-height: 5
#| echo: false
source(here::here("figures", "figS6_reallocation_temporal_ranked.R"))
print(p)
```

```{r}
#| label: fig-reallocation-time-div
#| fig-cap: "Supplementary (figS7): WT 6->12W vs Myc+ 6->12W mitoPPS change per pathway. On the y=x line = shared/developmental (r=0.75); off it = where Myc bends the WT trajectory; below-diagonal = the biogenesis arms Myc fades. n=6/group."
#| fig-width: 6.5
#| fig-height: 6.8
#| echo: false
source(here::here("figures", "figS7_reallocation_temporal_divergence.R"))
print(p)
```

# The global axis and the correlation ceiling {#sec-axis}

*(stub)* Why cross-sectional couplings are unreliable: the 884 pathway scores collapse onto
one global per-sample axis (`pathway_loading`: PC1=76% var, eff. dim 1.7; OXPHOS r=0.968 with
the global mean). A mito axis's median coupling to an *arbitrary* library set is ~0.80.
"OXPHOS is central" is therefore untestable (no separable variance); mito is mito-LED, not
mito-SPECIFIC. Panels: `pathway_loading/A_scree`, `A3_gene_axis_covariates`, `linear_pathway_coupling/A_ceiling_origin`.

# The oncogenic programme attenuates across progression {#sec-attenuation}

**SUPPLEMENTARY FIGURE 2** carries the four controls @sec-window depends on. PUMA's collapse
is read against a global x0.55 rescaling, so before that reading is allowed: the x0.55 is
**not** a falling dose (every measure that would register one *widens* while MYC's output
narrows); it **is** real and uniform (the 12W profile is the 6W profile at 55% amplitude, its
coherence at the 100th percentile of matched shuffles); what MYC adds over the window is a
**uniform offset**, not a different set of pathways, and that offset is the batch-clean
quantity; and PUMA's collapse **survives removal of the BCL-xL denominator** on a genome-wide
statistic.

```{r}
#| label: fig-controls
#| fig-cap: "SUPPLEMENTARY FIGURE 2 (figureS2). The controls behind Figure 2. (A) Driver against output: everything registering a falling MYC dose widens over the window while the genotype gap in Hallmark MYC targets V2 narrows, so the attenuation is downstream of dose. (B) The 12W MYC effect is the 6W effect at ~55% amplitude. (C) Myc+ against wild-type temporal change: the shared slope is not beyond its null, the MYC-specific intercept is at the 0th percentile. (D) The collapse statistic over 8,774 MYC-responsive genes, marking the two pre-specified genes and Myc itself."
#| fig-width: 9
#| fig-height: 7.7
#| echo: false
source(here::here("figures", "figureS2_controls.R"))
print(p)
```

Foreshadowed already in fig01b (@sec-content, tension 2): the per-gene induction shrinks
6W->12W (OXPHOS still 87% up at 12W but only 1 gene clears FDR), and the Time-Myc+ contrast
is net **down** (Nuclear MitoCarta 68 down vs 20 up) -- the fade, in the same panel. The
question this section settles is *why*: does the Myc effect fade because the **background**
the oncogene acts on has moved, or because the **programme itself** decays? @fig-background
answers it in four panels, all built on one exact identity -- the change in the genotype gap
equals (Myc+ temporal change) - (WT temporal change), which holds to machine precision on
both the content and the priority ruler.

**First, the trivial explanation: the driver does not fade** (@fig-myc-dose). Myc's own
genotype gap *widens* over the window, **+1.64 log2 at 6W (p=2.5e-4) to +1.80 at 12W
(p=9.6e-6)**, and no within-genotype decline could carry the fade (Myc+ 8142 -> 7578
normalised counts, p=0.65). The WT baseline itself drifts down borderline (2543 -> 2143,
p=0.052), but that sits on the batch-confounded axis -- which is precisely why the gap, the
batch-clean quantity, is the one read. The proximal MYC/MAX/MXD network is flat: across MAX,
the MXD1/3/4-MXI1-MNT-MGA repressors and the MLX arm, 40 simple-effect tests give 5 at nominal
p<0.05 and **none survive BH** (min 0.052), with no genome-wide interaction padj<0.05 for any
of them -- no repressor rises to titrate MAX away.

The contrast that settles it is driver against output. **Every** measure that would register a
falling dose widens over the window -- Myc **+0.17**, Myc:MAX **+0.32** (+1.42 -> +1.74),
Myc:repressor-sum **+0.25** (+1.65 -> +1.90) -- while Myc's *output*, the genotype gap in
Hallmark MYC targets V2, **narrows by -0.36** (+1.32 -> +0.95). Driver up, output down: the
attenuation is **downstream of dose**, and the rest of this section is the decomposition of
that downstream part. Two bounds on the claim. Reads map to the mouse *Myc* locus, so total
message in Myc+ is transgene plus endogenous and the two are not separable here (script 27
decomposes them at programme level) -- "constant dose" is a statement about total message,
which is what the substrate sees. And transcript is not protein: the 31-gene post-translational
stability panel (`outputs/myc_stability_panel/`, script 41) finds nothing Myc-specific (every
interaction padj>0.05), no excess movement over an expression-matched background, and
directions that predict a *more* stable MYC (Fbxw7 -0.35 and Huwe1 -0.24, the two established
degradative E3 ligases, both down) -- but that is absence of transcriptional evidence, not evidence of
absence. The settling experiments are protein-level: a MYC blot at 6W vs 12W, phospho-T58/S62,
and a cycloheximide chase.

```{r}
#| label: fig-myc-dose
#| fig-cap: "figS8: is the Myc dose constant? (A) Myc normalised counts across the four groups. (B) The proximal MYC/MAX/MXD network -- MAX, the MXD/MNT/MGA repressors and the MLX arm; brackets unlabelled, red = p<0.05. (C) Everything that would register a falling dose (Myc, Myc:MAX, Myc:repressor sum) against Myc's output (Hallmark MYC targets V2), as the genotype gap at each age; open = 6W, arrow to 12W. The driver gap widens while the output gap narrows -- the attenuation is downstream of dose. n=6/group."
#| fig-width: 9
#| fig-height: 7.4
#| echo: false
source(here::here("figures", "figS8_myc_network_levels.R"))
print(p)
```

**Myc's effect is rescaled, not reshaped.** Regressing the 12W Myc effect on the 6W effect
across the 144 MitoPathways gives **slope 0.55** (bootstrap 0.51-0.60), R2 0.80, rho 0.84 on
the content ruler and slope 0.64, R2 0.79, rho 0.93 on priority, with per-tier slopes
0.53-0.83 -- uniform (@fig-background B). The coherence sits at the **100th percentile** of
gene-label shuffles matched on set size, expression and pathway overlap. The 12W profile *is*
the 6W profile at 55% of the amplitude: the same pathways, in the same order, less of it.
This is the pathway-level form of the Issue #4 result that magnitude, not rank, shrinks.

**What Myc adds over the timeline is a uniform offset, not a different set of pathways.**
Regressing the Myc+ 6->12W change on the WT 6->12W change (@fig-background C) gives slope
0.75 and intercept -0.17. The **slope** -- the part both genotypes share -- is not
distinguishable from the null (null median 0.63, 82nd percentile). The **intercept** is, at
the 0th percentile. It is also the only batch-clean quantity in the comparison, since an
offset common to both temporal contrasts cancels in their difference.

**The background is not moving toward the Myc state.** The WT 6->12W vector projects onto the
Myc(6W) vector at just **0.064** -- against a null median of 0.24, the 9.8th percentile, i.e.
*less* than an arbitrary matched panel. Restricted to the biogenesis and OXPHOS arms the cos
angle is **-0.34** (content) / -0.39 (priority): the background moves *away* from where Myc
pushes. That aggregate is not homogeneous, and the per-tier breakdown is informative -- protein
import (**-0.63** content, **-0.77** priority) and OXPHOS (**-0.65**) move against Myc, while
mitochondrial central dogma (**+0.74**) and metabolism (+0.48) move with it. The Myc+
trajectory, by contrast, has cos angle **-0.84** with the Myc effect (0th percentile), and
-0.80 to -0.98 in every substantial tier: over the timeline the Myc+ compartment retraces the
Myc programme backwards. @fig-background A shows the same thing per tier -- a red Myc arrow
that keeps its direction and shortens, and a grey background arrow that points elsewhere.

**And the WT-convergence term does not survive its control.** The published decomposition
aligns each gene to Myc's induction direction and reads the WT temporal change along it -- but
the genotype contrast and the WT temporal contrast **share the 6W WT baseline with opposite
signs**, so noise in that baseline manufactures apparent convergence. The six 6W_neg mice are
therefore split over all 20 three-versus-three partitions -- and, because halving the baseline
also adds noise to the alignment direction (which pulls the statistic toward chance from either
side), each partition is run twice off the *same* half baseline, once shared between the two
contrasts and once split. Identical direction, identical universe, identical noise; the only
difference is the sharing.

Globally, sharing the baseline adds **+0.24** (sd 0.138) of apparent convergence, in 19 of the
20 partitions: `frac_wt_toward` falls from 0.720 to **0.478** -- chance -- and the percentage
from +61% to -12% (@fig-background D). And the artifact is **absent exactly where the published
analysis reported divergence**: +0.01 on the MYC-target core, -0.01 on OXPHOS subunits, so
their divergence stands as published. What survives on the other side is programme-specific:
the biosynthetic arms genuinely converge (amino acid 0.74 -> **0.65**, lipid 0.77 -> **0.63**,
TCA and nucleotide ~0.63-0.64). The programme-specificity was the whole result; the global
number was the artifact.

So the picture is **additive, not causal**: a reallocation of the mitochondrial compartment
that both genotypes undergo, plus a Myc programme that decays proportionally without changing
shape. The background's own move is 57-62% as large as the Myc effect and points in a
direction the Myc effect does not.

One number the section deliberately does not give is a single size for the attenuation. On
the same gene sets, the share ruler retains 77-83% of the genotype gap at 12W while the
set-average LFC rulers retain 54-63% -- the sign and direction agree, the magnitude is
normalisation-dependent (denominator: per-sample transcriptome sum vs median-of-ratios size
factors). The manuscript quotes the range.

::: {.callout-warning}
**Batch = timepoint.** Both temporal contrasts carry the same batch offset. It cancels in the
interaction -- @fig-background C's intercept, and the whole rescaling result -- but **not** in
the convergence/fade split, so calling the shared temporal vector "developmental" rather than
"batch" remains an interpretation. The sample-split removes the shared-*baseline* artifact; it
cannot remove the batch confound. Separately, a family of tempting statistics here is
**algebraically forced** and carries no information: cor(WT-time, change in Myc effect) is
-0.19 (content) / -0.33 (priority) and the value the marginals force is *exactly* the same,
because the change is the difference of the two temporal contrasts. The full list is in
`results/background_vs_myc.rds$artifact_ledger`. None of them may be cited.
:::

```{r}
#| label: fig-background
#| fig-cap: "fig03: is Myc's mitochondrial effect shaped by a changing background? (A) Each Level-1 MitoPathway tier as four states in priority x content; red arrows = the Myc effect at 6W (solid) and 12W (dashed), grey = each genotype's temporal move. (B) The 12W Myc effect is the 6W effect x0.55 -- rescaled, not reshaped. (C) Myc+ vs WT temporal change: the shared slope is not beyond the null, the Myc-specific intercept is. (D) WT-convergence on the matched control -- the same three-mouse 6W WT baseline shared between the two contrasts (open) vs split (filled), 20 partitions; tick = the published six-mouse value. The global term collapses to chance; the biosynthetic arms survive and OXPHOS/MYC-core divergence is untouched. n=6/group, exploratory."
#| fig-width: 9
#| fig-height: 8.6
#| echo: false
source(here::here("figures", "fig03_background_vs_myc.R"))
print(p)
```

Full numbers, nulls and the artifact ledger: `docs/2026-07-24_background_vs_myc_interpretation.md`
(script 40). The earlier per-gene decomposition (script 31) is unchanged on disk; script 40
tests it and reports the correction.

# Development opens the window {#sec-window}

**MANUSCRIPT FIGURE 2.** MYC builds that organelle on a substrate that is itself moving, and
what the substrate does over the same window is the other half of the story.

## The gland withdraws respiration, and only respiration

Between 6 and 12 weeks the wild-type gland withdraws from respiration **and from nothing else
that would explain it**. OXPHOS subunits fall **-0.255** in set-average log2FC, at the **0th
percentile** of 2000 expression-matched random sets, while the proliferative programme barely
moves (`PROLIF_*` pooled **-0.047**) and the **assembly factors of the very same complexes**
sit at **+0.001, the 50th percentile** (@fig-window A). Nucleotide metabolism and the
mitoribosome -- the two mitochondrial arms most tightly coupled to growth, chosen in advance
as the alternative hypothesis's best representatives -- do not move either. The steepest fall
in the whole comparator is the **TEB-versus-ductal signature, -0.422**: the terminal end bud
regresses, and the TEB is the apoptosis-loaded compartment, because duct cavitation is
achieved by apoptosis.

The claim is comparative, so it is tested as a comparison. `PROLIF_*` pooled is itself beyond
its own null (1.3rd percentile) -- it is *small*, not immobile -- so a one-set null cannot
carry the sentence. Redrawing **both** sets of a contrast together puts
OXPHOS-minus-proliferation at **-0.208, the 0th percentile** (@fig-window B). Nor is it an
artefact of the content rise: the withdrawal persists on mitoPPS, a within-compartment ratio
that is blind to how many mitochondria a cell has (@fig-window C). **The gland de-respires
without de-proliferating.**

## And it leaves the death machinery intact

Of the 25 pro- and 7 anti-apoptotic MitoCarta transcripts plus five non-MitoCarta brakes,
exactly **one** moves across the window at padj<0.05 -- and it moves *up* (`Bnip3` +0.63). No
brake is raised either: `Bcl2` padj 0.45, `Bcl2l1` 0.79, `Mcl1` 0.44, `Xiap` 0.79,
`Birc2/3/5` all non-significant (@fig-window D). **Development removes the input, not the
machine.** That is why what follows is a *competence* argument rather than an abundance one.

## What MYC therefore loses

Every arm of the MYC programme is rescaled by about **x0.55** across the window
(@sec-attenuation). Read against that rate, the `Bax:Bcl-xL` ratio retains **0.55** exactly --
the **50th percentile** of matched gene pairs conditioned on the same starting effect, i.e.
entirely unremarkable. The `PUMA:Bcl-xL` ratio does not fade: it **reverses**, from +0.674 at
6W (p 0.009) to -0.060 at 12W, a retention of **-0.09** at the **92nd percentile** of that
conditional null (interaction p 0.037; p 0.017 with epithelial and immune composition in the
model) (@fig-window E).

That result reproduces on a statistic that never touches the `Bcl2l1` denominator. Ranking all
8,774 MYC-responsive genes by their residual from the global rate puts `Bbc3` at the **0.51st
percentile** -- and `Bbc3` was **pre-specified** from the PGC-1a westerns, not found by
scanning (script 44; drawn in `figures/fig05_death_arm.R` panel D, which becomes Supplementary
Figure 2). Both halves of that prediction are reported and they split: `Bcl2l11`/BIM sits at
the **91.6th percentile**, so the in-vivo collapse is PUMA-specific. `Myc`'s own transcript
sits at the top of the same distribution (retention 1.09): the transgene message does not
attenuate, so the downstream fade is not transcript-level dose.

Two things this collapse is *not*. It has **no wild-type definition** -- the wild-type
`Bbc3:Bcl2l1` ratio drifts *up* (+0.168) and the whole effect sits in the genotype gap, the
batch-clean quantity. It is a loss of **inducibility**, not of level. And **PUMA is a solo**:
none of six pre-registered questions returns a module. What loses MYC-inducibility genome-wide
is luminal lineage identity (`MG_HS_GRAY` NES **-2.79**, padj 1.7e-17); what keeps it is the
MYC programme itself.

```{r}
#| label: fig-window
#| fig-cap: "MANUSCRIPT FIGURE 2 (figure2). Development opens the window. (A) Comparator arms on the wild-type window, each against its own expression-matched null; the assembly factors of the same complexes are the internal control. (B) The paired null on the difference, which is the form the comparative claim takes. (C) Content against priority: the withdrawal survives a content-blind ruler. (D) The apoptotic and brake transcripts across the same window: one moves, and it rises. (E) Retention against the x0.55 global rate, with the conditional matched-pair null. (F) The three propositions the mouse hands to the cell experiments, each with its control and its adjudicating experiment. Wild-type only in A-D, n=6/timepoint; F is a ranking, not a test."
#| fig-width: 9
#| fig-height: 10.7
#| echo: false
source(here::here("figures", "figure2_developmental_window.R"))
print(p)
```

## The three propositions the mouse hands to the cell experiments

These numbers rank hypotheses; they do not test them. Read that way the same data licenses
three propositions, two of which the cell experiments already answer (@fig-window F).

**H1 -- respiratory state gates PUMA priming.** Whether MYC raises the PUMA:BCL-xL ratio
depends on where an animal sits on the respiratory-priority axis: the interaction lands at the
**91.9th percentile** of a within-timepoint permutation null, while the same test on `redox` --
a matched mitochondrial axis MYC does *not* drive -- sits at the **50.9th, dead centre**. It is
not shared with `BAX:BCL-xL` in any readable way (67.2nd against its own control at 63.3rd).
*Adjudicated by: raising PGC-1a in MYC-driven cells.*

**H2 -- the TEB-to-ductal transition moves both together.** TEB is the steepest faller of any
arm and moderates the same ratio at the **90.8th percentile**, again PUMA-specific over BAX.
The terminal end bud both respires hardest and clears its lumen by apoptosis, so this is the
honest rival to H1 and it is **not separable from it at n=24**. *Adjudicated by: single-cell or
deconvolution, or TEB- versus ductal-derived cells -- open.*

**H3 -- respiration costs proliferation, unless death is blocked.** On the same axis, in the
same animals, MYC is *worse* at proliferating where respiratory priority is high (**6.3rd
percentile**; E2F **9.2nd**). In an animal death is intact, so the liability limb is the one
expressed. *Adjudicated by: PGC-1a in MYAZ with and without Bcl-xL -- and the sign inverts.*
**The mouse supplies the negative limb, the cells the positive one, and the switch is whether
the cell can die.**

::: {.callout-important collapse="true"}
## The bounds -- and they are severe

**Nothing in panel F is significant.** The best permutation p is 0.081. Three specific limits:
the respiratory axis is **collinear with the timepoint it would explain** (the term goes
p 0.005 -> 0.088 once `tp*myc` is in the model), which is the structure of a two-timepoint
cross-sectional design and is not repairable by statistics; H1 and H2 are not separable; and a
mediation test is uninformative here rather than supportive (adding OXPHOS makes the
interaction *stronger*, and the mediator is itself MYC-driven). One column of the source table
must be distrusted outright: adding timepoint makes the *proliferation* and composite terms
stronger while PUMA weakens, and terms that grow on adding a collinear variable are
collinearity artefacts.

**The composite version of the death result is dead three times over**, which is why the
finding is gene-level. (i) The GSVA-space mitochondria-to-death bridge is **circular** -- the
priming composite is built from `MITOCARTA_APOPTOSIS_PRO`/`_ANTI`, so it correlated
mitochondrial genes with mitochondrial genes. (ii) "MYC raises apoptotic priming" as a
composite **is the content effect**: against an expression-matched *MitoCarta* background the
pro-apoptotic set rises +0.196 where the background rises +0.364 (z -2.42), i.e. below average
for a mitochondrial gene, and on the 404 non-mitochondrial pro-death genes it does not
reproduce. (iii) The attenuation is **global**, so a composite priming signal fades because
the MYC programme fades. Nor does the gene-level result survive multiplicity: no pair beats BH
across the priming panel (minimum adjusted interaction p 0.33).

**One trap the panel encodes rather than captions:** `Bmf:Bcl-xL` sits *further* out on the
conditional null than `Bbc3:Bcl-xL` (97.5th percentile, p 0.025) but has **no 6W effect to
lose** (d6 0.24, p 0.40). It fails the precondition the whole statistic rests on.

**Batch = timepoint** bounds every wild-type temporal sentence above. Three things keep them
usable rather than uninterpretable: the comparison in panels A-B is *internal to one contrast*
(a batch offset shared by all gene sets cannot make OXPHOS subunits fall while the assembly
factors of the same complexes do not); it reproduces on two rulers with different
denominators; and the protein blots move the same way off the RNA batch entirely. Panel D's
flatness reads as "no detectable movement at n=6", not "no movement".

One caveat that is in no panel: within the 12 wild-type animals the OXPHOS and proliferation
scores correlate at **0.82**. The timepoint term survives adjustment for proliferation
(-0.499 -> -0.420, p 0.002) but at that collinearity a per-sample model cannot arbitrate. The
set-level dissociation in A-B, a contrast between gene sets rather than between samples, is
the load-bearing version.
:::

::: {.callout-note collapse="true"}
## A hypothesis about what carries the respiratory arm

Gray et al. (2023) describe a **"low-expressing" (LE)** state within each mammary lineage -- a
systematic suppression of the lineage programme, *not* an oestrogen axis; our library metadata
glosses it wrongly as "high- versus low-estrogen", recorded here so the error does not
propagate. LE states are pubertally enriched, and the ESRRA and GABPA mitochondrial regulons
are detected in the **AP (alveolar progenitor / LASP) LE** context. If the maturing gland loses
LE abundance, LE-marker sets should fall and HE-marker sets should not. In AP they do: LE
markers **-0.294** (0th percentile) against HE markers -0.049 (23rd); in BA the same. **In HS
the prediction fails** -- both arms fall -- so this is an AP/BA statement, not a global one.

Two reasons it stays a hypothesis. **25 of the 87 OXPHOS subunits are themselves AP LE-marker
genes**, so any lineage claim built on the raw sets is the OXPHOS claim restated; every number
here is from the mitochondria-removed version, and the raw self-adjusted figure (-0.044) must
never be quoted. And even mitochondria-removed, the two scores correlate at 0.93 across 12
samples. Separating a shift in state *abundance* from developmental *regulation* needs
deconvolution or single cells.

Relatedly, the withdrawal is **not** a whole-regulon event. `CORE_MITO` -- the ESRRA/NRF1/GABPA
targets intersecting MitoCarta, the closest DNA-level proxy for the PGC-1a axis -- is flat
overall (**+0.013**, 75th percentile) while its 59 OXPHOS-subunit members fall **-0.216**
(0.05th) and the other 241 rise **+0.069** (99.6th). `Ppargc1a` itself sits at baseMean 30 and
is not readable, which is why **PGC-1a is named in this paper only where the cell experiments
are described**.
:::

::: {.callout-note title="Planned experiment: MYAZ + PGC1a RNA-seq"}
Not yet done; the author intends to do it. It is **the only design that can define the
OXPHOS/apoptosis module by perturbation** rather than by correlation -- whatever PGC1a
co-induces that is both mitochondrial and apoptotic *is* the module, with no null to fight
and no correlation ceiling. In this dataset the module cannot be defined by co-expression:
the wild-type gland does not move its apoptotic transcripts at all, the PGC1a-axis regulon is
*depleted* of pro-apoptotic genes (fold 0.76) while MYC's is enriched (2.68x), and the
per-sample couplings sit at the ambient. See section 0.8 of
`docs/2026-07-26_introduction_alignment_and_the_question.md`. When the data exists it should
settle: whether the payload/trigger division of labour holds under direct PGC1a manipulation,
and why PUMA and BIM behave as a pair in cells but split in vivo.
:::

Sources on disk: `results/priming_arm_teb.rds` (script 42),
`results/substrate_specificity_tradeoff.rds` (43),
`results/collapse_module_ownership.rds` (44), and
`results/death_priming_reassessment.rds` (34, for the three composite negatives above). The
narrative and the propositions: `docs/2026-07-26_introduction_alignment_and_the_question.md`
sections 0, 0.5a, 0.7a, 0.8, 0.9, **0.10**. Blot values (MYC -50%; BAX/BIM/PUMA) and the
MYC-ER, MYAZ and Bcl-xL rescue results are the author's, reported here as given.

# External validation: Lesner et al. {#sec-lesner}

*(stub)* The MYC nuclear-OXPHOS-up / mtDNA-down discordance is independently established
(Lesner et al., bioRxiv 2026.07.13.738248) in TCGA, mouse HCC, proteomics, mtDNA qPCR --
killing the "our enzymatic-dissociation artifact" worry. Lesner validates the *question*; our
mtDNA-down arm is genotype-independent (p~0.49), i.e. underpowered, not demonstrated. Their
hypoxia->MXI1/JUN Myc-suppression route does NOT transfer to our model (script 39; liver-specific).
Panels: `lesner/A_mechanism_genes`, `lesner/B_fgsea_contrasts`, `lesner/C_mitophagy_specificity_null`.

# Leads for the bench {#sec-leads}

*(stub)* Hypothesis-generating, not results (n=24): `redox -> MB2 tumorigenic fork` (the one
robust Myc-independent coupling; test: redox/GSH assay) and `nuclear OXPHOS -> death-priming`
(test: **BH3 profiling** -- priming is an occupancy state at the outer membrane, and no
amount of transcriptomics substitutes for measuring it).

**`Foxo3`, new 2026-07-27 and the one live mechanistic lead** -- it collapses like PUMA
(0.87th percentile of its matched stratum; +0.243 at 6W -> -0.242 at 12W, interaction
p 0.0074, against `Bbc3`'s +0.258 -> -0.283, p 0.0081) and it is PUMA's canonical
p53-independent activator. Test: **FOXO3 western across the two ages, and in the PGC1a
arm.** Caveats that must travel with it: both 6W effects are non-significant, both
interactions are genome-wide null, two genes are not a module, and `Foxo3` *rises* in the
wild-type gland. Note also that our direction is **opposite** to the canonical
PGC-1a-sequesters-FOXO3a axis in `docs/library_reference/PUMA-and-its-relationships.md`. Not
for the Results on the RNA alone.

**Closed, do not propose it again:** the retrograde/ISR route to PUMA is tested and negative
(script 44 PART D -- the ATF4 target set and Hallmark UPR are null in the collapse ranking,
and `Chac1`, which looked collapsed, has no 6W effect to lose at p 0.47).

The OXPHOS->priming lead is **gene-level, not composite-level** (corrected 2026-07-27, it was
described here in its pre-script-43 composite form). The composite version is dead twice
over: script 34 showed it is circular, and script 43 PART B showed that of five outcomes only
the **gene-level `Bbc3:Bcl2l1` ratio** has a `redox` control that behaves (`myc x OXPHOS-priority`
+6.09, p 0.005, 91.7th percentile against a control at the 51.2nd). The claim is a **sign
contrast on one outcome**, not a magnitude comparison across five. Two further tests the
mechanism menu (section 0.9 of the alignment doc) names, neither answerable from RNA:
cardiolipin by lipidomics and cristae architecture by EM.

# Provenance {#sec-provenance}

```{r}
#| label: session
#| echo: false
sessionInfo()
```
